---
title: "CMIP6 prep"
author: "Camille"
date: "2024-02-19"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: hide
---
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = F, warning = F, error = F, fig.align = "center",
                      collapse = T)
```

```{r libraries}
library(raster)

library(plyr)
library(dplyr)

library(ggplot2)
library(viridisLite)
library(colorspace)
library(RColorBrewer)

library(sf)
library(stringr)

library(ggpubr)
library(kableExtra)

memory.limit(size = 160000)

#display.brewer.pal(n = 11, name ="RdYlGn")
# colorvec <- brewer.pal(n = 11, name ="RdYlGn")
# col.cat <- c("Hyperarid" = colorvec[2], "Arid"= colorvec[4],"Semi-arid"= colorvec[5], "Dry subhumid" = colorvec[8], "Humid"= colorvec[10], "Cold" = "powderblue")

#handmade
col.cat <- c("Hyperarid" = "#D73027", "Arid"= "#FDAE61","Semi-arid"= "#FEE08B", "Dry subhumid" = "#ABD9E9", "Humid"= "#4575B4", "Cold" = "#DAE2F7")

breaks.unesco <- c(-Inf,0,0.03,0.2,0.5,0.65,Inf)
cat.unesco <- c("(0,0.03]"= "Hyperarid", "(0.03,0.2]" = "Arid", "(0.2,0.5]" = "Semi-arid", "(0.5,0.65]" = "Dry subhumid","(0.65, Inf]" = "Humid", "(-Inf,0]" = "Cold")
```

# Data

Land-sea mask and IPCC region are taken from Iturbide et al. 2020: https://github.com/SantanderMetGroup/ATLAS/tree/devel/reference-grids

Elevation data is taken from Worldclim: https://www.worldclim.org/data/worldclim21.html


```{r land-masks}
land_mask <- raster("Masks/land_sea_mask_1degree.nc4") 
land_mask.df <- as.data.frame(land_mask, xy = T) %>% setNames(c("lon","lat","lm")) # 0 if ocean, 1 if land


elev <- raster("Worldclim/wc2.1_10m/wc2.1_10m_elev.tif") 
elev.df <- elev %>% projectRaster(to = land_mask) %>% as.data.frame(xy = T) %>% setNames(c("lon","lat", "z"))

ipcc_regions <- shapefile("Masks/IPCC-WGI-reference-regions-v4.shp") %>% spTransform(crs("EPSG:4326")) 
ipcc_regions.raster <- ipcc_regions %>% rasterize(land_mask)
ipcc_regions.df <- as.data.frame(ipcc_regions.raster, xy = T) %>% setNames(c("lon","lat","Continent","Type","Name","Acronym"))

```

CMIP6 data is dowloaded from https://aims2.llnl.gov/search 

The data was downloaded by model and extracted by model, period and SSP with the "data-prep-xxx.R" scripts, then grouped by variable ("bind-tables_xxx.R"), on GRICAD.

```{r cmip6-data, eval = F}
tas.all_annual <- read.table("CMIP6/tas/tas.all_annual.txt")
pr.all_annual <- read.table("CMIP6/pr/pr.all_annual.txt")
sfcWind.all_annual <- read.table("CMIP6/sfcWind/sfcWind.all_annual.txt")
hfls.all_annual <- read.table("CMIP6/hfls/hfls.all_annual.txt")
hfss.all_annual <- read.table("CMIP6/hfss/hfss.all_annual.txt")
hurs.all_annual <- read.table("CMIP6/hurs/hurs.all_annual.txt")

cmip6_annual <- tas.all_annual %>% merge(pr.all_annual, by = c("lon","lat","model","period","lm", "Continent","Type","Name","Acronym","source")) %>%
  merge(sfcWind.all_annual, by = c("lon","lat","model","period","lm", "Continent","Type","Name","Acronym","source")) %>%
  merge(hfls.all_annual, by = c("lon","lat","model","period","lm", "Continent","Type","Name","Acronym","source")) %>%
  merge(hfss.all_annual, by = c("lon","lat","model","period","lm", "Continent","Type","Name","Acronym","source")) %>%
  merge(hurs.all_annual, by = c("lon","lat","model","period","lm", "Continent","Type","Name","Acronym","source")) %>% 
  merge(elev.df, by = c("lon", "lat"))

write.table(cmip6_annual, "cmip6_annual.txt")
```

```{r cmip6-january, eval = F}
tas.all_january <- read.table("CMIP6/tas/tas.all_january.txt")
pr.all_january <- read.table("CMIP6/pr/pr.all_january.txt")
sfcWind.all_january <- read.table("CMIP6/sfcWind/sfcWind.all_january.txt")
hfls.all_january <- read.table("CMIP6/hfls/hfls.all_january.txt")
hfss.all_january <- read.table("CMIP6/hfss/hfss.all_january.txt")
hurs.all_january <- read.table("CMIP6/hurs/hurs.all_january.txt")

cmip6_january <- tas.all_january %>% merge(pr.all_january, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>%
  merge(sfcWind.all_january, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>%
  merge(hfls.all_january, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>%
  merge(hfss.all_january, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>%
  merge(hurs.all_january, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>% 
  merge(elev.df, by = c("lon", "lat"))

write.table(cmip6_january, "cmip6_january.txt")
```

```{r cmip6-february, eval = F}
tas.all_february <- read.table("CMIP6/tas/tas.all_february.txt")
pr.all_february <- read.table("CMIP6/pr/pr.all_february.txt")
sfcWind.all_february <- read.table("CMIP6/sfcWind/sfcWind.all_february.txt")
hfls.all_february <- read.table("CMIP6/hfls/hfls.all_february.txt")
hfss.all_february <- read.table("CMIP6/hfss/hfss.all_february.txt")
hurs.all_february <- read.table("CMIP6/hurs/hurs.all_february.txt")

cmip6_february <- tas.all_february %>% merge(pr.all_february, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>%
  merge(sfcWind.all_february, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>%
  merge(hfls.all_february, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>%
  merge(hfss.all_february, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>%
  merge(hurs.all_february, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>% 
  merge(elev.df, by = c("lon", "lat"))

write.table(cmip6_february, "cmip6_february.txt")
```

```{r cmip6-march, eval = F}
tas.all_march <- read.table("CMIP6/tas/tas.all_march.txt")
pr.all_march <- read.table("CMIP6/pr/pr.all_march.txt")
sfcWind.all_march <- read.table("CMIP6/sfcWind/sfcWind.all_march.txt")
hfls.all_march <- read.table("CMIP6/hfls/hfls.all_march.txt")
hfss.all_march <- read.table("CMIP6/hfss/hfss.all_march.txt")
hurs.all_march <- read.table("CMIP6/hurs/hurs.all_march.txt")

cmip6_march <- tas.all_march %>% merge(pr.all_march, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>%
  merge(sfcWind.all_march, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>%
  merge(hfls.all_march, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>%
  merge(hfss.all_march, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>%
  merge(hurs.all_march, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>% 
  merge(elev.df, by = c("lon", "lat"))

write.table(cmip6_march, "cmip6_march.txt")
```

```{r cmip6-april, eval = F}
tas.all_april <- read.table("CMIP6/tas/tas.all_april.txt")
pr.all_april <- read.table("CMIP6/pr/pr.all_april.txt")
sfcWind.all_april <- read.table("CMIP6/sfcWind/sfcWind.all_april.txt")
hfls.all_april <- read.table("CMIP6/hfls/hfls.all_april.txt")
hfss.all_april <- read.table("CMIP6/hfss/hfss.all_april.txt")
hurs.all_april <- read.table("CMIP6/hurs/hurs.all_april.txt")

cmip6_april <- tas.all_april %>% merge(pr.all_april, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>%
  merge(sfcWind.all_april, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>%
  merge(hfls.all_april, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>%
  merge(hfss.all_april, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>%
  merge(hurs.all_april, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>% 
  merge(elev.df, by = c("lon", "lat"))

write.table(cmip6_april, "cmip6_april.txt")
```

```{r cmip6-may, eval = F}
tas.all_may <- read.table("CMIP6/tas/tas.all_may.txt")
pr.all_may <- read.table("CMIP6/pr/pr.all_may.txt")
sfcWind.all_may <- read.table("CMIP6/sfcWind/sfcWind.all_may.txt")
hfls.all_may <- read.table("CMIP6/hfls/hfls.all_may.txt")
hfss.all_may <- read.table("CMIP6/hfss/hfss.all_may.txt")
hurs.all_may <- read.table("CMIP6/hurs/hurs.all_may.txt")

cmip6_may <- tas.all_may %>% merge(pr.all_may, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>%
  merge(sfcWind.all_may, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>%
  merge(hfls.all_may, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>%
  merge(hfss.all_may, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>%
  merge(hurs.all_may, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>% 
  merge(elev.df, by = c("lon", "lat"))

write.table(cmip6_may, "cmip6_may.txt")
```

```{r cmip6-june, eval = F}
tas.all_june <- read.table("CMIP6/tas/tas.all_june.txt")
pr.all_june <- read.table("CMIP6/pr/pr.all_june.txt")
sfcWind.all_june <- read.table("CMIP6/sfcWind/sfcWind.all_june.txt")
hfls.all_june <- read.table("CMIP6/hfls/hfls.all_june.txt")
hfss.all_june <- read.table("CMIP6/hfss/hfss.all_june.txt")
hurs.all_june <- read.table("CMIP6/hurs/hurs.all_june.txt")

cmip6_june <- tas.all_june %>% merge(pr.all_june, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>%
  merge(sfcWind.all_june, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>%
  merge(hfls.all_june, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>%
  merge(hfss.all_june, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>%
  merge(hurs.all_june, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>% 
  merge(elev.df, by = c("lon", "lat"))

write.table(cmip6_june, "cmip6_june.txt")
```

```{r cmip6-july, eval = F}
tas.all_july <- read.table("CMIP6/tas/tas.all_july.txt")
pr.all_july <- read.table("CMIP6/pr/pr.all_july.txt")
sfcWind.all_july <- read.table("CMIP6/sfcWind/sfcWind.all_july.txt")
hfls.all_july <- read.table("CMIP6/hfls/hfls.all_july.txt")
hfss.all_july <- read.table("CMIP6/hfss/hfss.all_july.txt")
hurs.all_july <- read.table("CMIP6/hurs/hurs.all_july.txt")

cmip6_july <- tas.all_july %>% merge(pr.all_july, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>%
  merge(sfcWind.all_july, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>%
  merge(hfls.all_july, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>%
  merge(hfss.all_july, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>%
  merge(hurs.all_july, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>% 
  merge(elev.df, by = c("lon", "lat"))

write.table(cmip6_july, "cmip6_july.txt")
```

```{r cmip6-august, eval = F}
tas.all_august <- read.table("CMIP6/tas/tas.all_august.txt")
pr.all_august <- read.table("CMIP6/pr/pr.all_august.txt")
sfcWind.all_august <- read.table("CMIP6/sfcWind/sfcWind.all_august.txt")
hfls.all_august <- read.table("CMIP6/hfls/hfls.all_august.txt")
hfss.all_august <- read.table("CMIP6/hfss/hfss.all_august.txt")
hurs.all_august <- read.table("CMIP6/hurs/hurs.all_august.txt")

cmip6_august <- tas.all_august %>% merge(pr.all_august, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>%
  merge(sfcWind.all_august, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>%
  merge(hfls.all_august, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>%
  merge(hfss.all_august, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>%
  merge(hurs.all_august, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>% 
  merge(elev.df, by = c("lon", "lat"))

write.table(cmip6_august, "cmip6_august.txt")
```

```{r cmip6-september, eval = F}
tas.all_september <- read.table("CMIP6/tas/tas.all_september.txt")
pr.all_september <- read.table("CMIP6/pr/pr.all_september.txt")
sfcWind.all_september <- read.table("CMIP6/sfcWind/sfcWind.all_september.txt")
hfls.all_september <- read.table("CMIP6/hfls/hfls.all_september.txt")
hfss.all_september <- read.table("CMIP6/hfss/hfss.all_september.txt")
hurs.all_september <- read.table("CMIP6/hurs/hurs.all_september.txt")

cmip6_september <- tas.all_september %>% merge(pr.all_september, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>%
  merge(sfcWind.all_september, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>%
  merge(hfls.all_september, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>%
  merge(hfss.all_september, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>%
  merge(hurs.all_september, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>% 
  merge(elev.df, by = c("lon", "lat"))

write.table(cmip6_september, "cmip6_september.txt")
```

```{r cmip6-october, eval = F}
tas.all_october <- read.table("CMIP6/tas/tas.all_october.txt")
pr.all_october <- read.table("CMIP6/pr/pr.all_october.txt")
sfcWind.all_october <- read.table("CMIP6/sfcWind/sfcWind.all_october.txt")
hfls.all_october <- read.table("CMIP6/hfls/hfls.all_october.txt")
hfss.all_october <- read.table("CMIP6/hfss/hfss.all_october.txt")
hurs.all_october <- read.table("CMIP6/hurs/hurs.all_october.txt")

cmip6_october <- tas.all_october %>% merge(pr.all_october, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>%
  merge(sfcWind.all_october, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>%
  merge(hfls.all_october, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>%
  merge(hfss.all_october, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>%
  merge(hurs.all_october, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>% 
  merge(elev.df, by = c("lon", "lat"))

write.table(cmip6_october, "cmip6_october.txt")
```

```{r cmip6-november, eval = F}
tas.all_november <- read.table("CMIP6/tas/tas.all_november.txt")
pr.all_november <- read.table("CMIP6/pr/pr.all_november.txt")
sfcWind.all_november <- read.table("CMIP6/sfcWind/sfcWind.all_november.txt")
hfls.all_november <- read.table("CMIP6/hfls/hfls.all_november.txt")
hfss.all_november <- read.table("CMIP6/hfss/hfss.all_november.txt")
hurs.all_november <- read.table("CMIP6/hurs/hurs.all_november.txt")

cmip6_november <- tas.all_november %>% merge(pr.all_november, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>%
  merge(sfcWind.all_november, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>%
  merge(hfls.all_november, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>%
  merge(hfss.all_november, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>%
  merge(hurs.all_november, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>% 
  merge(elev.df, by = c("lon", "lat"))

write.table(cmip6_november, "cmip6_november.txt")
```

```{r cmip6-december, eval = F}
tas.all_december <- read.table("CMIP6/tas/tas.all_december.txt")
pr.all_december <- read.table("CMIP6/pr/pr.all_december.txt")
sfcWind.all_december <- read.table("CMIP6/sfcWind/sfcWind.all_december.txt")
hfls.all_december <- read.table("CMIP6/hfls/hfls.all_december.txt")
hfss.all_december <- read.table("CMIP6/hfss/hfss.all_december.txt")
hurs.all_december <- read.table("CMIP6/hurs/hurs.all_december.txt")

cmip6_december <- tas.all_december %>% merge(pr.all_december, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>%
  merge(sfcWind.all_december, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>%
  merge(hfls.all_december, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>%
  merge(hfss.all_december, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>%
  merge(hurs.all_december, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>% 
  merge(elev.df, by = c("lon", "lat"))

write.table(cmip6_december, "cmip6_december.txt")
```


# Aridity Index

## De Martonne



```{r aridity-index-martonne, fig.dim = c(15, 20), cache = T}
cmip6 <- read.table("cmip6.ref.txt") # created in "figures_article.Rmd", already converted in mm and °C

breaks.martonne <- c(0,5,10,20,"30","40",Inf)
cat.martonne <- c("[0,5]" = "Desert", "(5,10]"= "Arid", "(10,20]" = "Semi-arid","(20,30]" = "Temperate", "(30,40]" = "Humid", "(40,Inf]" = "Forest")
col.martonne <- c("Desert" =  "#D73027", "Arid"= "#FDAE61","Semi-arid"= "#FEE08B", "Temperate" = "#ABD9E9", "Humid"= "#4575B4","Forest"=  "#DAE2F7")

cmip6$AIm <- with(cmip6,spr.ref/(t.ref)) # pr in mm/y, tas in Ceslsius

cmip6$cat.AIm <- cmip6$AIm %>% cut(breaks.martonne, include.lowest = T) %>% revalue(cat.martonne)

map_list <- list()

for(i in unique(cmip6$source)){
  g <- ggplot(data = subset(cmip6, period == "1970_2000" & source == i)) + geom_raster(aes(x=lon, y = lat,  fill = cat.AIm))+
    scale_fill_manual(values = col.martonne, na.translate = F)+
    labs(title = paste(i, "1970-2000", sep = ", "), fill = "")+
    theme_void()
  map_list[[i]] <- g
}

ggpubr::ggarrange(plotlist = map_list, ncol = 3, nrow = 5, common.legend = T, legend = "bottom")
```

## AI FAO (with Penman-Monteith) 

ea from Rh:

ea = 6.108*(RH/100)*exp(17.27*T/(T+237.3))

Refs: https://www.nature.com/articles/s41598-023-40499-6 : Allen and Tetens

All variables are calculated by month, so that we know the monthly evapotranspiration. The monthly values are summed to obtain the annual values.

```{r aridity-index-unesco, eval = F}
id_vars <- c("lon","lat","model","period","lm", "Continent","Type","Name","Acronym","source")

cmip6_annual <- mutate(read.table("cmip6_annual.txt"),
                       t = tas - 273.15,
                       Rn = (hfls + hfss) * 86400 * 1e-6, # conversion from W/m2 to MJ/m2/day.
                       P = 101.3 * ((293-0.0065*z)/293)^(5.26),
                       es = 0.61078*exp(17.27*t/(t+237.3)),
                       ea = (hurs/100)*es,
                       delta = 4098*es/(t+237.3)^2, #slope vapor pressure
                       gamma = 0.00163*P/(2.501-52.6361e-3*t), #psychrometric constant
                       ET0_annual = ((0.408*delta*Rn+gamma*900/((t)+273)*0.748*sfcWind*(es-ea))/(delta+gamma*(1+0.34*0.748*sfcWind)))) 

cmip6_01 <- mutate(read.table("cmip6_january.txt"),
                   t = tas - 273.15,
                   pr_01 = pr*60*60*24, # mm/day
                   Rn = (hfls + hfss) * 86400 * 1e-6, # conversion from W/m2 to MJ/m2/day.
                   P = 101.3 * ((293-0.0065*z)/293)^(5.26),
                   es = 0.61078*exp(17.27*t/(t+237.3)),
                   ea = (hurs/100)*es,
                   delta = 4098*es/(t+237.3)^2, #slope vapor pressure
                   gamma = 0.00163*P/(2.501-52.6361e-3*t), #psychrometric constant
                   ET0_01 = ((0.408*delta*Rn+gamma*900/(t+273)*0.748*sfcWind*(es-ea))/(delta+gamma*(1+0.34*0.748*sfcWind)))) # Evapotranspiration, G considered to be negligible. in mm/day. Wind speed at 10 m is converted to wind speed at 2 m by multiplying by 0.748

cmip6_02 <- mutate(read.table("cmip6_february.txt"),
                   t = tas - 273.15,
                   pr_02 = pr*60*60*24, # mm/day
                   Rn = (hfls + hfss) * 86400 * 1e-6, # conversion from W/m2 to MJ/m2/day.
                   P = 101.3 * ((293-0.0065*z)/293)^(5.26),
                   es = 0.61078*exp(17.27*t/(t+237.3)),
                   ea = (hurs/100)*es,
                   delta = 4098*es/(t+237.3)^2, #slope vapor pressure
                   gamma = 0.00163*P/(2.501-52.6361e-3*t), #psychrometric constant
                   ET0_02 = ((0.408*delta*Rn+gamma*900/(t+273)*0.748*sfcWind*(es-ea))/(delta+gamma*(1+0.34*0.748*sfcWind))))

cmip6_03 <- mutate(read.table("cmip6_march.txt"),
                   t = tas - 273.15,
                   pr_03 = pr*60*60*24, # mm/day
                   Rn = (hfls + hfss) * 86400 * 1e-6, # conversion from W/m2 to MJ/m2/day.
                   P = 101.3 * ((293-0.0065*z)/293)^(5.26),
                   es = 0.61078*exp(17.27*t/(t+237.3)),
                   ea = (hurs/100)*es,
                   delta = 4098*es/(t+237.3)^2, #slope vapor pressure
                   gamma = 0.00163*P/(2.501-52.6361e-3*t), #psychrometric constant
                   ET0_03 = ((0.408*delta*Rn+gamma*900/(t+273)*0.748*sfcWind*(es-ea))/(delta+gamma*(1+0.34*0.748*sfcWind))))

cmip6_04 <- mutate(read.table("cmip6_april.txt"),
                   t = tas - 273.15,
                   pr_04 = pr*60*60*24, # mm/day
                   Rn = (hfls + hfss) * 86400 * 1e-6, # conversion from W/m2 to MJ/m2/day.
                   P = 101.3 * ((293-0.0065*z)/293)^(5.26),
                   es = 0.61078*exp(17.27*t/(t+237.3)),
                   ea = (hurs/100)*es,
                   delta = 4098*es/(t+237.3)^2, #slope vapor pressure
                   gamma = 0.00163*P/(2.501-52.6361e-3*t), #psychrometric constant
                   ET0_04 = ((0.408*delta*Rn+gamma*900/(t+273)*0.748*sfcWind*(es-ea))/(delta+gamma*(1+0.34*0.748*sfcWind))))

cmip6_05 <- mutate(read.table("cmip6_may.txt"),
                   t = tas - 273.15,
                   pr_05 = pr*60*60*24, # mm/day
                   Rn = (hfls + hfss) * 86400 * 1e-6, # conversion from W/m2 to MJ/m2/day.
                   P = 101.3 * ((293-0.0065*z)/293)^(5.26),
                   es = 0.61078*exp(17.27*t/(t+237.3)),
                   ea = (hurs/100)*es,
                   delta = 4098*es/(t+237.3)^2, #slope vapor pressure
                   gamma = 0.00163*P/(2.501-52.6361e-3*t), #psychrometric constant
                   ET0_05 = ((0.408*delta*Rn+gamma*900/(t+273)*0.748*sfcWind*(es-ea))/(delta+gamma*(1+0.34*0.748*sfcWind))))

cmip6_06 <- mutate(read.table("cmip6_june.txt"),
                   t = tas - 273.15,
                   pr_06 = pr*60*60*24, # mm/day
                   Rn = (hfls + hfss) * 86400 * 1e-6, # conversion from W/m2 to MJ/m2/day.
                   P = 101.3 * ((293-0.0065*z)/293)^(5.26),
                   es = 0.61078*exp(17.27*t/(t+237.3)),
                   ea = (hurs/100)*es,
                   delta = 4098*es/(t+237.3)^2, #slope vapor pressure
                   gamma = 0.00163*P/(2.501-52.6361e-3*t), #psychrometric constant
                   ET0_06 = ((0.408*delta*Rn+gamma*900/(t+273)*0.748*sfcWind*(es-ea))/(delta+gamma*(1+0.34*0.748*sfcWind))))

cmip6_07 <- mutate(read.table("cmip6_july.txt"),
                   t = tas - 273.15,
                   pr_07 = pr*60*60*24, # mm/day
                   Rn = (hfls + hfss) * 86400 * 1e-6, # conversion from W/m2 to MJ/m2/day.
                   P = 101.3 * ((293-0.0065*z)/293)^(5.26),
                   es = 0.61078*exp(17.27*t/(t+237.3)),
                   ea = (hurs/100)*es,
                   delta = 4098*es/(t+237.3)^2, #slope vapor pressure
                   gamma = 0.00163*P/(2.501-52.6361e-3*t), #psychrometric constant
                   ET0_07 = ((0.408*delta*Rn+gamma*900/(t+273)*0.748*sfcWind*(es-ea))/(delta+gamma*(1+0.34*0.748*sfcWind))))

cmip6_08 <- mutate(read.table("cmip6_august.txt"),
                   t = tas - 273.15,
                   pr_08 = pr*60*60*24, # mm/day
                   Rn = (hfls + hfss) * 86400 * 1e-6, # conversion from W/m2 to MJ/m2/day.
                   P = 101.3 * ((293-0.0065*z)/293)^(5.26),
                   es = 0.61078*exp(17.27*t/(t+237.3)),
                   ea = (hurs/100)*es,
                   delta = 4098*es/(t+237.3)^2, #slope vapor pressure
                   gamma = 0.00163*P/(2.501-52.6361e-3*t), #psychrometric constant
                   ET0_08 = ((0.408*delta*Rn+gamma*900/(t+273)*0.748*sfcWind*(es-ea))/(delta+gamma*(1+0.34*0.748*sfcWind))))

cmip6_09 <- mutate(read.table("cmip6_september.txt"),
                   t = tas - 273.15,
                   pr_09 = pr*60*60*24, # mm/day
                   Rn = (hfls + hfss) * 86400 * 1e-6, # conversion from W/m2 to MJ/m2/day.
                   P = 101.3 * ((293-0.0065*z)/293)^(5.26),
                   es = 0.61078*exp(17.27*t/(t+237.3)),
                   ea = (hurs/100)*es,
                   delta = 4098*es/(t+237.3)^2, #slope vapor pressure
                   gamma = 0.00163*P/(2.501-52.6361e-3*t), #psychrometric constant
                   ET0_09 = ((0.408*delta*Rn+gamma*900/(t+273)*0.748*sfcWind*(es-ea))/(delta+gamma*(1+0.34*0.748*sfcWind))))

cmip6_10 <- mutate(read.table("cmip6_october.txt"),
                   t = tas - 273.15,
                   pr_10 = pr*60*60*24, # mm/day
                   Rn = (hfls + hfss) * 86400 * 1e-6, # conversion from W/m2 to MJ/m2/day.
                   P = 101.3 * ((293-0.0065*z)/293)^(5.26),
                   es = 0.61078*exp(17.27*t/(t+237.3)),
                   ea = (hurs/100)*es,
                   delta = 4098*es/(t+237.3)^2, #slope vapor pressure
                   gamma = 0.00163*P/(2.501-52.6361e-3*t), #psychrometric constant
                   ET0_10 = ((0.408*delta*Rn+gamma*900/(t+273)*0.748*sfcWind*(es-ea))/(delta+gamma*(1+0.34*0.748*sfcWind))))

cmip6_11 <- mutate(read.table("cmip6_november.txt"),
                   t = tas - 273.15,
                   pr_11 = pr*60*60*24, # mm/day
                   Rn = (hfls + hfss) * 86400 * 1e-6, # conversion from W/m2 to MJ/m2/day.
                   P = 101.3 * ((293-0.0065*z)/293)^(5.26),
                   es = 0.61078*exp(17.27*t/(t+237.3)),
                   ea = (hurs/100)*es,
                   delta = 4098*es/(t+237.3)^2, #slope vapor pressure
                   gamma = 0.00163*P/(2.501-52.6361e-3*t), #psychrometric constant
                   ET0_11 = ((0.408*delta*Rn+gamma*900/(t+273)*0.748*sfcWind*(es-ea))/(delta+gamma*(1+0.34*0.748*sfcWind))))

cmip6_12 <- mutate(read.table("cmip6_december.txt"),
                   t = tas - 273.15,
                   pr_12 = pr*60*60*24, # mm/day
                   Rn = (hfls + hfss) * 86400 * 1e-6, # conversion from W/m2 to MJ/m2/day.
                   P = 101.3 * ((293-0.0065*z)/293)^(5.26),
                   es = 0.61078*exp(17.27*t/(t+237.3)),
                   ea = (hurs/100)*es,
                   delta = 4098*es/(t+237.3)^2, #slope vapor pressure
                   gamma = 0.00163*P/(2.501-52.6361e-3*t), #psychrometric constant
                   ET0_12 = ((0.408*delta*Rn+gamma*900/(t+273)*0.748*sfcWind*(es-ea))/(delta+gamma*(1+0.34*0.748*sfcWind))))

et0_monthly <- select(cmip6_annual, id_vars) %>%
  merge(select(cmip6_01, c(id_vars, "ET0_01", "pr_01")), by = id_vars) %>%
  merge(select(cmip6_02, c(id_vars, "ET0_02", "pr_02")), by = id_vars) %>%
  merge(select(cmip6_03, c(id_vars, "ET0_03", "pr_03")), by = id_vars) %>%
  merge(select(cmip6_04, c(id_vars, "ET0_04", "pr_04")), by = id_vars) %>%
  merge(select(cmip6_05, c(id_vars, "ET0_05", "pr_05")), by = id_vars) %>%
  merge(select(cmip6_06, c(id_vars, "ET0_06", "pr_06")), by = id_vars) %>%
  merge(select(cmip6_07, c(id_vars, "ET0_07", "pr_07")), by = id_vars) %>%
  merge(select(cmip6_08, c(id_vars, "ET0_08", "pr_08")), by = id_vars) %>%
  merge(select(cmip6_09, c(id_vars, "ET0_09", "pr_09")), by = id_vars) %>%
  merge(select(cmip6_10, c(id_vars, "ET0_10", "pr_10")), by = id_vars) %>%   
  merge(select(cmip6_11, c(id_vars, "ET0_11", "pr_11")), by = id_vars) %>%
  merge(select(cmip6_12, c(id_vars, "ET0_12", "pr_12")), by = id_vars)

et0_monthly$spr <- with(et0_monthly, pr_01*31 + pr_02 * 28 + pr_03 * 31 + pr_04 * 30 + pr_05 * 31 + pr_06 * 30 +
                          pr_07 * 31 + pr_08 * 31 + pr_09 * 30 + pr_10 * 31 + pr_11 * 30 + pr_12 * 31)
et0_monthly$sET0 <- with(et0_monthly, ET0_01*31 + ET0_02 * 28 + ET0_03 * 31 + ET0_04 * 30 + ET0_05 * 31 + ET0_06 * 30 +
                           ET0_07 * 31 + ET0_08 * 31 + ET0_09 * 30 + ET0_10 * 31 + ET0_11 * 30 + ET0_12 * 31)

cmip6 <- merge(cmip6_annual, et0_monthly, by = id_vars)
cmip6$AI <- with(cmip6, abs(spr/sET0))  
cmip6$AI_annual <- with(cmip6, abs((pr*60*60*24*365)/(ET0_annual*365)))

breaks.unesco <- c(-Inf,0,0.03,0.2,0.5,0.65,Inf)
cat.unesco <- c("(0,0.03]"= "Hyperarid", "(0.03,0.2]" = "Arid", "(0.2,0.5]" = "Semi-arid", "(0.5,0.65]" = "Dry subhumid","(0.65, Inf]" = "Humid", "(-Inf,0]" = "Cold")
#col.cat <- c("Hyperarid" = colorvec[2], "Arid"= colorvec[4],"Semi-arid"= colorvec[5], "Dry subhumid" = colorvec[8], "Humid"= colorvec[10], "Cold" = "powderblue")
col.cat <- c("Hyperarid" = "#D73027", "Arid"= "#FDAE61","Semi-arid"= "#FEE08B", "Dry subhumid" = "#ABD9E9", "Humid"= "#4575B4", "Cold" = "#DAE2F7")

cmip6$cat.AI <- cmip6$AI %>% cut(breaks = breaks.unesco) %>% revalue(cat.unesco)
cmip6$cat.AI[which(cmip6$sET0 < 400)] <- "Cold"
cmip6$cat.AI %>% unique()

cmip6$cat.AI <- cmip6$AI %>% cut(breaks = breaks.unesco) %>% 
  revalue(cat.unesco)


cmip6$cat.AI_annual <- cmip6$AI_annual %>% cut(breaks = breaks.unesco) %>% 
  revalue(cat.unesco)

cmip6$cat.AI_annual[which(cmip6$ET0_annual*365 < 400)] <- "Cold"

write.table(cmip6, "cmip6_AI.txt")
```

# Maps AI by model, all % calculated without Antarctica

```{r read-data, cache = T}
cmip6 <- read.table("cmip6_AI.txt")

breaks.unesco <- c(-Inf,0.03,0.2,0.5,0.65,Inf)
cat.unesco <- c("(-Inf,0.03]"= "Hyperarid", "(0.03,0.2]" = "Arid", "(0.2,0.5]" = "Semi-arid", "(0.5,0.65]" = "Dry subhumid","(0.65, Inf]" = "Humid")
cmip6$cat.AI <- factor(cmip6$cat.AI, levels =  c("Hyperarid","Arid","Semi-arid","Dry subhumid", "Humid","Cold"))
```

## 1850-1880

```{r maps-hist-1850-1880, fig.dim = c(15, 20)}
map_list <- list()

for(i in unique(cmip6$source)){
  g <- ggplot(data = subset(cmip6, period == "1850_1880" & source == i)) + geom_raster(aes(x=lon, y = lat,  fill = cat.AI))+
    scale_fill_manual(values = col.cat, na.translate = F)+
    labs(title = paste(i, "1850-1880", sep = ", "), fill = "")+
    theme_void()
  map_list[[i]] <- g
}

ggpubr::ggarrange(plotlist = map_list, ncol = 3, nrow = 5, common.legend = T, legend = "bottom")
```

```{r table-of-percent-1850}

sdmm <- cmip6 %>% subset(period == "1850_1880"  & !Continent %in% c("SOUTHERN","PACIFIC","ATLANTIC","INDIAN","ARCTIC") & lat > -55) %>%
  group_by(source, cat.AI) %>%
  summarise(count = n()) %>%
  ungroup() %>% group_by(source) %>% mutate(percent = round(count/sum(count)*100, 1), count = NULL) %>%
  ungroup() %>% group_by(cat.AI) %>% summarise(mean = round(mean(percent, na.rm = T),1), sd = round(sd(percent, na.rm = T),1)) %>%
  t() %>% as.data.frame() %>% setNames(c(.[1,1:6], "NA")) %>% .[-1, ] %>% sapply(as.numeric) %>% as.data.frame() %>%
  mutate("Sum drylands" = rowSums(.[c("Hyperarid","Arid","Semi-arid","Dry subhumid")]), "source" = c("mean","sd")) %>%
  select(c("source","Hyperarid","Arid","Semi-arid","Dry subhumid","Sum drylands","Humid","Cold","NA"))


tab <- cmip6 %>% subset(period == "1850_1880"  & !Continent %in% c("SOUTHERN","PACIFIC","ATLANTIC","INDIAN","ARCTIC") & lat > -55) %>%
  group_by(source, cat.AI) %>%
  summarise(count = n()) %>%
  ungroup() %>% group_by(source) %>% mutate(percent = round(count/sum(count)*100, 1), count = NULL) %>%
  reshape2::dcast(source~cat.AI) %>%
  mutate("Sum drylands" = rowSums(.[c("Hyperarid","Arid","Semi-arid","Dry subhumid")])) %>%
  select(c("source","Hyperarid","Arid","Semi-arid","Dry subhumid","Sum drylands","Humid","Cold","NA")) %>% 
  rbind(sdmm)


knitr::kable(tab, digits = 2, escape = F)  %>%
  kable_styling(bootstrap_options = "bordered") %>% 
  column_spec(c(6,9), italic = T, include_thead = T) %>% row_spec(c(14,15), bold = T) 
```

## 1970-2000

```{r maps-hist-1970-2000, fig.dim = c(15, 20)}
map_list <- list()

for(i in unique(cmip6$source)){
  g <- ggplot(data = subset(cmip6, period == "1970_2000" & source == i)) + geom_raster(aes(x=lon, y = lat,  fill = cat.AI))+
    scale_fill_manual(values = col.cat, na.translate = F)+
    labs(title = paste(i, "1970-2000", sep = ", "), fill = "")+
    theme_void()
  map_list[[i]] <- g
}

ggpubr::ggarrange(plotlist = map_list, ncol = 3, nrow = 5, common.legend = T, legend = "bottom")
```

```{r percent-of-each-class-by-by-source, fig.dim = c(16,16)}
library(ggrepel) 

pie <- cmip6 %>% subset(period == "1970_2000" & !Continent %in% c("SOUTHERN","PACIFIC","ATLANTIC","INDIAN","ARCTIC") & lat > -55) %>%
  select(c("source", "cat.AI")) %>%
  reshape2::melt(id.vars = c("source")) %>%
  group_by(source, value) %>%
  summarise(count = n()) %>%
  mutate(perc = round(count/sum(count)*100, 1),
         lab.y = rev(cumsum(rev(count))) - count*0.5) %>% 
  ungroup() 

pie_list <- list()

for(i in unique(pie$source)){
  p <- ggplot(subset(pie, source == i))+
    geom_bar(aes(x = "", y = count, fill = value), width = 1, stat = "identity")+
    coord_polar("y", start = 0)+
    scale_fill_manual(values = col.cat)+
    geom_label_repel(aes(x = "", y = lab.y, label = perc), nudge_x = 0.5)+
    labs(fill = "", title = i)+
    theme_void()+theme(legend.position = "right")
  pie_list[[i]] <- p
}

ggpubr::ggarrange(plotlist = pie_list, nrow = 4, ncol = 4, common.legend = T, legend = "bottom")
```

```{r pie-multimodel-cmip6}
pie <- cmip6 %>% subset(period == "1970_2000" & !Continent %in% c("SOUTHERN","PACIFIC","ATLANTIC","INDIAN","ARCTIC") & lat > -55) %>%
  select(c("cat.AI")) %>%
  table() %>% as.data.frame() %>%
  mutate(perc = round(Freq/sum(Freq)*100, 1),
   lab.y = rev(cumsum(rev(Freq))) - Freq*0.5) 

p <- ggplot(pie)+
    geom_bar(aes(x = "", y = Freq, fill = cat.AI), width = 1, stat = "identity")+
    coord_polar("y", start = 0)+
    scale_fill_manual(values = col.cat)+
    geom_label_repel(aes(x = "", y = lab.y, label = perc), nudge_x = 0.5)+
    labs(fill = "", title = "Multimodel CMIP6")+
    theme_void()+theme(legend.position = "right")

print(p)
```

```{r table-of-percent-ref}

sdmm <- cmip6 %>% subset(period == "1970_2000"  & !Continent %in% c("SOUTHERN","PACIFIC","ATLANTIC","INDIAN","ARCTIC") & lat > -55 ) %>%
  group_by(source, cat.AI) %>%
  summarise(count = n()) %>%
  ungroup() %>% group_by(source) %>% mutate(percent = round(count/sum(count)*100, 1), count = NULL) %>%
  ungroup() %>% group_by(cat.AI) %>% summarise(mean = round(mean(percent, na.rm = T),1), sd = round(sd(percent, na.rm = T),1)) %>%
  t() %>% as.data.frame() %>% setNames(c(.[1,1:6], "NA")) %>% .[-1, ] %>% sapply(as.numeric) %>% as.data.frame() %>%
  mutate("Sum drylands" = rowSums(.[c("Hyperarid","Arid","Semi-arid","Dry subhumid")]), "source" = c("mean","sd")) %>%
  select(c("source","Hyperarid","Arid","Semi-arid","Dry subhumid","Sum drylands","Humid","Cold","NA"))

tab <- cmip6 %>% subset(period == "1970_2000"   & !Continent %in% c("SOUTHERN","PACIFIC","ATLANTIC","INDIAN","ARCTIC") & lat > -55 ) %>%
  group_by(source, cat.AI) %>%
  summarise(count = n()) %>%
  ungroup() %>% group_by(source) %>% mutate(percent = round(count/sum(count)*100, 1), count = NULL) %>%
  reshape2::dcast(source~cat.AI) %>%
  mutate("Sum drylands" = rowSums(.[c("Hyperarid","Arid","Semi-arid","Dry subhumid")])) %>%
  select(c("source","Hyperarid","Arid","Semi-arid","Dry subhumid","Sum drylands","Humid","Cold","NA")) %>% 
  rbind(sdmm)


knitr::kable(tab, digits = 2, escape = F)  %>%
  kable_styling(bootstrap_options = "bordered") %>% 
  column_spec(c(6,9), italic = T, include_thead = T) %>% row_spec(c(14,15), bold = T) 
```

## 1985-2010

```{r maps-hist-1985-2010, fig.dim = c(15,20)}
map_list <- list()

for(i in unique(cmip6$source)){
  g <- ggplot(data = subset(cmip6, period == "1985_2015" & source == i)) + geom_raster(aes(x=lon, y = lat,  fill = cat.AI))+
    scale_fill_manual(values = col.cat, na.translate = F)+
    labs(title = paste(i, "1985-2010", sep = ", "), fill = "")+
    theme_void()
  map_list[[i]] <- g
}

ggpubr::ggarrange(plotlist = map_list, ncol = 3, nrow = 5, common.legend = T, legend = "bottom")
```

```{r table-of-percent-1985}

sdmm <- cmip6 %>% subset(period == "1985_2015"   & !Continent %in% c("SOUTHERN","PACIFIC","ATLANTIC","INDIAN","ARCTIC") & lat > -55 ) %>%
  group_by(source, cat.AI) %>%
  summarise(count = n()) %>%
  ungroup() %>% group_by(source) %>% mutate(percent = round(count/sum(count)*100, 1), count = NULL) %>%
  ungroup() %>% group_by(cat.AI) %>% summarise(mean = round(mean(percent, na.rm = T),1), sd = round(sd(percent, na.rm = T),1)) %>%
  t() %>% as.data.frame() %>% setNames(c(.[1,1:6], "NA")) %>% .[-1, ] %>% sapply(as.numeric) %>% as.data.frame() %>%
  mutate("Sum drylands" = rowSums(.[c("Hyperarid","Arid","Semi-arid","Dry subhumid")]), "source" = c("mean","sd")) %>%
  select(c("source","Hyperarid","Arid","Semi-arid","Dry subhumid","Sum drylands","Humid","Cold","NA"))


tab <- cmip6 %>% subset(period == "1985_2015"   & !Continent %in% c("SOUTHERN","PACIFIC","ATLANTIC","INDIAN","ARCTIC") & lat > -55 ) %>%
  group_by(source, cat.AI) %>%
  summarise(count = n()) %>%
  ungroup() %>% group_by(source) %>% mutate(percent = round(count/sum(count)*100, 1), count = NULL) %>%
  reshape2::dcast(source~cat.AI) %>%
  mutate("Sum drylands" = rowSums(.[c("Hyperarid","Arid","Semi-arid","Dry subhumid")])) %>%
  select(c("source","Hyperarid","Arid","Semi-arid","Dry subhumid","Sum drylands","Humid","Cold","NA")) %>% 
  rbind(sdmm)


knitr::kable(tab, digits = 2, escape = F)  %>%
  kable_styling(bootstrap_options = "bordered") %>% 
  column_spec(c(6,9), italic = T, include_thead = T) %>% row_spec(c(14,15), bold = T) 
```

## SSP 2-4.5 

###  2030-2060

```{r maps-ssp245-2030-2060, fig.dim = c(15,20)}
map_list <- list()

for(i in unique(cmip6$source)){
  g <- ggplot(data = subset(cmip6, period == "2030_2060" & model == "SSP245" & source == i)) + geom_raster(aes(x=lon, y = lat,  fill = cat.AI))+
    scale_fill_manual(values = col.cat, na.translate = F)+
    labs(title = paste(i, "2030-2060", sep = ", "), fill = "")+
    theme_void()
  map_list[[i]] <- g
}

ggpubr::ggarrange(plotlist = map_list, ncol = 3, nrow = 5, common.legend = T, legend = "bottom")
```

```{r table-of-percent-SSP245-2030}

sdmm <- cmip6 %>% subset(period == "2030_2060" & model == "SSP245" &!Continent %in% c("SOUTHERN","PACIFIC","ATLANTIC","INDIAN","ARCTIC")) %>%
  group_by(source, cat.AI) %>%
  summarise(count = n()) %>%
  ungroup() %>% group_by(source) %>% mutate(percent = round(count/sum(count)*100, 1), count = NULL) %>%
  ungroup() %>% group_by(cat.AI) %>% summarise(mean = round(mean(percent, na.rm = T),1), sd = round(sd(percent, na.rm = T),1)) %>%
  t() %>% as.data.frame() %>% setNames(c(.[1,1:6], "NA")) %>% .[-1, ] %>% sapply(as.numeric) %>% as.data.frame() %>%
  mutate("Sum drylands" = rowSums(.[c("Hyperarid","Arid","Semi-arid","Dry subhumid")]), "source" = c("mean","sd")) %>%
  select(c("source","Hyperarid","Arid","Semi-arid","Dry subhumid","Sum drylands","Humid","Cold","NA"))

tab <- cmip6 %>% subset(period == "2030_2060" & model == "SSP245"   & !Continent %in% c("SOUTHERN","PACIFIC","ATLANTIC","INDIAN","ARCTIC") & lat > -55 ) %>%
  group_by(source, cat.AI) %>%
  summarise(count = n()) %>%
  ungroup() %>% group_by(source) %>% mutate(percent = round(count/sum(count)*100, 1), count = NULL) %>%
  reshape2::dcast(source~cat.AI) %>%
  mutate("Sum drylands" = rowSums(.[c("Hyperarid","Arid","Semi-arid","Dry subhumid")])) %>%
  select(c("source","Hyperarid","Arid","Semi-arid","Dry subhumid","Sum drylands","Humid","Cold","NA")) %>% 
  rbind(sdmm)

knitr::kable(tab, digits = 2, escape = F)  %>%
  kable_styling(bootstrap_options = "bordered") %>% 
  column_spec(c(6,9), italic = T, include_thead = T) %>% row_spec(c(14,15), bold = T) 
```

### 2070-2100

```{r maps-ssp245-2070-2100, fig.dim = c(15,20)}
map_list <- list()

for(i in unique(cmip6$source)){
  g <- ggplot(data = subset(cmip6, period == "2070_2100" & model == "SSP245" & source == i)) + geom_raster(aes(x=lon, y = lat,  fill = cat.AI))+
    scale_fill_manual(values = col.cat, na.translate = F)+
    labs(title = paste(i, "2070-2100", sep = ", "), fill = "")+
    theme_void()
  map_list[[i]] <- g
}

ggpubr::ggarrange(plotlist = map_list, ncol = 3, nrow = 5, common.legend = T, legend = "bottom")
```

```{r table-of-percent-SSP245-2070}
sdmm <- cmip6 %>% subset(period == "2070_2100" & model == "SSP245" &!Continent %in% c("SOUTHERN","PACIFIC","ATLANTIC","INDIAN","ARCTIC")) %>%
  group_by(source, cat.AI) %>%
  summarise(count = n()) %>%
  ungroup() %>% group_by(source) %>% mutate(percent = round(count/sum(count)*100, 1), count = NULL) %>%
  ungroup() %>% group_by(cat.AI) %>% summarise(mean = round(mean(percent, na.rm = T),1), sd = round(sd(percent, na.rm = T),1)) %>%
  t() %>% as.data.frame() %>% setNames(c(.[1,1:6], "NA")) %>% .[-1, ] %>% sapply(as.numeric) %>% as.data.frame() %>%
  mutate("Sum drylands" = rowSums(.[c("Hyperarid","Arid","Semi-arid","Dry subhumid")]), "source" = c("mean","sd")) %>%
  select(c("source","Hyperarid","Arid","Semi-arid","Dry subhumid","Sum drylands","Humid","Cold","NA"))

tab <- cmip6 %>% subset(period == "2070_2100" & model == "SSP245"   & !Continent %in% c("SOUTHERN","PACIFIC","ATLANTIC","INDIAN","ARCTIC") & lat > -55) %>%
  group_by(source, cat.AI) %>%
  summarise(count = n()) %>%
  ungroup() %>% group_by(source) %>% mutate(percent = round(count/sum(count)*100, 1), count = NULL) %>%
  reshape2::dcast(source~cat.AI) %>%
  mutate("Sum drylands" = rowSums(.[c("Hyperarid","Arid","Semi-arid","Dry subhumid")])) %>%
  select(c("source","Hyperarid","Arid","Semi-arid","Dry subhumid","Sum drylands","Humid","Cold","NA")) %>% 
  rbind(sdmm)

knitr::kable(tab, digits = 2, escape = F)  %>%
  kable_styling(bootstrap_options = "bordered") %>% 
  column_spec(c(6,9), italic = T, include_thead = T) %>% row_spec(c(14,15), bold = T) 
```

## SSP 3-7.0

### 2030-2060

```{r maps-ssp370-2030-2060, fig.dim = c(15,20)}
map_list <- list()

for(i in unique(cmip6$source)){
  g <- ggplot(data = subset(cmip6, period == "2030_2060" & model == "SSP370" & source == i)) + geom_raster(aes(x=lon, y = lat,  fill = cat.AI))+
    scale_fill_manual(values = col.cat, na.translate = F)+
    labs(title = paste(i, "2030-2060", sep = ", "), fill = "")+
    theme_void()
  map_list[[i]] <- g
}

ggpubr::ggarrange(plotlist = map_list, ncol = 3, nrow = 5, common.legend = T, legend = "bottom")
```

```{r table-of-percent-SSP370-2030}
sdmm <- cmip6 %>% subset(period == "2030_2060" & model == "SSP370" &!Continent %in% c("SOUTHERN","PACIFIC","ATLANTIC","INDIAN","ARCTIC")) %>%
  group_by(source, cat.AI) %>%
  summarise(count = n()) %>%
  ungroup() %>% group_by(source) %>% mutate(percent = round(count/sum(count)*100, 1), count = NULL) %>%
  ungroup() %>% group_by(cat.AI) %>% summarise(mean = round(mean(percent, na.rm = T),1), sd = round(sd(percent, na.rm = T),1)) %>%
  t() %>% as.data.frame() %>% setNames(c(.[1,1:6], "NA")) %>% .[-1, ] %>% sapply(as.numeric) %>% as.data.frame() %>%
  mutate("Sum drylands" = rowSums(.[c("Hyperarid","Arid","Semi-arid","Dry subhumid")]), "source" = c("mean","sd")) %>%
  select(c("source","Hyperarid","Arid","Semi-arid","Dry subhumid","Sum drylands","Humid","Cold","NA"))

tab <- cmip6 %>% subset(period == "2030_2060" & model == "SSP370"  & !Continent %in% c("SOUTHERN","PACIFIC","ATLANTIC","INDIAN","ARCTIC") & lat > -55) %>%
  group_by(source, cat.AI) %>%
  summarise(count = n()) %>%
  ungroup() %>% group_by(source) %>% mutate(percent = round(count/sum(count)*100, 1), count = NULL) %>%
  reshape2::dcast(source~cat.AI) %>%
  mutate("Sum drylands" = rowSums(.[c("Hyperarid","Arid","Semi-arid","Dry subhumid")])) %>%
  select(c("source","Hyperarid","Arid","Semi-arid","Dry subhumid","Sum drylands","Humid","Cold","NA")) %>% 
  rbind(sdmm)

knitr::kable(tab, digits = 2, escape = F)  %>%
  kable_styling(bootstrap_options = "bordered") %>% 
  column_spec(c(6,9), italic = T, include_thead = T) %>% row_spec(c(14,15), bold = T) 
```

### 2070-2100

```{r maps-ssp370-2070-2100, fig.dim = c(15,20)}
map_list <- list()

for(i in unique(cmip6$source)){
  g <- ggplot(data = subset(cmip6, period == "2070_2100" & model == "SSP370" & source == i)) + geom_raster(aes(x=lon, y = lat,  fill = cat.AI))+
    scale_fill_manual(values = col.cat, na.translate = F)+
    labs(title = paste(i, "2070-2100", sep = ", "), fill = "")+
    theme_void()
  map_list[[i]] <- g
}

ggpubr::ggarrange(plotlist = map_list, ncol = 3, nrow = 5, common.legend = T, legend = "bottom")
```

```{r table-of-percent-SSP370-2070}
sdmm <- cmip6 %>% subset(period == "2070_2100" & model == "SSP370" &!Continent %in% c("SOUTHERN","PACIFIC","ATLANTIC","INDIAN","ARCTIC")) %>%
  group_by(source, cat.AI) %>%
  summarise(count = n()) %>%
  ungroup() %>% group_by(source) %>% mutate(percent = round(count/sum(count)*100, 1), count = NULL) %>%
  ungroup() %>% group_by(cat.AI) %>% summarise(mean = round(mean(percent, na.rm = T),1), sd = round(sd(percent, na.rm = T),1)) %>%
  t() %>% as.data.frame() %>% setNames(c(.[1,1:6], "NA")) %>% .[-1, ] %>% sapply(as.numeric) %>% as.data.frame() %>%
  mutate("Sum drylands" = rowSums(.[c("Hyperarid","Arid","Semi-arid","Dry subhumid")]), "source" = c("mean","sd")) %>%
  select(c("source","Hyperarid","Arid","Semi-arid","Dry subhumid","Sum drylands","Humid","Cold","NA"))

tab <- cmip6 %>% subset(period == "2070_2100" & model == "SSP370"  & !Continent %in% c("SOUTHERN","PACIFIC","ATLANTIC","INDIAN","ARCTIC") & lat > -55) %>%
  group_by(source, cat.AI) %>%
  summarise(count = n()) %>%
  ungroup() %>% group_by(source) %>% mutate(percent = round(count/sum(count)*100, 1), count = NULL) %>%
  reshape2::dcast(source~cat.AI) %>%
  mutate("Sum drylands" = rowSums(.[c("Hyperarid","Arid","Semi-arid","Dry subhumid")])) %>%
  select(c("source","Hyperarid","Arid","Semi-arid","Dry subhumid","Sum drylands","Humid","Cold","NA")) %>% 
  rbind(sdmm)

knitr::kable(tab, digits = 2, escape = F)  %>%
  kable_styling(bootstrap_options = "bordered") %>% 
  column_spec(c(6,9), italic = T, include_thead = T) %>% row_spec(c(14,15), bold = T) 
```

## SSP 5-8.5

### 2030-2060

```{r maps-ssp585-2030-2060, fig.dim = c(15,20)}
map_list <- list()

for(i in unique(cmip6$source)){
  g <- ggplot(data = subset(cmip6, period == "2030_2060" & model == "SSP585" & source == i)) + geom_raster(aes(x=lon, y = lat,  fill = cat.AI))+
    scale_fill_manual(values = col.cat, na.translate = F)+
    labs(title = paste(i, "2030-2060", sep = ", "), fill = "")+
    theme_void()
  map_list[[i]] <- g
}

ggpubr::ggarrange(plotlist = map_list, ncol = 3, nrow = 5, common.legend = T, legend = "bottom")
```

```{r table-of-percent-SSP585-2030}
sdmm <- cmip6 %>% subset(period == "2030_2060" & model == "SSP585" &!Continent %in% c("SOUTHERN","PACIFIC","ATLANTIC","INDIAN","ARCTIC")) %>%
  group_by(source, cat.AI) %>%
  summarise(count = n()) %>%
  ungroup() %>% group_by(source) %>% mutate(percent = round(count/sum(count)*100, 1), count = NULL) %>%
  ungroup() %>% group_by(cat.AI) %>% summarise(mean = round(mean(percent, na.rm = T),1), sd = round(sd(percent, na.rm = T),1)) %>%
  t() %>% as.data.frame() %>% setNames(c(.[1,1:6], "NA")) %>% .[-1, ] %>% sapply(as.numeric) %>% as.data.frame() %>%
  mutate("Sum drylands" = rowSums(.[c("Hyperarid","Arid","Semi-arid","Dry subhumid")]), "source" = c("mean","sd")) %>%
  select(c("source","Hyperarid","Arid","Semi-arid","Dry subhumid","Sum drylands","Humid","Cold","NA"))

tab <- cmip6 %>% subset(period == "2030_2060" & model == "SSP585"  & !Continent %in% c("SOUTHERN","PACIFIC","ATLANTIC","INDIAN","ARCTIC") & lat > -55) %>%
  group_by(source, cat.AI) %>%
  summarise(count = n()) %>%
  ungroup() %>% group_by(source) %>% mutate(percent = round(count/sum(count)*100, 1), count = NULL) %>%
  reshape2::dcast(source~cat.AI) %>%
  mutate("Sum drylands" = rowSums(.[c("Hyperarid","Arid","Semi-arid","Dry subhumid")])) %>%
  select(c("source","Hyperarid","Arid","Semi-arid","Dry subhumid","Sum drylands","Humid","Cold","NA")) %>% 
  rbind(sdmm)

knitr::kable(tab, digits = 2, escape = F)  %>%
  kable_styling(bootstrap_options = "bordered") %>% 
  column_spec(c(6,9), italic = T, include_thead = T) %>% row_spec(c(14,15), bold = T) 
```

### 2070-2100

```{r maps-ssp585-2070-2100, fig.dim = c(15,20)}
map_list <- list()

for(i in unique(cmip6$source)){
  g <- ggplot(data = subset(cmip6, period == "2070_2100" & model == "SSP585" & source == i)) + geom_raster(aes(x=lon, y = lat,  fill = cat.AI))+
    scale_fill_manual(values = col.cat, na.translate = F)+
    labs(title = paste(i, "2070-2100", sep = ", "), fill = "")+
    theme_void()
  map_list[[i]] <- g
}

ggpubr::ggarrange(plotlist = map_list, ncol = 3, nrow = 5, common.legend = T, legend = "bottom")
```

```{r table-of-percent-SSP585-2070}
sdmm <- cmip6 %>% subset(period == "2070_2100" & model == "SSP585" &!Continent %in% c("SOUTHERN","PACIFIC","ATLANTIC","INDIAN","ARCTIC")) %>%
  group_by(source, cat.AI) %>%
  summarise(count = n()) %>%
  ungroup() %>% group_by(source) %>% mutate(percent = round(count/sum(count)*100, 1), count = NULL) %>%
  ungroup() %>% group_by(cat.AI) %>% summarise(mean = round(mean(percent, na.rm = T),1), sd = round(sd(percent, na.rm = T),1)) %>%
  t() %>% as.data.frame() %>% setNames(c(.[1,1:6], "NA")) %>% .[-1, ] %>% sapply(as.numeric) %>% as.data.frame() %>%
  mutate("Sum drylands" = rowSums(.[c("Hyperarid","Arid","Semi-arid","Dry subhumid")]), "source" = c("mean","sd")) %>%
  select(c("source","Hyperarid","Arid","Semi-arid","Dry subhumid","Sum drylands","Humid","Cold","NA"))

tab <- cmip6 %>% subset(period == "2070_2100" & model == "SSP585"  & !Continent %in% c("SOUTHERN","PACIFIC","ATLANTIC","INDIAN","ARCTIC") & lat > -55) %>%
  group_by(source, cat.AI) %>%
  summarise(count = n()) %>%
  ungroup() %>% group_by(source) %>% mutate(percent = round(count/sum(count)*100, 1), count = NULL) %>%
  reshape2::dcast(source~cat.AI) %>%
  mutate("Sum drylands" = rowSums(.[c("Hyperarid","Arid","Semi-arid","Dry subhumid")])) %>%
  select(c("source","Hyperarid","Arid","Semi-arid","Dry subhumid","Sum drylands","Humid","Cold","NA")) %>% 
  rbind(sdmm)

knitr::kable(tab, digits = 2, escape = F)  %>%
  kable_styling(bootstrap_options = "bordered") %>% 
  column_spec(c(6,9), italic = T, include_thead = T) %>% row_spec(c(13,14), bold = T) 
```

# Multimodel average

Here we compute the multimodel mean and median aridity index. We compare the category obtained by using the mean aridity index (cat.mean) to the category obtained in most models (cat.maj).

## Summary dataframe

### Difference with AI ref for all SSP 

Compute the difference between future AI, precipitation and temperature for future scenarios, for all models. 

```{r diff-AI-cmip6, eval = F}
cmip6 <- read.table("cmip6_AI.txt")

cmip6.ref <- cmip6 %>% filter(period == "1970_2000") %>% select(c("lon","lat","model","period","source","AI","spr","t","sET0")) %>% setNames(c("lon","lat","model","period","source","AI.ref","spr.ref","t.ref","ET0.ref"))

cmip6ext <- merge(cmip6, select(cmip6.ref, c("lon","lat","source","AI.ref","spr.ref","t.ref","ET0.ref")) , by = c("lon","lat","source"), all = T) %>%  mutate(diff.AI = AI - AI.ref, diff.pr = spr - spr.ref, diff.t = t - t.ref, diff.ET0 = sET0 - ET0.ref)

write.table(cmip6ext, "cmip6ext.txt")
```

Create a summary table with multimodel average. 

```{r agreement-between-models, eval = F}
cmip6 <- read.table("cmip6ext.txt")
cmip6s <- cmip6 %>% 
  group_by(lon,lat,Continent, Type, Name, Acronym, lm, model, period, z) %>% 
  dplyr::summarise(t.mean = mean(t, na.rm = T), t.sd = sd(t, na.rm = T), # t in °C
                   pr.mean = mean(spr, na.rm = T) , pr.sd = sd(spr, na.rm = T),  # pr in mm/y
                   sfcWind.mean = mean(sfcWind, na.rm = T), sfcWind.sd = sd(sfcWind, na.rm = T),
                   Rn.mean = mean(Rn, na.rm = T), Rn.sd = sd(Rn, na.rm = T), # in MJ/m2/day
                   ea.mean = mean(ea, na.rm = T), ea.sd = sd(ea, na.rm = T), # in kPa
                   ET0 = mean(sET0, na.rm = T), ET0.sd = sd(sET0, na.rm = T), # in mm/y
                   ET0.ref = mean(ET0.ref, na.rm = T), ET0.ref.sd = sd(ET0.ref, na.rm = T), # in mm/y
                   diff.ET0.mean = mean(diff.ET0, na.rm = T), diff.ET0.sd = sd(diff.ET0, na.rm = T),
                   AI.mean = mean(AI, na.rm = T), AI.sd = sd(AI, na.rm = T),
                   AI.med = median(AI, na.rm = T),
                   AI.ref.mean = mean(AI.ref, na.rm = T), AI.ref.sd = sd(AI.ref, na.rm = T),
                   AI.ref.med = median(AI.ref, na.rm = T),
                   diff.AI.mean = mean(diff.AI, na.rm = T), diff.AI.sd = sd(diff.AI, na.rm = T),
                   diff.AI.med = median(diff.AI, na.rm = T), 
                   spr.ref.mean = mean(spr.ref, na.rm = T), spr.ref.sd = sd(spr.ref, na.rm = T),
                   diff.pr.mean = mean(diff.pr, na.rm = T), diff.pr.sd = sd(diff.pr, na.rm = T),
                   t.ref.mean = mean(t.ref, na.rm = T), t.ref.sd = sd(t.ref, na.rm =T),
                   diff.t.mean = mean(diff.t, na.rm = T), diff.t.sd = sd(diff.t, na.rm = T),
                   Hyperarid = sum(cat.AI == "Hyperarid"),
                   Arid = sum(cat.AI == "Arid"),
                   'Semi-arid' = sum(cat.AI == "Semi-arid"),
                   'Dry subhumid' = sum(cat.AI == "Dry subhumid") ,
                   'Humid' = sum(cat.AI == "Humid"),
                   'Cold' = sum(cat.AI == "Cold"),
                   'NA' = sum(is.na(cat.AI)),
                   nb.models = n(),
                    plus.AI = table(diff.AI > 0)["TRUE"],
                    minus.AI = table(diff.AI < 0)["TRUE"]) %>%
ungroup() %>% as.data.frame()

cmip6s$cat.maj <- colnames(select(cmip6s, c("Hyperarid","Arid",'Semi-arid',"Dry subhumid",'Humid',"Cold","NA")))[max.col(select(cmip6s, c("Hyperarid","Arid",'Semi-arid',"Dry subhumid",'Humid',"Cold","NA")),ties.method = "first")]
cmip6s$nb.maj <- cmip6s %>% select(c("Hyperarid","Arid",'Semi-arid',"Dry subhumid",'Humid',"Cold","NA")) %>% apply(1, max)
cmip6s$prop.maj <- with(cmip6s, nb.maj / nb.models)



breaks.unesco <- c(-Inf,0,0.03,0.2,0.5,0.65,Inf)
cat.unesco <- c("(0,0.03]"= "Hyperarid", "(0.03,0.2]" = "Arid", "(0.2,0.5]" = "Semi-arid", "(0.5,0.65]" = "Dry subhumid","(0.65, Inf]" = "Humid", "(-Inf,0]" = "Cold")

cmip6s$cat.AI <- cmip6s$AI.mean %>% cut(breaks = breaks.unesco) %>% revalue(cat.unesco)
cmip6s$cat.AI[which(cmip6s$ET0 < 400)] <- "Cold"

cmip6s$cat.AI.ref <- cmip6s$AI.ref.mean %>% cut(breaks = breaks.unesco) %>% revalue(cat.unesco)
cmip6s$cat.AI.ref[which(cmip6s$ET0.ref < 400)] <- "Cold"

cmip6s$same.catAI <- with(cmip6s, ifelse(cat.maj == cat.AI, "same", "diff"))

cmip6s$cat.dryer <- with(cmip6s, 
          ifelse(cat.AI.ref == "Arid" & cat.AI %in% c("Cold","Humid","Dry subhumid","Semi-arid"), "wetter", 
          ifelse(cat.AI.ref == "Arid" & cat.AI == "Hyperarid", "dryer",
          ifelse(cat.AI.ref == "Semi-arid" & cat.AI %in% c("Cold","Humid","Dry subhumid"), "wetter",
          ifelse(cat.AI.ref == "Semi-arid" & cat.AI %in% c("Hyperarid","Arid"), "dryer",
          ifelse(cat.AI.ref == "Dry subhumid" & cat.AI %in% c("Cold","Humid"), "wetter",
          ifelse(cat.AI.ref == "Dry subhumid" & cat.AI %in% c("Semi-arid","Arid","Hyperarid"), "dryer",
          ifelse(cat.AI.ref == "Humid" & cat.AI == "Cold","wetter",
          ifelse(cat.AI.ref == "Humid" & cat.AI %in% c("Dry subhumid","Semi-arid","Arid","Hyperarid"), "dryer",
          ifelse(cat.AI.ref == "Cold" & cat.AI %in% c("Humid","Dry subhumid","Semi-arid","Arid","Hyperarid"), "dryer",
          ifelse(cat.AI.ref == "Hyperarid" & cat.AI %in% c("Cold","Humid","Dry subhumid","Semi-arid","Arid"), "wetter",
          ifelse(cat.AI.ref == cat.AI, "same",
          NA))))))))))))

cmip6sref <- cmip6s %>% filter(period == "1970_2000") %>% select(c("lon","lat","model","period","cat.maj")) %>% setNames(c("lon","lat","model","period","cat.maj.ref"))

cmip6s <- merge(cmip6s, select(cmip6sref, c("lon","lat","cat.maj.ref")), 
                 by = c("lon","lat"), all.x = T) 


cmip6s$cat.maj.dryer <- with(cmip6s, 
        ifelse(cat.maj.ref == "Arid" & cat.maj %in% c("Cold","Humid","Dry subhumid","Semi-arid"), "wetter", 
        ifelse(cat.maj.ref == "Arid" & cat.maj == "Hyperarid", "dryer",
        ifelse(cat.maj.ref == "Semi-arid" & cat.maj %in% c("Cold","Humid","Dry subhumid"), "wetter",
        ifelse(cat.maj.ref == "Semi-arid" & cat.maj %in% c("Hyperarid","Arid"), "dryer",
        ifelse(cat.maj.ref == "Dry subhumid" & cat.maj %in% c("Cold","Humid"), "wetter",
        ifelse(cat.maj.ref == "Dry subhumid" & cat.maj %in% c("Semi-arid","Arid","Hyperarid"), "dryer",
        ifelse(cat.maj.ref == "Humid" & cat.maj == "Cold","wetter",
        ifelse(cat.maj.ref == "Humid" & cat.maj %in% c("Dry subhumid","Semi-arid","Arid","Hyperarid"), "dryer",
        ifelse(cat.maj.ref == "Cold" & cat.maj %in% c("Humid","Dry subhumid","Semi-arid","Arid","Hyperarid"), "dryer",
        ifelse(cat.maj.ref == "Hyperarid" & cat.maj %in% c("Cold","Humid","Dry subhumid","Semi-arid","Arid"), "wetter",
        ifelse(cat.maj.ref == cat.maj, "same",
        NA))))))))))))


cmip6s$Continent[which(cmip6s$Continent == "EUROPE-AFRICA")] <- "MED-BASIN"

write.table(cmip6s, "cmip6s.txt")

```

```{r read-cmip6s}
cmip6s <- read.table("cmip6s.txt")
```

## Compare cat mean and cat maj

cat mean = category of aridity obtained by using the multimodel mean AI
cat maj = category of aridity obtained by using the category majoritarily obtained in the 13 models

### Where are cat mean and cat maj different

```{r table-comparing-catmaj-catmean, fig.dim = c(8,6)}

cmip6s %>% group_by(model, period) %>% summarise(maj.equal.mean = table(same.catAI)[2], maj.diff.mean = table(same.catAI)[1], prop.same = round(maj.equal.mean/n()*100,1)) %>% knitr::kable(col.names = c("Model","Period","Number of gridcells in which cat.maj = cat.mean", "Number of gridcells in which cat.maj /= cat.mean", "Proportion of cells with the same category (%)")) %>% kableExtra::kable_styling(bootstrap_options = "bordered")

cmip6s$cat.AI <- factor(cmip6s$cat.AI, levels =  c("Hyperarid","Arid","Semi-arid","Dry subhumid", "Humid","Cold"))
cmip6s$cat.maj <- factor(cmip6s$cat.maj, levels =  c("Hyperarid","Arid","Semi-arid","Dry subhumid", "Humid","Cold"))

ggplot(subset(cmip6s, cat.AI != "Cold"))+
  geom_col(aes(x= period, y = cat.AI, fill = cat.AI), just = -0.2, width = 0.3)+
  geom_col(aes(x = period, y = cat.maj, fill = cat.maj), just = 1.2, width = 0.3)+
  scale_fill_manual(values = col.cat, na.translate = F)+
  labs(fill = "", y ="")+
  facet_grid(rows = vars(model), switch = "y")+
  theme_minimal()+theme(axis.text.y = element_blank())



tab <- rbind(table(cmip6s$cat.AI)/length(cmip6s$cat.AI)*100, table(cmip6s$cat.maj)/length(cmip6s$cat.maj)*100) %>% t() %>% as.data.frame() %>% setNames(c("Cat mean", "Cat maj"))
knitr::kable(tab, digits =2, caption = "Proportion of gridcells in each aridity category") %>% kable_styling(bootstrap_options = "bordered")

```

```{r maps-of-cat-differences, fig.dim = c(10,27)}
map_list <- list()

for(i in unique(cmip6s$model)){
  df <- subset(cmip6s, model == i)
  for(j in unique(df$period)){
    index <- paste(i, j, sep = "")
    g <- ggplot()+geom_raster(data = subset(df, period == j & same.catAI == "diff"), aes(x = lon, y = lat, fill = cat.AI))+
      borders(colour = "grey60")+
      scale_fill_manual(values =  col.cat, na.translate = F)+
      labs(title = paste(i, j , sep = " "), fill = "")+
      theme_void()+ylim(-55,90)+
      theme(legend.position = "bottom")
    map_list[[index]] <- g
  }
}

col1 <- ggarrange(plotlist = map_list, nrow = 9, ncol = 1, common.legend = T, legend = "bottom") %>% annotate_figure(top = "cat.mean for gridcells in which\ncat.maj and cat.mean are different")

map_list2 <- list()

for(i in unique(cmip6s$model)){
  df <- subset(cmip6s, model == i)
  for(j in unique(df$period)){
    index <- paste(i, j, sep = "")
    g <- ggplot()+geom_raster(data = subset(df, period == j & same.catAI == "diff"), aes(x = lon, y = lat, fill = cat.maj))+
      borders(colour = "grey60")+
      scale_fill_manual(values =  col.cat, na.translate = F)+
      labs(title = paste(i, j , sep = " "), fill = "")+
      theme_void()+ylim(-55,90)+
      theme(legend.position = "bottom")
    map_list2[[index]] <- g
  }
}

col2 <- ggarrange(plotlist = map_list2, nrow = 9, ncol = 1, common.legend = T, legend = "bottom") %>% annotate_figure(top = "cat.maj for gridcells in which\ncat.maj and cat.mean are different")

ggarrange(col1,col2,ncol = 2, common.legend = T)

```

### Plot cat maj and cat mean for reference period

```{r ref-cat-maj, fig.dim = c(10,4), eval = F}
ggpubr::ggarrange(plotlist = list(
  
  ggplot() + 
    geom_raster(data = subset(cmip6s, period == "1970_2000"), aes(x=lon, y = lat,  fill = cat.maj))+
    geom_point(data = subset(cmip6s, period == "1970_2000" & nb.maj > 10), aes(x =lon, y = lat), shape = "'", size = 0.5, col = "grey60")+
    scale_fill_manual(values = col.cat)+
    labs(title = "Category chosen in majority by models", fill = "")+
    ylim(-55,90)+
    theme_void(base_size = 15)+
    theme(legend.position = "none"),
  
  ggplot() + 
    geom_raster(data = subset(cmip6s, period == "1970_2000"), aes(x=lon, y = lat,  fill = cat.AI))+
   geom_point(data = subset(cmip6s, period == "1970_2000" & nb.maj > 10), aes(x =lon, y = lat), shape = "'", size = 0.5, col = "grey60")+
    scale_fill_manual(values = col.cat)+
    labs(title = "Multimodel average categories of AI", fill = "")+
    ylim(-55,90)+
    theme_void(base_size = 15)+
    theme(legend.position = "none")),
  
  ncol = 2)

```


```{r is-cat-maj-same-as-recomputed-cat, fig.dim = c(10,10)}
map_list <- list()

for(i in c("1850_1880","1970_2000","1985_2015")){
  g <- ggplot() + geom_raster(data = subset(cmip6s, period == i & model == "historical"), aes(x=lon, y = lat,  fill = cat.maj))+
    geom_point(data = subset(cmip6s, period == i & model == "historical" & nb.maj > 10), aes(x= lon, y = lat), shape = "'", size = 0.5, col = "grey60")+
    scale_fill_manual(values =  col.cat, na.translate = F)+
    labs(title = i, fill = "")+
    theme_void()+ylim(-55,90)+
    theme(legend.position = "bottom")
  map_list[[i]] <- g
}


map_list2 <- list()

for(i in c("1850_1880","1970_2000","1985_2015")){
  g <- ggplot() + geom_raster(data = subset(cmip6s, period == i & model == "historical"), aes(x=lon, y = lat,  fill = cat.AI))+
   # geom_point(data = subset(cmip6s, period == i & model == "historical" & nb.maj > 10), aes(x= lon, y = lat), size = 0.5, col = "grey60")+
    scale_fill_manual(values =  col.cat, na.translate = F)+
    labs(title = i, fill = "")+
    theme_void()+ylim(-55,90)+
    theme(legend.position = "bottom")
  map_list2[[i]] <- g
}


ggarrange(
  annotate_figure(ggarrange(plotlist = map_list, nrow = 3, common.legend = T, legend = "bottom"), top = "Majority category among models"),
  annotate_figure(ggarrange(plotlist = map_list2, nrow = 3, common.legend = T, legend = "bottom"), top = "Multimodel average category"),
  ncol = 2, legend = "bottom")
```

```{r is-cat-maj-same-as-recomputed-cat-future, fig.dim = c(10,10)}
map_list <- list()

for(i in c("SSP245","SSP370","SSP585")){
  for (j in c("2030_2060","2070_2100")){
    index <- paste(i, j, sep = " , ")
    g <- ggplot() + geom_raster(data = subset(cmip6s, period == j & model == i), aes(x=lon, y = lat,  fill = cat.maj))+
         geom_point(data = subset(cmip6s, period == j & model == i & nb.maj > 10), aes(x= lon, y = lat), shape = "'", size = 0.5, col = "grey60")+
      scale_fill_manual(values = col.cat, na.translate = F)+
      labs(fill = "Cat AI maj", title = paste(i, j, sep = ", "))+
      theme_void()+ylim(-55,90)+
      theme(legend.position = "bottom")
    map_list[[index]] <- g
  }}


map_list2 <- list()

for(i in c("SSP245","SSP370","SSP585")){
  for (j in c("2030_2060","2070_2100")){
    index <- paste(i, j, sep = " , ")
    g <- ggplot() + geom_raster(data = subset(cmip6s, period == j & model == i), aes(x=lon, y = lat,  fill = cat.AI))+
    #     geom_point(data = subset(cmip6s, period == j & model == i & nb.maj > 10), aes(x= lon, y = lat), shape = "'", size = 0.5, col = "grey60")+
      scale_fill_manual(values = col.cat, na.translate = F)+
      labs(fill = "Cat AI mean", title = paste(i, j, sep = ", "))+
      theme_void()+ylim(-55,90)+
      theme(legend.position = "bottom")
    map_list2[[index]] <- g
  }}

ggarrange(plotlist = map_list, nrow = 3, ncol = 2, common.legend = T, legend = "bottom")
ggarrange(plotlist = map_list2, nrow = 3, ncol = 2, common.legend = T, legend = "bottom")

```



## Multimodel maps of cat mean

```{r map-historical-changes-AI-catmean, fig.dim = c(3, 8)}

cmip6s$cat.AI <- factor(cmip6s$cat.AI, levels = c("Hyperarid","Arid","Semi-arid","Dry subhumid","Humid","Cold"))

map_list <- list()

for(i in c("1850_1880", "1970_2000","1985_2015")){
  g <- ggplot() + 
    geom_tile(data = subset(cmip6s, period == i & model == "historical"), aes(x=lon, y = lat,  fill = cat.AI))+
    geom_point(data = subset(cmip6s,nb.maj > 10), aes(x = lon, y = lat), shape = ".", col = "grey20", size = 0.1)+
    borders(colour = "grey60")+
    scale_fill_manual(values =  col.cat, na.translate = F)+
    labs(title = paste("Category AI",i , sep = " "), fill = "")+
    theme_void()+ylim(-55,90)+
    theme(legend.position = "bottom")
  map_list[[i]] <- g
}

ggarrange(plotlist = map_list, ncol = 1, nrow = 3, common.legend = T, legend = "bottom")

```

```{r map-future-categories-AI, fig.dim = c(10,10)}

map_list <- list()

for(i in c("SSP245","SSP370","SSP585")){
  for (j in c("2030_2060","2070_2100")){
    index <- paste(i, j, sep = " , ")
    g <- ggplot() + 
      geom_tile(data = subset(cmip6s, period == j & model == i), aes(x=lon, y = lat,  fill = cat.AI))+
      geom_point(data = subset(cmip6s,nb.maj > 10), aes(x = lon, y = lat), shape = ".", col = "grey20", size = 0.1)+
      borders(colour = "grey60")+
      scale_fill_manual(values =  col.cat, na.translate = F)+
      labs(title = paste("Category AI",index, sep = " "), fill = "")+
      theme_void()+ylim(-55,90)+
      theme(legend.position = "bottom")
    map_list[[index]] <- g
  }}

ggpubr::ggarrange(plotlist = map_list,ncol = 2, nrow = 3, common.legend = T, legend = "bottom")
```

```{r map-historical-changes-AI-catmean-just-borders, fig.dim = c(10, 6)}

cmip6s$cat.AI <- factor(cmip6s$cat.AI, levels = c("Hyperarid","Arid","Semi-arid","Dry subhumid","Humid","Cold"))

ggplot() + 
    geom_tile(data = subset(cmip6s, period == "1970_2000" & model == "historical"), aes(x=lon, y = lat,  fill = cat.AI))+
    borders(colour = "grey60")+
    scale_fill_manual(values =  col.cat, na.translate = F)+
    labs(title = "Category AI 1970-2000", fill = "")+
    theme_void()+ylim(-55,90)+
    theme(legend.position = "bottom")

ggplot() + 
    geom_tile(data = subset(cmip6s, period == "2070_2100" & model == "SSP585"), aes(x=lon, y = lat,  fill = cat.AI))+
    borders(colour = "grey60")+
    scale_fill_manual(values =  col.cat, na.translate = F)+
    labs(title = "Category AI 2070-2100, SSP 585", fill = "")+
    theme_void()+ylim(-55,90)+
    theme(legend.position = "bottom")
  map_list[[i]] <- g


```


## Table of multimodel average category in percent

```{r percent-categories}
# cmip6s$cat.AI <- cmip6s$AI.mean %>% cut(breaks = breaks.unesco) %>% revalue(cat.unesco)
# cmip6s$cat.AI[which(cmip6s$ET0 < 400)] <- "Cold"
cmip6s$cat.AI.med <- cmip6s$AI.med %>% cut(breaks = breaks.unesco) %>% revalue(cat.unesco)
cmip6s$cat.AI.med[which(cmip6s$ET0 < 400)] <- "Cold"

table.percent <- cmip6s %>%  subset(!Continent %in% c("SOUTHERN","PACIFIC","ATLANTIC","INDIAN","ARCTIC") & lat > -55) %>%
  select(period, model, cat.AI) %>% group_by(period, model, cat.AI) %>% summarise(count = n()) %>% 
  ungroup() %>% group_by(period, model) %>% mutate(percent = round(count/sum(count)*100,1))

table.percent.med <- cmip6s %>%  subset(!Continent %in% c("SOUTHERN","PACIFIC","ATLANTIC","INDIAN","ARCTIC") & lat > -55) %>%
  select(period, model, cat.AI.med) %>% setNames(c("period","model","cat.AI")) %>% group_by(period, model, cat.AI) %>% summarise(count.med = n()) %>% 
  ungroup() %>% group_by(period, model) %>% mutate(percent.med = round(count.med/sum(count.med)*100,1))
  
table <- merge(table.percent, table.percent.med, by = c("period","model","cat.AI"), all.x = T) 
kable(table)

ggplot(subset(table, period == "1970_2000"), aes(x = cat.AI))+
  geom_col(aes(y = percent, fill = "mean"), position = position_nudge(x = - 0.2), width = 0.4)+
  geom_col(aes(y = percent.med, fill = "med"), position = position_nudge(x = + 0.2), width = 0.4)+
  scale_fill_manual(values = c("#ABD9E9", "#FDAE61"))+
  labs(x="Aridity category", fill = "Using mean or median")+
  theme_minimal()


```

# Maps multimodel AI

```{r aridity-index}
quantile(cmip6s$AI.mean, probs = seq(0,1,0.1), na.rm = T) 

ai.breaks <- c(0,0.03,0.2,0.5,0.65,1,10,20,30)
colscale <- brewer.pal(n = 11, name ="RdYlBu")[-c(5,6)]
```

```{r map-historical-changes-quantiles-AI, fig.dim = c(5,10)}
map_list <- list()

for(i in c("1850_1880", "1970_2000","1985_2015")){
  g <- ggplot() + geom_raster(data = subset(cmip6s, period == i & model == "historical"), aes(x=lon, y = lat,  fill = AI.mean))+
    borders(colour = "grey60")+
    binned_scale(aesthetics = "fill", breaks = ai.breaks, palette = function(x) colscale,
                 guide = guide_legend(label.theme = element_text(angle = 0)))+
    labs(title = i, fill = "Aridity index")+
    theme_void()+ylim(-55,90)+
    theme(legend.position = "right")
  map_list[[i]] <- g
}

ggpubr::ggarrange(plotlist = map_list, ncol = 1, nrow = 3, common.legend = T, legend = "bottom")

```

```{r map-future-changes-quantiles-AI, fig.dim = c(10,10)}

map_list <- list()

for(i in c("SSP245","SSP370","SSP585")){
  for (j in c("2030_2060","2070_2100")){
    index <- paste(i, j, sep = " , ")
    g <- ggplot() + geom_raster(data = subset(cmip6s, period == j & model == i), aes(x=lon, y = lat,  fill = AI.mean))+
      borders(colour = "grey60")+
      binned_scale(aesthetics = "fill", breaks = ai.breaks, palette = function(x) colscale,
                   guide = guide_legend(label.theme = element_text(angle = 0)))+
      labs(fill = "Aridity index", title = index)+
      theme_void()+ylim(-55,90)+
      theme(legend.position = "bottom")
    map_list[[index]] <- g
  }}

ggpubr::ggarrange(plotlist = map_list,ncol = 2, nrow = 3, common.legend = T, legend = "bottom")
```

# AI changes over time and scenarios

```{r validation}
v<-NULL
v2<-NULL
for (N in 5:30) {
  v<-c(v,sum(sapply(floor(2*N/3):N,function(i) choose(N,i)*(1/2)^N*2)))# proba d'avoir 2/3 de même signe à partir de N modèles, lorsque N change
  v2<-c(v2,sum(sapply(floor(2*N/3):N,function(i) choose(N,i)*(1/2)^N)))# proba d'avoir 2/3 de signe >0 à partir de N modèles, lorsque N change
}


plot(5:30,v,type="l",ylab="Proba(>2/3 models...)",xlab="Nb of models")
lines(5:30,v2,col="red")

```

## Changes AI for summary CMIP6

## Boxplot of changes AI

```{r boxplot-diff, fig.dim = c(8,5)}
ggpubr::ggarrange(plotlist = list(
  ggplot(subset(cmip6s, diff.AI.mean < 1 & diff.AI.mean > -1 & model == "historical"))+
    geom_boxplot(aes(x=period, y = - diff.AI.mean, col = period), outliers = F)+
    scale_color_manual(values = c("#D73027","#FDAE61","#FEE08B"))+
    labs(x = "", y = "Difference of AI between each period and\nthe reference period 1970-2000")+
    facet_grid(cols = vars(model))+
    theme_minimal() + theme(legend.position = "none", axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)),
  ggplot(subset(cmip6s, diff.AI.mean < 1 & diff.AI.mean > -1 & model != "historical"))+
    geom_boxplot(aes(x=period, y = diff.AI.mean, col = period), outliers = F)+
    scale_color_manual(values = c("#ABD9E9","#4575B4"))+
    scale_y_continuous(position = "right")+
    labs(y = "", x = "")+
    facet_grid(cols = vars(model))+
    theme_minimal() + theme(legend.position = "none", axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))),
  widths = c(1.5,3),
  ncol = 2, nrow = 1
)

   
```

## Strictly wetter or dryer (reference period: 1970-2000)

```{r map-plusminus, fig.dim = c(10,10)}
map_list <- list()

for(i in c("SSP245","SSP370","SSP585")){
  for (j in c("2030_2060","2070_2100")){
  index <- paste(i, j, sep = ", ")
  g <- ggplot()+
  geom_point(data = subset(cmip6s, model == i & period == j & plus.AI > 10), aes(x = lon, y = lat, col = "plus"), shape = "+")+
  geom_point(data = subset(cmip6s, model == i & period == j & minus.AI > 10), aes(x = lon, y = lat, col = "minus"), shape = "-")+
    scale_color_manual(values = c("#ABD9E9", "#D73027"))+
    labs(color = "", x = "", y = "", title = index)+
  theme_minimal()
  map_list[[index]] <- g
  }
}

ggarrange(plotlist = map_list, ncol = 2, nrow = 3, common.legend = T)

   
```

```{r map-historical-changes-strict, fig.dim = c(8,3)}
map_list <- list()

for(i in c("1850_1880","1985_2015")){
    index <- paste(i, j, sep = " , ")
  g <- ggplot() + geom_tile(data = subset(cmip6s, period == i & model == "historical" & diff.AI.mean > 0), aes(x=lon, y = lat,  fill = "wetter"))+
    geom_tile(data = subset(cmip6s, period == i & model == "historical" & diff.AI.mean < 0), aes(x=lon, y = lat,  fill = "dryer"))+
    geom_point(data = subset(cmip6s, period == i & model == "historical" & plus.AI > 10), aes(x=lon, y = lat), shape = "+", size = 0.5)+
     geom_point(data = subset(cmip6s, period == i & model == "historical" & minus.AI > 10), aes(x=lon, y = lat), shape = "-", size = 0.5)+
    borders(colour = "grey60")+
    scale_fill_manual(values =  c("#FDAE61", "#ABD9E9"), na.translate = F)+
    labs(title = paste("AI changes between",i , "and 1970-2000", sep = " "), fill = "")+
    theme_void()+ylim(-55,90)+
    theme(legend.position = "bottom")
  map_list[[i]] <- g
  }

ggarrange(plotlist = map_list, ncol = 2)

```

```{r map-future-changes-strict, fig.dim = c(10,10)}

map_list <- list()

for(i in c("SSP245","SSP370","SSP585")){
  for (j in c("2030_2060","2070_2100")){
    index <- paste(i, j, sep = " , ")
    g <- ggplot() + geom_tile(data = subset(cmip6s, period == j & model == i & diff.AI.mean > 0), aes(x=lon, y = lat,  fill = "wetter"))+
      geom_tile(data = subset(cmip6s, period == j & model == i & diff.AI.mean < 0), aes(x=lon, y = lat,  fill = "dryer"))+
       geom_point(data = subset(cmip6s, period == j & model == i & plus.AI > 10), aes(x=lon, y = lat), shape = "+", col = "grey20", size = 0.5)+
       geom_point(data = subset(cmip6s, period == j & model == i & minus.AI > 10), aes(x=lon, y = lat), shape = "-", col = "grey20", size = 0.5)+
      borders(colour = "grey60")+
      scale_fill_manual(values = c("#F46D43", "#74ADD1"), na.translate = F)+
      labs(fill = "Compared to 1970-2000", title = i)+
      theme_void()+ylim(-55,90)+
      theme(legend.position = "bottom")
    map_list[[index]] <- g
  }}



ggpubr::ggarrange(plotlist = map_list,ncol = 2, nrow = 3, common.legend = T, legend = "bottom")
```

## Wetter or dryer, binned (reference period: 1970-2000)

```{r summary-of-AI-changes}
ano.quantiles <- cmip6s %>% group_by(model, period) %>% summarise(q10 = quantile(diff.AI.mean, probs = 0.1, na.rm =T), 
                                                                  q25 = quantile(diff.AI.mean, probs = 0.25, na.rm =T), 
                                                                  q50 = quantile(diff.AI.mean, probs = 0.5, na.rm =T), 
                                                                  q75 = quantile(diff.AI.mean, probs = 0.75, na.rm =T), 
                                                                  q90 = quantile(diff.AI.mean, probs = 0.9, na.rm =T)) 

ano.quantiles <- cmip6s %>% summarise(q10 = quantile(diff.AI.mean, probs = 0.1, na.rm =T), 
                                      q25 = quantile(diff.AI.mean, probs = 0.25, na.rm =T), 
                                      q50 = quantile(diff.AI.mean, probs = 0.5, na.rm =T), 
                                      q75 = quantile(diff.AI.mean, probs = 0.75, na.rm =T), 
                                      q90 = quantile(diff.AI.mean, probs = 0.9, na.rm =T)) 
qbreaks <- c(-5, -1,-0.1, -0.01, 0, 0.01, 0.1, 1,5)
colscale <- c("#08519c", "#3182bd", "#6baed6", "#9ecae1", "#c6dbef", "#fddbc7", "#f4a582", "#d6604d", "#b2182b", "#67001f")
```

"diff.AI.mean" is the multimodel AI average minus the reference AI (period 1970-2000). More blue: AI mean is superior to AI ref, hence wetter than the ref. More red: the anomalie is negative, hence AI mean is inferior to AI ref, hence dryer. 

AS OF NOW THE SAME QUANTILES ARE USED FOR ALL MAPS.

```{r map-historical-changes-quantiles, fig.dim = c(8,3)}
map_list <- list()

for(i in c("1850_1880","1985_2015")){
  #quant = subset(ano.quantiles, model == "historical" & period == i)
  #b = quant[1,c(3:7)] %>% as.numeric
  b = qbreaks
  g <- ggplot() + geom_raster(data = subset(cmip6s, period == i & model == "historical"), aes(x=lon, y = lat,  fill = diff.AI.mean))+
    borders(colour = "grey60")+
           geom_point(data = subset(cmip6s, period == i & model == "historical" & plus.AI > 10), aes(x=lon, y = lat), shape = "+", col = "grey20", size = 0.5)+
       geom_point(data = subset(cmip6s,  period == i & model == "historical"  & minus.AI > 10), aes(x=lon, y = lat), shape = "-", col = "grey20", size = 0.5)+
    binned_scale(aesthetics = "fill", breaks = b, palette = function(x) rev(colscale),
                 guide = guide_legend(label.theme = element_text(angle = 0)))+
    labs(title = i, fill = "")+
    theme_void()+ylim(-55,90)+
    theme(legend.position = "right")
  map_list[[i]] <- g
}

ggpubr::ggarrange(plotlist = map_list, ncol = 2, common.legend = T, legend = "bottom")

```

```{r map-future-changes-quantiles, fig.dim = c(10,10)}

map_list <- list()

for(i in c("SSP245","SSP370","SSP585")){
  for (j in c("2030_2060","2070_2100")){
    index <- paste(i, j, sep = " , ")
    b = qbreaks
    g <- ggplot() + geom_raster(data = subset(cmip6s, period == j & model == i), aes(x=lon, y = lat,  fill = diff.AI.mean))+
      borders(colour = "grey60")+
                 geom_point(data = subset(cmip6s, period == i & model == "historical" & plus.AI > 10), aes(x=lon, y = lat), shape = "+", col = "grey20", size = 0.5)+
       geom_point(data = subset(cmip6s,  period == i & model == "historical"  & minus.AI > 10), aes(x=lon, y = lat), shape = "-", col = "grey20", size = 0.5)+
      binned_scale(aesthetics = "fill", breaks = b, palette = function(x) rev(colscale),
                   guide = guide_legend(label.theme = element_text(angle = 0)))+
      labs(fill = "Compared to 1970-2000", title = index)+
      theme_void()+ylim(-55,90)+
      theme(legend.position = "bottom")
    map_list[[index]] <- g
  }}



ggpubr::ggarrange(plotlist = map_list,ncol = 2, nrow = 3, common.legend = T, legend = "bottom")
```

## Wetter or dryer, binned, % (reference period: 1970-2000)

```{r summary-of-AI-changes-percent}
cmip6s$diff.AI.mean.percent <- with(cmip6s, (AI.mean-AI.ref.mean)/AI.ref.mean*100)


pbreaks <- c(-40, -30,-20, -10, 0, 10, 20, 30, 40)
colscale <- c("#08519c", "#3182bd", "#6baed6", "#9ecae1", "#c6dbef","white","white", "#fddbc7", "#f4a582", "#d6604d", "#b2182b")
```

"diff.AI.mean" is the multimodel AI average minus the reference AI (period 1970-2000). More blue: AI mean is superior to AI ref, hence wetter than the ref. More red: the anomalie is negative, hence AI mean is inferior to AI ref, hence dryer. 

```{r map-historical-changes-percent, fig.dim = c(8,3)}
map_list <- list()

for(i in c("1850_1880","1985_2015")){
  #quant = subset(ano.quantiles, model == "historical" & period == i)
  #b = quant[1,c(3:7)] %>% as.numeric
  b = pbreaks
  g <- ggplot() + geom_raster(data = subset(cmip6s, period == i & model == "historical"), aes(x=lon, y = lat,  fill = diff.AI.mean.percent))+
    borders(colour = "grey60")+
           geom_point(data = subset(cmip6s, period == i & model == "historical" & plus.AI > 10), aes(x=lon, y = lat), shape = "+", col = "grey20", size = 0.5)+
       geom_point(data = subset(cmip6s,  period == i & model == "historical"  & minus.AI > 10), aes(x=lon, y = lat), shape = "-", col = "grey20", size = 0.5)+
    binned_scale(aesthetics = "fill", breaks = b, palette = function(x) rev(colscale))+
    labs(title = i, fill = "")+
    theme_void()+ylim(-55,90)+
    theme(legend.position = "right")
  map_list[[i]] <- g
}

ggpubr::ggarrange(plotlist = map_list, ncol = 2, common.legend = T, legend = "bottom") %>% annotate_figure(top = "% change of AI index compared to 1970-2000")

#,guide = guide_legend(label.theme = element_text(angle = 0))
```

```{r map-future-changes-percent, fig.dim = c(15,15)}

map_list <- list()

for(i in c("SSP245","SSP370","SSP585")){
  for (j in c("2030_2060","2070_2100")){
    index <- paste(i, j, sep = " , ")
    b = pbreaks
    g <- ggplot() + geom_tile(data = subset(cmip6s, period == j & model == i), aes(x=lon, y = lat,  fill = diff.AI.mean.percent))+
      borders(colour = "grey60", size = 0.5)+
                 geom_point(data = subset(cmip6s, period == j & model == i & plus.AI > 10), aes(x=lon, y = lat), shape = "+", col = "grey20", size = 0.5)+
       geom_point(data = subset(cmip6s,  period == j & model == i & minus.AI > 10), aes(x=lon, y = lat), shape = "-", col = "grey20", size = 0.5)+
      binned_scale(aesthetics = "fill", breaks = b, palette = function(x) rev(colscale))+
      labs(fill = "% change of AI index\ncompared to 1970-2000", title = index)+
      theme_void()+ylim(-55,90)+
      theme(legend.position = "bottom")
    map_list[[index]] <- g
  }}


ggpubr::ggarrange(plotlist = map_list,ncol = 2, nrow = 3, common.legend = T, legend = "bottom")

ggsave("map_AIpercent_SSP585_2100.png", map_list[[index]])
```


# Maps of difference between future and reference category of AI, using cat maj

## Detailed changes and colors cat maj

```{r detailed-cat-change-catmaj}
cmip6s$cat.maj[which(is.na(cmip6s$cat.maj) == T)] <- "NA"
cmip6s$cat.maj.ref[which(is.na(cmip6s$cat.maj.ref) == T)] <- "NA"
cmip6s$change.catmaj <- with(cmip6s, ifelse(cat.maj == cat.maj.ref, 0, 1))

cmip6s$change.cat.maj <- with(cmip6s, paste(cat.maj.ref, cat.maj, sep = " to "))

cmip6s$change.cat.maj <- factor(cmip6s$change.cat.maj, 
                                levels = c("Cold to Humid", 
                                           "Cold to Dry subhumid",
                                           "Cold to Semi-arid",
                                           "Humid to Arid",
                                           "Humid to Semi-arid",
                                           "Humid to Dry subhumid",
                                           "Dry subhumid to Semi-arid",
                                           "Dry subhumid to Arid",
                                           "Semi-arid to Arid",
                                           "Arid to Hyperarid", 
                                           "Hyperarid to Arid",
                                           "Arid to Semi-arid",
                                           "Arid to Dry subhumid",
                                           "Arid to Humid",
                                           "Semi-arid to Dry subhumid",
                                           "Semi-arid to Humid",
                                           "Semi-arid to Cold",
                                           "Dry subhumid to Humid",
                                           "Dry subhumid to Cold",
                                           "Humid to Cold"))

colors_changes <- c("Cold to Humid" = "#FEE08B", 
                                           "Cold to Dry subhumid" = "#FDCB6A",
                                           "Cold to Semi-arid" = "#FDB463",
                                           "Humid to Arid" = "#FDAE61",
                                           "Humid to Semi-arid" = "#F68D54",
                                           "Humid to Dry subhumid" =  "#F46B43",
                                           "Dry subhumid to Semi-arid" = "#F03B20",
                                           "Dry subhumid to Arid" = "#E63125",
                                           "Semi-arid to Arid" = "#D73027",
                                           "Arid to Hyperarid" = "#BD0026", 
                                           "Hyperarid to Arid" = "#ABD9E9",
                                           "Arid to Semi-arid" = "#91C4DC",
                                           "Arid to Dry subhumid" = "#7FB2D0",
                                           "Arid to Humid" = "#6A9EC4",
                                           "Semi-arid to Dry subhumid" = "#5893B8",
                                           "Semi-arid to Humid" = "#4575B4",
                                           "Semi-arid to Cold" = "#5D83C9",
                                           "Dry subhumid to Humid" = "#77A1E2",
                                           "Dry subhumid to Cold" = "#A7C7E9",
                                           "Humid to Cold" = "#DAE2F7")
```

## Changes towards dryer categories  cat maj

### Historical

```{r map-historical-changes-dry-catmaj, fig.dim = c(8,3)}
map_list <- list()

for(i in c("1850_1880","1985_2015")){
  g <- ggplot() + 
    geom_raster(data = subset(cmip6s, period == i & model == "historical" & change.catmaj == 1 & cat.maj.dryer == "dryer"), 
                aes(x=lon, y = lat,  fill = change.cat.maj))+
    borders(colour = "grey60")+
    scale_fill_manual(values =  colors_changes, na.translate = F)+
    labs(title = i, fill = "Change of category\ncompared to\n1970-2000")+
    theme_void()+ylim(-55,90)+
    theme(legend.position = "bottom")+guides(fill = guide_legend(nrow = 3))
  map_list[[i]] <- g
}


ggarrange(plotlist = map_list, ncol = 2, common.legend = T, legend = "bottom")
```

#### Future

```{r map-future-changes-dry-catmaj, fig.dim = c(10,10)}

map_list <- list()

for(i in c("SSP245","SSP370","SSP585")){
  for (j in c("2030_2060","2070_2100")){
    index <- paste(i, j, sep = " , ")
    g <- ggplot() + geom_tile(data = subset(cmip6s, period == j & model == i & change.catmaj == 1 & cat.maj.dryer == "dryer"), 
                              aes(x=lon, y = lat,  fill = change.cat.maj))+
      borders(colour = "grey60")+
      scale_fill_manual(values = colors_changes, na.translate = F)+
      labs(fill = "Change of category\ncompared to\n1970-2000", title = index)+
      theme_void()+ylim(-55,90)+
      theme(legend.position = "bottom")+guides(fill = guide_legend(nrow = 3))
    map_list[[index]] <- g
  }}

ggarrange(plotlist = map_list,ncol = 2, nrow = 3, common.legend = T, legend = "bottom")
```

## Changes towards wetter categories  cat maj

#### Historical

```{r map-historical-changes-wet-catmaj, fig.dim = c(8,2)}
map_list <- list()

for(i in c("1850_1880","1985_2015")){
  g <- ggplot() + geom_raster(data = subset(cmip6s, period == i & model == "historical" & change.catmaj == 1 & cat.maj.dryer == "wetter"), 
                              aes(x=lon, y = lat,  fill = change.cat.maj))+
    borders(colour = "grey60", ylim = c(-40,40))+
    scale_fill_manual(values =  colors_changes, na.translate = F)+
    labs(title = i, fill = "Change of category\ncompared to\n1970-2000")+
    theme_void()+
    ylim(-50,50)+
    theme(legend.position = "bottom")+guides(fill = guide_legend(nrow = 3))
  map_list[[i]] <- g
}

ggarrange(plotlist = map_list, ncol = 2, common.legend = T, legend = "bottom")

```

#### Future

```{r map-future-changes-wet-catmaj, fig.dim = c(16,8)}

map_list <- list()

for(i in c("SSP245","SSP370","SSP585")){
  for (j in c("2030_2060","2070_2100")){
    index <- paste(i, j, sep = " , ")
    g <- ggplot() + geom_tile(data = subset(cmip6s, period == j & model == i & change.catmaj == 1 & cat.maj.dryer == "wetter"), 
                              aes(x=lon, y = lat,  fill = change.cat.maj))+
      borders(colour = "grey60", ylim = c(-40,40))+
      scale_fill_manual(values = colors_changes, na.translate = F)+
      labs(fill = "Change of category\ncompared to\n1970-2000", title = index)+
      theme_void()+ylim(-50,50)+
      theme(legend.position = "bottom")+guides(fill = guide_legend(nrow = 3))
    map_list[[index]] <- g
  }}

ggpubr::ggarrange(plotlist = map_list, nrow = 3, ncol = 2, common.legend = T, legend = "bottom")
```

# Map of difference between future and reference category of AI, using cat mean

## Detailed changes and colors 

```{r detailed-cat-mean-change}
cmip6s$change.catmean <- with(cmip6s, ifelse(cat.AI == cat.AI.ref, 0, 1))

cmip6s$change.cat.mean <- with(cmip6s, paste(cat.AI.ref, cat.AI, sep = " to "))
unique(cmip6s$change.cat.mean)

table(cmip6s$change.cat.mean)
cmip6s$change.cat.mean <- factor(cmip6s$change.cat.mean, 
                                  levels = c("Hyperarid to Arid",
                                             "Arid to Semi-arid",
                                             "Semi-arid to Dry subhumid",
                                             "Dry subhumid to Humid",
                                             "Humid to Cold",
                                             "Cold to Humid",
                                             "Humid to Dry subhumid",
                                             "Humid to Semi-arid",
                                             "Dry subhumid to Semi-arid",
                                             "Semi-arid to Arid",
                                             "Arid to Hyperarid"))

wtd <- colorRampPalette(c("#FEE08B","firebrick"))
col.wtd <- wtd(6)

dtw <- colorRampPalette(c( "#DAE2F7","#4575B4" ))
col.dtw <- dtw(5)

colors_changes <- c("Hyperarid to Arid" = col.dtw[1],
                    "Arid to Semi-arid" = col.dtw[2],
                    "Semi-arid to Dry subhumid" = col.dtw[3],
                    "Dry subhumid to Humid" = col.dtw[4],
                    "Humid to Cold" = col.dtw[5],
                    "Cold to Humid" = col.wtd[1],
                    "Humid to Dry subhumid" = col.wtd[2],
                    "Humid to Semi-arid" = col.wtd[3],
                    "Dry subhumid to Semi-arid" = col.wtd[4],
                    "Semi-arid to Arid" = col.wtd[5],
                    "Arid to Hyperarid" = col.wtd[6])
   
```

## Changes towards dryer categories cat mean
### Historical

```{r map-historical-changes-dry-catmean, fig.dim = c(8,3)}
map_list <- list()

for(i in c("1850_1880","1985_2015")){
  g <- ggplot() + 
    geom_raster(data = subset(cmip6s, period == i & model == "historical" & change.catmean == 1 & cat.dryer == "dryer"), 
                aes(x=lon, y = lat,  fill = change.cat.mean))+
    borders(colour = "grey60")+
    scale_fill_manual(values =  colors_changes, na.translate = F)+
    labs(title = i, fill = "Change of category\ncompared to\n1970-2000")+
    theme_void()+ylim(-55,90)+
    theme(legend.position = "bottom")+guides(fill = guide_legend(nrow = 2))
  map_list[[i]] <- g
}

ggarrange(plotlist = map_list, ncol = 2, common.legend = T, legend = "bottom")
```

### Future

```{r map-future-changes-dry-catmean, fig.dim = c(10,10)}

map_list <- list()

for(i in c("SSP245","SSP370","SSP585")){
  for (j in c("2030_2060","2070_2100")){
    index <- paste(i, j, sep = " , ")
    g <- ggplot() + geom_tile(data = subset(cmip6s, period == j & model == i & change.catmean == 1 & cat.dryer == "dryer"), 
                              aes(x=lon, y = lat,  fill = change.cat.mean))+
      borders(colour = "grey60")+
      scale_fill_manual(values = colors_changes)+
      labs(fill = "Change of category\ncompared to\n1970-2000", title = index)+
      theme_void()+ylim(-55,90)+
      theme(legend.position = "bottom")+guides(fill = guide_legend(nrow = 2))
    map_list[[index]] <- g
  }}

ggarrange(plotlist = map_list, ncol = 2, nrow = 3, common.legend = T, legend = "bottom")
```


## Changes towards wetter categories cat mean

### Historical

```{r map-historical-changes-wet-, fig.dim = c(8, 2)}
map_list <- list()

for(i in c("1850_1880","1985_2015")){
  g <- ggplot() + geom_tile(data = subset(cmip6s, period == i & model == "historical" & change.catmaj == 1 & cat.dryer == "wetter"), 
                            aes(x=lon, y = lat,  fill = change.cat.mean))+
    borders(colour = "grey60", ylim = c(-40,40))+
    scale_fill_manual(values =  colors_changes, na.translate = F)+
    labs(title = i, fill = "Change of category\ncompared to\n1970-2000")+
    theme_void()+ylim(-50,50)+
    theme(legend.position = "bottom")+guides(fill = guide_legend(nrow = 2))
  map_list[[i]] <- g
}

ggarrange(plotlist = map_list, ncol = 2, common.legend = T, legend = "bottom")

```

### Future

```{r map-future-changes-wet-catmean, fig.dim = c(10,8)}

map_list <- list()

for(i in c("SSP245","SSP370","SSP585")){
  for (j in c("2030_2060","2070_2100")){
    index <- paste(i, j, sep = " , ")
    g <- ggplot() + geom_tile(data = subset(cmip6s, period == j & model == i & change.catmean == 1 & cat.dryer == "wetter"), 
                              aes(x=lon, y = lat,  fill = change.cat.mean))+
      borders(colour = "grey60", ylim = c(-40,40))+
      scale_fill_manual(values = colors_changes, na.translate = F)+
      labs(fill = "Change of category\ncompared to\n1970-2000", title = index)+
      theme_void()+
      theme(legend.position = "bottom")+guides(fill = guide_legend(nrow = 2))
    map_list[[index]] <- g
  }}

ggpubr::ggarrange(plotlist = map_list,ncol = 2, nrow = 3, common.legend = T, legend = "bottom")
```




## Big map of SSP585 2070-2100

```{r bigmap, fig.dim = c(10,7)}

g <- ggplot() + 
  geom_tile(data = subset(cmip6s, period == "2070_2100" & model == "SSP585" & change.catmean == 1), aes(x=lon, y = lat,  fill = change.cat.mean))+
      borders(colour = "grey60")+
      scale_fill_manual(values = colors_changes, na.translate = F)+
      labs(fill = "Change of aridity category\nfor the period 2070-2100\nand SSP 5-8.5,\ncompared to 1970-2000", title = "")+
      theme_void()+ylim(-55,90)+
      theme(legend.position = "bottom")+guides(fill = guide_legend(nrow = 4))
   
print(g)

```

# Tables

## Changes of aridity category using cat mean

```{r changes-in-table}
tab <- cmip6s %>% group_by(model, period, Continent) %>% summarise(Cold = table(cat.AI)[1], Hyperarid = table(cat.AI)[2], Arid = table(cat.AI)[3], Semiarid = table(cat.AI)[4], Drysubhumid = table(cat.AI)[5], Humid = table(cat.AI)[6], total = n()) 
write.table(tab, "tab.catAI.txt")

tab.future <- cmip6s %>% subset(!Continent %in% c("SOUTHERN","PACIFIC","ATLANTIC","INDIAN","ARCTIC")) %>%
  group_by(period, model, cat.AI) %>%
  summarise(count = n()) %>%
  ungroup() %>% group_by(period, model) %>% mutate(percent = round(count/sum(count)*100, 1), count = NULL) %>%
  reshape2::dcast(period + model ~cat.AI) %>%
  mutate("Sum drylands" = rowSums(.[c("Hyperarid","Arid","Semi-arid","Dry subhumid")])) %>%
  select(c("model","period","Hyperarid","Arid","Semi-arid","Dry subhumid","Sum drylands","Humid","Cold","NA"))

kable(tab.future[order(tab.future$model, decreasing = F),]) %>% kable_styling(bootstrap_options = "bordered") %>%
  column_spec(c(7,10), italic = T, include_thead = T) %>% row_spec(2, bold = T) 

# ggplot(tab.future, aes(x = period))+
#   geom_point(aes(y = Hyperarid, col = "Hyperarid", shape = model))+
#   geom_line(aes(y=Hyperarid, col = "Hyperarid", group = model, lty = model))+
#   geom_point(aes(y = Arid, col = "Arid", shape = model))+
#   geom_line(aes(y=Arid, col = "Arid", group = model, lty = model))+
#   geom_point(aes(y = get('Semi-arid'), col = "Semi-arid", shape = model))+
#   geom_line(aes(y=get('Semi-arid'), col = "Semi-arid", group = model, lty = model))+
#   geom_point(aes(y = get('Dry subhumid'), col = "Dry subhumid", shape = model))+
#   geom_line(aes(y=get('Dry subhumid'), col = "Dry subhumid", group = model, lty = model))+
#   geom_point(aes(y = Humid, col = "Humid", shape = model))+
#   geom_line(aes(y= Humid, col = "Humid", group = model, lty = model))+
#   scale_color_manual(values = col.cat)+
#   theme_minimal()
```

```{r tab-percent-catmean, eval = F}

cmip6$cat.AI <- factor(cmip6$cat.AI, levels = c("Hyperarid", "Arid", "Semi-arid", "Dry subhumid", "Humid", "Cold"))
tab.percent <- cmip6 %>% subset(!Continent %in% c("SOUTHERN","PACIFIC","ATLANTIC","INDIAN","ARCTIC") & lat > -55 ) %>%
  group_by(period, model, cat.AI, source) %>%
  summarise(count = n()) %>%
  ungroup() %>% 
  group_by(period, model, cat.AI) %>% 
  summarise(mmmean = mean(count, na.rm = T), mmsd = sd(count, na.rm = T)) %>%
  mutate(percent = round(mmmean/sum(mmmean)*100, 1), sd.percent = round(mmsd/sum(mmmean)*100,1)) 
write.table(tab.percent, "tab.percent.txt")
```

```{r colplots-futurecatmean, fig.dim = c(10,5)}
tab.percent <- read.table("tab.percent.txt")

df245 <- rbind(subset(tab.percent, model == "historical" & period %in% c("1970_2000")), subset(tab.percent, model == "SSP245")) %>%
  group_by(period, model) %>% 
  mutate(lab.y = (rev(cumsum(rev(percent)))) - percent*0.5) %>% 
  subset(cat.AI %in% c("Hyperarid", "Arid", "Semi-arid", "Dry subhumid", "Humid","Cold"))

df245$cat.AI <- factor(df245$cat.AI, levels = c("Hyperarid", "Arid", "Semi-arid", "Dry subhumid", "Humid", "Cold"))

b245 <- ggplot(data = df245, aes(x = period, y = percent))+
  geom_col(aes(group = cat.AI, col = cat.AI, fill = cat.AI))+
  geom_label(aes(y = lab.y, label = paste(percent, " ± ", sd.percent, sep = "")), size = 3)+
  scale_color_manual(values = col.cat, aesthetics = c("col", "fill"), na.translate = F)+
  labs(title = "SSP 2-4.5", y = "%", x = "", col = "", fill = "")+
  theme_minimal()

df370 <- rbind(subset(tab.percent, model == "historical" & period %in% c( "1970_2000")), subset(tab.percent, model == "SSP370")) %>%
  group_by(period, model) %>% 
  mutate(lab.y = (rev(cumsum(rev(percent)))) - percent*0.5) %>% 
  subset(cat.AI %in% c("Hyperarid", "Arid", "Semi-arid", "Dry subhumid", "Humid","Cold"))

df370$cat.AI <- factor(df370$cat.AI, levels = c("Hyperarid", "Arid", "Semi-arid", "Dry subhumid", "Humid", "Cold"))

b370 <- ggplot(data = df370, aes(x = period, y = percent))+
  geom_col(aes(group = cat.AI, col = cat.AI, fill = cat.AI))+
  geom_label(aes(y = lab.y, label = paste(percent, " ± ", sd.percent, sep = "")), size = 3)+
  scale_color_manual(values = col.cat, aesthetics = c("col", "fill"), na.translate = F)+
  labs(title = "SSP 3-7.0", y = "%", x = "", col = "", fill = "")+
  theme_minimal()


df585 <- rbind(subset(tab.percent, model == "historical" & period %in% c("1970_2000")), subset(tab.percent, model == "SSP585")) %>%
  group_by(period, model) %>% 
  mutate(lab.y = (rev(cumsum(rev(percent)))) - percent*0.5) %>% 
  subset(cat.AI %in% c("Hyperarid", "Arid", "Semi-arid", "Dry subhumid", "Humid","Cold"))

df585$cat.AI <- factor(df585$cat.AI, levels = c("Hyperarid", "Arid", "Semi-arid", "Dry subhumid", "Humid", "Cold"))

b585 <- ggplot(data = df585, aes(x = period, y = percent))+
  geom_col(aes(group = cat.AI, col = cat.AI, fill = cat.AI))+
  geom_label(aes(y = lab.y, label = paste(percent, " ± ", sd.percent, sep = "")), size = 3)+
  scale_color_manual(values = col.cat, aesthetics = c("col", "fill"), na.translate = F)+
  labs(title = "SSP 5-8.5", y = "%", x = "", col = "", fill = "")+
  theme_minimal()


ggarrange(plotlist = list(b245, b370, b585), common.legend = T, legend = "bottom", ncol = 3)
```



## Changes of aridity category using cat maj

```{r changes-in-table-maj}
tab.maj <- cmip6s %>% group_by(model, period, Continent) %>% summarise(Cold = table(cat.maj)[1], Hyperarid = table(cat.maj)[2], Arid = table(cat.maj)[3], Semiarid = table(cat.maj)[4], Drysubhumid = table(cat.maj)[5], Humid = table(cat.maj)[6], total = n()) 
write.table(tab.maj, "tab.catmaj.txt")

tab.future.maj <- cmip6s %>% subset(!Continent %in% c("SOUTHERN","PACIFIC","ATLANTIC","INDIAN","ARCTIC")) %>%
  group_by(period, model, cat.maj) %>%
  summarise(count = n()) %>%
  ungroup() %>% group_by(period, model) %>% mutate(percent = round(count/sum(count)*100, 1), count = NULL) %>%
  reshape2::dcast(period + model ~cat.maj) %>%
  mutate("Sum drylands" = rowSums(.[c("Hyperarid","Arid","Semi-arid","Dry subhumid")])) %>%
  select(c("model","period","Hyperarid","Arid","Semi-arid","Dry subhumid","Sum drylands","Humid","Cold","NA"))

kable(tab.future.maj[order(tab.future.maj$model, decreasing = F),]) %>% kable_styling(bootstrap_options = "bordered") %>%
  column_spec(c(7,10), italic = T, include_thead = T) %>% row_spec(2, bold = T) 
```

```{r colplots-futurecat, fig.dim = c(10,5)}
tab.flow.catmaj <- cmip6s %>% subset(!Continent %in% c("SOUTHERN","PACIFIC","ATLANTIC","INDIAN","ARCTIC") & lat > -55 ) %>%
  group_by(period, model, cat.maj) %>%
  summarise(count = n()) %>%
  ungroup() %>% group_by(period, model) %>% mutate(percent = round(count/sum(count)*100, 1), count = NULL)

df245 <- rbind(subset(tab.flow.catmaj, model == "historical" & period %in% c("1970_2000")), subset(tab.flow.catmaj, model == "SSP245")) %>%
  mutate(lab.y = (rev(cumsum(rev(percent)))) - percent*0.5) %>% 
  subset(cat.maj %in% c("Hyperarid", "Arid", "Semi-arid", "Dry subhumid", "Humid","Cold"))

b245 <- ggplot(data = df245, aes(x = period, y = percent))+
  geom_col(aes(group = cat.maj, col = cat.maj, fill = cat.maj))+
  geom_label(aes(y = lab.y, label = percent))+
  scale_color_manual(values = col.cat, aesthetics = c("col", "fill"), na.translate = F)+
  labs(title = "SSP 2-4.5", y = "%", x = "")+
  theme_minimal()

df370 <- rbind(subset(tab.flow.catmaj, model == "historical" & period %in% c("1970_2000")), subset(tab.flow.catmaj, model == "SSP370")) %>% mutate(lab.y = (rev(cumsum(rev(percent)))) - percent*0.5) %>% 
  subset(cat.maj %in% c("Hyperarid", "Arid", "Semi-arid", "Dry subhumid", "Humid","Cold"))

b370 <- ggplot(data = df370, aes(x = period, y = percent))+
  geom_col(aes(group = cat.maj, col = cat.maj, fill = cat.maj))+
  geom_label(aes(y = lab.y, label = percent))+
  scale_color_manual(values = col.cat, aesthetics = c("col", "fill"), na.translate = F)+
  labs(title = "SSP 3-7.0", y = "%", x = "")+
  theme_minimal()


df585 <- rbind(subset(tab.flow.catmaj, model == "historical" & period %in% c("1970_2000")), subset(tab.flow.catmaj, model == "SSP585")) %>% mutate(lab.y = (rev(cumsum(rev(percent)))) - percent*0.5) %>% 
  subset(cat.maj %in% c("Hyperarid", "Arid", "Semi-arid", "Dry subhumid", "Humid","Cold"))

b585 <- ggplot(data = df585, aes(x = period, y = percent))+
  geom_col(aes(group = cat.maj, col = cat.maj, fill = cat.maj))+
  geom_label(aes(y = lab.y, label = percent))+
  scale_color_manual(values = col.cat, aesthetics = c("col", "fill"), na.translate = F)+
  labs(title = "SSP 5-8.5", y = "%", x = "")+
  theme_minimal()


ggarrange(plotlist = list(b245, b370, b585), common.legend = T, legend = "bottom", ncol = 3)
```



## Changes by continent, cat mean

```{r tab-percent-continent-catmean, eval = F}

cmip6$cat.AI <- factor(cmip6$cat.AI, levels = c("Hyperarid", "Arid", "Semi-arid", "Dry subhumid", "Humid", "Cold"))

tab.percent.cont <- cmip6 %>% subset(!Continent %in% c("SOUTHERN","PACIFIC","ATLANTIC","INDIAN","ARCTIC") & lat > -55 ) %>%
  group_by(period, model, Continent, cat.AI, source) %>%
  summarise(count = n()) %>%
  ungroup() %>% 
  group_by(period, model, Continent, cat.AI) %>% 
  summarise(mmmean = mean(count, na.rm = T), mmsd = sd(count, na.rm = T)) %>%
  mutate(percent = round(mmmean/sum(mmmean)*100, 1), sd.percent = round(mmsd/sum(mmmean)*100,1)) 
write.table(tab.percent.cont, "tab.percent.continent.txt")
```

```{r changes-in-table-by-continent-mean, results = "asis"}
tab.percent.cont <- read.table("tab.percent.continent.txt")

tab_list_percent <- list()
tab_list_sd <- list()

for(i in unique(tab.percent.cont$Continent)){
  tab.i <- subset(tab.percent.cont, Continent == i)
  
  df.percent <- tab.i %>% reshape2::dcast(period + model ~ cat.AI, value.var = "percent")
  
  df.sd <- tab.i %>% reshape2::dcast(period + model ~ cat.AI, value.var = "sd.percent")
  # sd for a sum: square root of sum of squared sd
  
  drycats <- which(names(df.percent) %in% c("Hyperarid","Arid","Semi-arid","Dry subhumid"))
  
  df.percent <- df.percent %>% mutate("Sum drylands" = rowSums(.[drycats], na.rm = T))
  df.sd <- df.sd %>% mutate("Sum drylands" = round(sqrt(rowSums((.[drycats])^2, na.rm = T)),2))
  
  missing <- setdiff(c("model","period","Hyperarid","Arid","Semi-arid","Dry subhumid","Sum drylands","Humid","Cold","NA"), names(df.percent))
  
  df.percent[, missing] <- 0
  df.sd[,missing] <- 0
  
  df.percent <- df.percent %>% select(c("model","period","Hyperarid","Arid","Semi-arid","Dry subhumid","Sum drylands","Humid","Cold","NA"))
  df.percent[is.na(df.percent)] <- 0
  
  tab_list_percent[[i]] <- df.percent
  
  df.sd <- df.sd %>% select(c("model","period","Hyperarid","Arid","Semi-arid","Dry subhumid","Sum drylands","Humid","Cold","NA"))
  df.sd[is.na(df.sd)] <- 0
  
  tab_list_sd[[i]] <- df.sd
}

k_list <- list()


kdiff_list <- list()

for(i in names(tab_list_percent)){
  
  df <- tab_list_percent[[i]] %>% select(c("model", "period"))
  
  
  
  prop.ref <- tab_list_percent[[i]] %>% filter(period == "1970_2000")
  prop.ref <- prop.ref %>% rbind(prop.ref[rep(1,dim(df)[1]-1),])
  df2 <- tab_list_percent[[i]][,c(3:10)] - prop.ref[,c(3:10)] %>% round(1)
  diff.prop.percent <- cbind(df, df2)
  
  sd.ref <- tab_list_sd[[i]] %>% filter(period == "1970_2000")
  sd.ref <- sd.ref %>% rbind(sd.ref[rep(1,dim(df)[1]-1),])
  df3 <- sqrt(tab_list_sd[[i]][,c(3:10)]^2 + sd.ref[,c(3:10)]^2) %>% round(1)
  diff.sd.percent <- cbind(df, df3)
  
  df$Hyperarid <- paste(tab_list_percent[[i]]$Hyperarid, tab_list_sd[[i]]$Hyperarid, sep = " ± ")
  df$Arid <- paste(tab_list_percent[[i]]$Arid, tab_list_sd[[i]]$Arid, sep = " ± ")
  df$'Semi-Arid' <- paste(tab_list_percent[[i]][,5], tab_list_sd[[i]][,5], sep = " ± ")
  df$'Dry subhumid' <- paste(tab_list_percent[[i]][,6], tab_list_sd[[i]][,6], sep = " ± ")
  df$'Sum drylands' <- paste(tab_list_percent[[i]][,7], tab_list_sd[[i]][,7], sep = " ± ")
  df$Humid <- paste(tab_list_percent[[i]]$Humid, tab_list_sd[[i]]$Humid, sep = " ± ")
  df$Cold <- paste(tab_list_percent[[i]]$Cold, tab_list_sd[[i]]$Cold, sep = " ± ")
  
  df.diff <- df
  df.diff$Hyperarid <- paste(round(diff.prop.percent$Hyperarid, 1), round(diff.sd.percent$Hyperarid, 1), sep = " ± ")
  df.diff$Arid <- paste(round(diff.prop.percent$Arid, 1), round(diff.sd.percent$Arid, 1), sep = " ± ")
  df.diff$'Semi-Arid' <- paste(round(diff.prop.percent[,5], 1), round(diff.sd.percent[,5], 1), sep = " ± ")
  df.diff$'Dry subhumid' <- paste(round(diff.prop.percent[,6], 1), round(diff.sd.percent[,6], 1), sep = " ± ")
  df.diff$'Sum drylands' <- paste(round(diff.prop.percent[,7],1), round(diff.sd.percent[,7], 1), sep = " ± ")
  df.diff$Humid <- paste(round(diff.prop.percent$Humid, 1), round(diff.sd.percent$Humid, 1), sep = " ± ")
  df.diff$Cold <- paste(round(diff.prop.percent$Cold, 1), round(diff.sd.percent$Cold, 1), sep = " ± ")
  
  k_list[[i]] <- kable(df[order(df$model, decreasing = F),], caption = i) %>% kable_styling(bootstrap_options = "bordered") %>%
    column_spec(8, italic = T, background = "#DAE2F7") %>% row_spec(2, bold = T) 
  kdiff_list[[i]] <- kable(df.diff[order(df.diff$model, decreasing = F),], caption = i) %>% kable_styling(bootstrap_options = "bordered") %>% 
    column_spec(8, italic = T, background = "#DAE2F7") %>% row_spec(2, bold = T) 
  
}

```

### Proportion of aridity categories by continent

```{r prop-aridity-category, results = "asis"}
k_list
```

### Changes in proportion of aridity categories by continent

```{r prop-change-aridity-category, results = "asis"}
kdiff_list
```

### Changes in proportion of aridity categories, by continent, calculated by model

```{r tab-percent-continent-bymodel-catmean, eval = F}

cmip6$cat.AI <- factor(cmip6$cat.AI, levels = c("Hyperarid", "Arid", "Semi-arid", "Dry subhumid", "Humid", "Cold"))

tab.percent.cont <- cmip6 %>% subset(!Continent %in% c("SOUTHERN","PACIFIC","ATLANTIC","INDIAN","ARCTIC") & lat > -55 ) %>%
  group_by(period, model, Continent, cat.AI, source) %>%
  summarise(count = n()) %>%
  ungroup() %>% 
  group_by(period, model, Continent, cat.AI) %>% 
  summarise(mmmean = mean(count, na.rm = T), mmsd = sd(count, na.rm = T)) %>%
  mutate(percent = round(mmmean/sum(mmmean)*100, 1), sd.percent = round(mmsd/sum(mmmean)*100,1)) 
write.table(tab.percent.cont, "tab.percent.continent.txt")

tab.cont <- cmip6 %>% subset(!Continent %in% c("SOUTHERN","PACIFIC","ATLANTIC","INDIAN","ARCTIC") & lat > -55 ) %>%
  group_by(period, model, Continent, cat.AI, source) %>%
  summarise(count = n()) %>%
  ungroup() 

list_ref <- list()

for(i in unique(tab.cont$source)){
  tab.cont.i <- filter(tab.cont, source == i)
  tab.cont.i.ref <- filter(tab.cont.i, period == "1970_2000")
  
  for(j in unique(tab.cont.i$Continent)){
    tab.cont.i.j <- filter(tab.cont.i, Continent == j)
    tab.cont.i.j.ref <- filter(tab.cont.i.ref, Continent == j)
    
    for(k in unique(tab.cont.i$cat.AI)){
      
      index <- paste(i, j, k, sep = "_")
      tab.cont.i.j.k <- filter(tab.cont.i.j, cat.AI == k)
      
      lines <- ifelse(dim(tab.cont.i.j.k)[1] > 0, dim(tab.cont.i.j.k)[1] - 1, 1)
      
      tab.cont.i.j.k.ref <- subset(tab.cont.i.j.ref, cat.AI == k)
      
      if(dim(tab.cont.i.j.k.ref)[1] == 0){
        names1 <- names(tab.cont.i.j.k.ref)
        line1 <- c("1970_2000","historical", i, j, k, 0)
        tab.cont.i.j.k.ref <- rbind(tab.cont.i.j.k.ref, line1) %>%
          setNames(names1)
        
      }else{
        tab.cont.i.j.k.ref <- tab.cont.i.j.k.ref
      }
      
      tab.cont.i.j.k.ref <- tab.cont.i.j.k.ref %>% rbind(tab.cont.i.j.k.ref[rep(1,lines),])
      tab.cont.i.j.k$diff <- tab.cont.i.j.k$count - as.numeric(tab.cont.i.j.k.ref$count)  
      
      list_ref[[index]] <- tab.cont.i.j.k
    }
  }
}


tab.cont.diff <- bind_rows(list_ref, .id = "column_label") %>%
  group_by(period, model, Continent, cat.AI) %>% 
  summarise(diff.mean = mean(diff, na.rm = T), diff.sd = sd(diff, na.rm = T))
write.table(tab.cont.diff, "tab.cont.diff.txt")

```

```{r changes-in-table-by-continent-diff, results = "asis"}
tab.cont.diff <- read.table("tab.cont.diff.txt")

tab_list_percent <- list()
tab_list_sd <- list()

for(i in unique(tab.cont.diff$Continent)){
  tab.i <- subset(tab.cont.diff, Continent == i)
  
  df.percent <- tab.i %>% reshape2::dcast(period + model ~ cat.AI, value.var = "diff.mean")
  
  df.sd <- tab.i %>% reshape2::dcast(period + model ~ cat.AI, value.var = "diff.sd")
  # sd for a sum: square root of sum of squared sd
  
  drycats <- which(names(df.percent) %in% c("Hyperarid","Arid","Semi-arid","Dry subhumid"))
  
  df.percent <- df.percent %>% mutate("Sum drylands" = rowSums(.[drycats], na.rm = T))
  df.sd <- df.sd %>% mutate("Sum drylands" = sqrt(rowSums((.[drycats])^2, na.rm = T)),2)
  
  missing <- setdiff(c("model","period","Hyperarid","Arid","Semi-arid","Dry subhumid","Sum drylands","Humid","Cold","NA"), names(df.percent))
  
  df.percent[, missing] <- 0
  df.sd[,missing] <- 0
  
  df.percent <- df.percent %>% select(c("model","period","Hyperarid","Arid","Semi-arid","Dry subhumid","Sum drylands","Humid","Cold","NA"))
  df.percent[is.na(df.percent)] <- 0
  df.percent[,c(3:9)] <- round(df.percent[,c(3:9)], 1)
  
  tab_list_percent[[i]] <- df.percent
  
  df.sd <- df.sd %>% select(c("model","period","Hyperarid","Arid","Semi-arid","Dry subhumid","Sum drylands","Humid","Cold","NA"))
  df.sd[is.na(df.sd)] <- 0
  df.sd[ , c(3:9)] <- round(df.sd[, c(3:9)], 1)
  
  tab_list_sd[[i]] <- df.sd
}

k_list <- list()

for(i in names(tab_list_percent)){
  
  df <- tab_list_percent[[i]] %>% select(c("model", "period"))
  
  df$Hyperarid <- paste(tab_list_percent[[i]]$Hyperarid, tab_list_sd[[i]]$Hyperarid, sep = " ± ")
  df$Arid <- paste(tab_list_percent[[i]]$Arid, tab_list_sd[[i]]$Arid, sep = " ± ")
  df$'Semi-Arid' <- paste(tab_list_percent[[i]][,5], tab_list_sd[[i]][,5], sep = " ± ")
  df$'Dry subhumid' <- paste(tab_list_percent[[i]][,6], tab_list_sd[[i]][,6], sep = " ± ")
  df$'Sum drylands' <- paste(tab_list_percent[[i]][,7], tab_list_sd[[i]][,7], sep = " ± ")
  df$Humid <- paste(tab_list_percent[[i]]$Humid, tab_list_sd[[i]]$Humid, sep = " ± ")
  df$Cold <- paste(tab_list_percent[[i]]$Cold, tab_list_sd[[i]]$Cold, sep = " ± ")
  
  k_list[[i]] <- kable(df[order(df$model, decreasing = F)], caption = i, digits = 2) %>% kable_styling(bootstrap_options = "bordered") %>%
    column_spec(8, italic = T, background = "#DAE2F7") %>% row_spec(2, bold = T) 
  
}

k_list

```

# Changes in proportion of aridity categories, figure

## Colplot by continent

```{r colplots-futurecat-continent, fig.dim = c(15,25)}
tab.flow <- cmip6s %>% subset(!Continent %in% c("SOUTHERN","PACIFIC","ATLANTIC","INDIAN","ARCTIC") & lat > -55 ) %>%
  group_by(period, model, Continent, cat.AI) %>%
  summarise(count = n()) %>%
  ungroup() %>% group_by(period, model, Continent) %>% mutate(percent = round(count/sum(count)*100, 1), count = NULL)

tab.flow$Continent <- str_replace(tab.flow$Continent, "-", "\n")

df245 <- rbind(subset(tab.flow, model == "historical" & period %in% c("1850_1880", "1970_2000")), subset(tab.flow, model == "SSP245")) %>% mutate(lab.y = (rev(cumsum(rev(percent)))) - percent*0.5) %>% 
  subset(cat.AI %in% c("Hyperarid", "Arid", "Semi-arid", "Dry subhumid", "Humid","Cold"))

b245 <- ggplot(data = df245, aes(x = period, y = percent))+
  geom_col(aes(group = cat.AI, col = cat.AI, fill = cat.AI))+
  geom_label(aes(y = lab.y, label = percent))+
  scale_color_manual(values = col.cat, aesthetics = c("col", "fill"), na.translate = F)+
  facet_grid(rows = vars(Continent), switch = "y")+
  scale_y_continuous(position = "right")+
  labs(title = "SSP 2-4.5", y = "%", x = "")+
  theme_minimal()

df370 <- rbind(subset(tab.flow, model == "historical" & period %in% c("1850_1880", "1970_2000")), subset(tab.flow, model == "SSP370")) %>% mutate(lab.y = (rev(cumsum(rev(percent)))) - percent*0.5) %>% 
  subset(cat.AI %in% c("Hyperarid", "Arid", "Semi-arid", "Dry subhumid", "Humid","Cold"))

b370 <- ggplot(data = df370, aes(x = period, y = percent))+
  geom_col(aes(group = cat.AI, col = cat.AI, fill = cat.AI))+
  geom_label(aes(y = lab.y, label = percent))+
  scale_color_manual(values = col.cat, aesthetics = c("col", "fill"), na.translate = F)+
  facet_grid(rows = vars(Continent), switch = "y")+
  scale_y_continuous(position = "right")+  labs(title = "SSP 3-7.0", y = "%", x = "")+
  theme_minimal()


df585 <- rbind(subset(tab.flow, model == "historical" & period %in% c("1850_1880", "1970_2000")), subset(tab.flow, model == "SSP585")) %>% mutate(lab.y = (rev(cumsum(rev(percent)))) - percent*0.5) %>% 
  subset(cat.AI %in% c("Hyperarid", "Arid", "Semi-arid", "Dry subhumid", "Humid","Cold"))

b585 <- ggplot(data = df585, aes(x = period, y = percent))+
  geom_col(aes(group = cat.AI, col = cat.AI, fill = cat.AI))+
  geom_label(aes(y = lab.y, label = percent))+
  scale_color_manual(values = col.cat, aesthetics = c("col", "fill"), na.translate = F)+
  facet_grid(rows = vars(Continent), switch = "y")+
  scale_y_continuous(position = "right")+  labs(title = "SSP 5-8.5", y = "%", x = "")+
  theme_minimal()


ggarrange(plotlist = list(b245, b370, b585), common.legend = T, legend = "bottom", ncol = 3)
```

# Evolution of aridity index by Continent and period

## Map with labels

```{r plot-evolution}

tab.percent.region <- cmip6s %>% subset(!Continent %in% c("SOUTHERN","PACIFIC","ATLANTIC","INDIAN","ARCTIC") & lat > -55 & period != "1985_2015") %>%
  group_by(period, model, Continent, Name) %>% select(c("diff.AI.mean.percent")) %>%
  summarise(AI.diff.mean = mean(diff.AI.mean.percent, na.rm = T), AI.diff.sd= sd(diff.AI.mean.percent, na.rm = T)) %>%
  ungroup()

ipcc_regions.sf <- st_as_sf(ipcc_regions)

percent.region.sf <- merge(ipcc_regions.sf, tab.percent.region, all.y = T)

```

```{r plot_sf_changes, fig.dim = c(10,10)}
pbreaks <- c(-10, 0, 10, 20, 30, 40)
colscale <- c("#08519c", "#3182bd", "#6baed6", "#9ecae1", "azure","cornsilk", "#f4a582")
#colscale <- c("#c6dbef","white","#fddbc7", "#f4a582", "#d6604d", "#b2182b", "#993344")

map_list <- list()

for(i in c("SSP245","SSP370","SSP585")){
  for (j in c("2030_2060","2070_2100")){
    index <- paste(i, j, sep = " , ")
    g <- ggplot(data = subset(percent.region.sf, period == j & model == i)) + geom_sf(aes(fill = AI.diff.mean))+
          binned_scale(aesthetics = "fill", breaks = pbreaks, palette = function(x) rev(colscale))+
      borders(colour = "grey60")+
      geom_sf_label(aes(label = round(AI.diff.mean,1)),size = 2) +
      labs(title = index)+
      theme_void()+ylim(-55,90)+
      theme(legend.position = "bottom")
    map_list[[index]] <- g
  }}

ggpubr::ggarrange(plotlist = map_list,ncol = 2, nrow = 3, common.legend = T, legend = "bottom")
```

## Point figure

#### By Continent


```{r plot-evolution-continent, fig.dim = c(10,6)}
cmip6s$diff.AI.mean.percent <- with(cmip6s, (AI.mean-AI.ref.mean)/AI.ref.mean*100)

tab.percent.continent <- cmip6s %>% subset(!Continent %in% c("SOUTHERN","PACIFIC","ATLANTIC","INDIAN","ARCTIC") & lat > -55 & period != "1985_2015") %>%
  group_by(period, model, Continent) %>% select(c("diff.AI.mean.percent")) %>%
  summarise(AI.diff.mean = mean(diff.AI.mean.percent, na.rm = T), AI.diff.sd= sd(diff.AI.mean.percent, na.rm = T), AI.diff.cv = AI.diff.sd/abs(AI.diff.mean)) %>%
  ungroup()

df <- tab.percent.continent %>% filter(period != "1850_1880" & Continent != "POLAR") 
ref <- df[which(df$model == "historical"),]
ref1 <- ref %>% mutate(model = "SSP245")
ref2 <- ref %>% mutate(model = "SSP370")
ref3 <- ref %>% mutate(model = "SSP585")

dfref <- rbind(subset(df, model != "historical"), ref1) %>% rbind(ref2) %>% rbind(ref3)  

dfref$Continent <- str_replace(dfref$Continent, "-", "\n")
dfref$Continent <- factor(dfref$Continent, levels = c("NORTH\nAMERICA", "CENTRAL\nAMERICA", "SOUTH\nAMERICA", "EUROPE", "MED\nBASIN", "AFRICA", "ASIA", "OCEANIA"))


dfref$model[which(dfref$model == "SSP245")] <- "SSP 2-4.5"
dfref$model[which(dfref$model == "SSP370")] <- "SSP 3-7.0"
dfref$model[which(dfref$model == "SSP585")] <- "SSP 5-8.5"

dfref$period[which(dfref$period == "1970_2000")] <- "1970-2000"
dfref$period[which(dfref$period == "2030_2060")] <- "2030-2060"
dfref$period[which(dfref$period == "2070_2100")] <- "2070-2100"
    
ggplot(dfref, aes(x = period, y = AI.diff.mean))+
  geom_point(aes(size = AI.diff.cv, col = AI.diff.cv))+
  geom_line(aes(group = Continent), size = 0.1)+
  facet_grid(rows = vars(model), cols = vars(Continent))+
  scale_size_continuous(breaks = c(0,1,5,10,100), transform = "pseudo_log")+
  scale_color_gradient(low = "gray", high = "#D73027", breaks = c(0,1,5,10,100), transform = "pseudo_log")+
  labs(y = "Difference in aridity index compared to 1970-2000, %", x = "")+
  guides(color = guide_legend("Coefficient of variance"), size = guide_legend("Coefficient of variance"))+
  ylim(-40,25)+
  theme_minimal()+
  theme(legend.position = "bottom", axis.text.x = element_text(angle = 90))

```

```{r table-evolution-continent, results = "as.is", fig.dim = c(8,10)}
knitr::kable(dfref) %>% kable_styling(bootstrap_options = "bordered")
```

#### By subregion

```{r plot-evolution-subregion, fig.dim = c(10,10)}
names(tab.percent.continent) <- c("period","model","Continent","AI.diff.mean.cont","AI.diff.sd.cont","AI.diff.cv.cont")

tab.percent.subregion <- cmip6s %>% subset(!Continent %in% c("SOUTHERN","PACIFIC","ATLANTIC","INDIAN","ARCTIC","POLAR") & lat > -55 & period != "1985_2015") %>%
  group_by(period, model, Name, Continent) %>% select(c("diff.AI.mean.percent")) %>%
  summarise(AI.diff.mean = mean(diff.AI.mean.percent, na.rm = T), AI.diff.sd= sd(diff.AI.mean.percent, na.rm = T), AI.diff.cv = AI.diff.sd/AI.diff.mean) %>%
  ungroup()

tab.percent.subregion <- merge(tab.percent.subregion, tab.percent.continent, by = c("period","model","Continent"), all.x = T)
  
df <- tab.percent.subregion %>% filter(period != "1850_1880") 
ref <- df[which(df$model == "historical"),]
ref1 <- ref %>% mutate(model = "SSP245")
ref2 <- ref %>% mutate(model = "SSP370")
ref3 <- ref %>% mutate(model = "SSP585")

dfref <- rbind(subset(df, model != "historical"), ref1) %>% rbind(ref2) %>% rbind(ref3)  

dfref$Name <- str_replace(dfref$Name, "-", "\n") 

dfref$model[which(dfref$model == "SSP245")] <- "SSP 2-4.5"
dfref$model[which(dfref$model == "SSP370")] <- "SSP 3-7.0"
dfref$model[which(dfref$model == "SSP585")] <- "SSP 5-8.5"

dfref$period[which(dfref$period == "1970_2000")] <- "1970-2000"
dfref$period[which(dfref$period == "2030_2060")] <- "2030-2060"
dfref$period[which(dfref$period == "2070_2100")] <- "2070-2100"

for(i in unique(dfref$Continent)){
df <- filter(dfref, Continent == i)
g <- ggplot(df, aes(x = period))+
  geom_point(aes(y = AI.diff.mean, size = AI.diff.cv, col = AI.diff.cv))+
  geom_line(aes(y = AI.diff.mean, group = Name), size = 0.1)+
  geom_point(aes(y = AI.diff.mean.cont, size = AI.diff.cv.cont), col = "gray", alpha = 0.5)+
  geom_line(aes(y = AI.diff.mean.cont, group = Name), size = 0.1, alpha = 0.5)+
  facet_grid(cols = vars(model), rows = vars(Name), switch = "both")+
  scale_size_continuous(breaks = c(0,1,5,10,100), transform = "pseudo_log")+
 scale_color_gradient(low = "#ABD9E9", high = "#D73027", breaks = c(0,1,5,10,100), transform = "pseudo_log")+
  labs(y = "Difference in aridity index compared to 1970-2000, %", x = "", title = i)+
  guides(color = guide_legend("Standard deviation"), size = guide_legend("Standard deviation"))+
  theme_minimal()+
  theme(legend.position = "bottom", axis.text.x = element_text(angle = 90))+  
  scale_y_continuous(position = "right", limits = c(-40,30))
print(g)
}

```

## Change AI in boxplot

```{r same-with-boxplot, fig.dim = c(10,10)}
tab.percent.continent <- cmip6s %>% 
  subset(!Continent %in% c("SOUTHERN","PACIFIC","ATLANTIC","INDIAN","ARCTIC") & lat > -55 & period %in% c("1970_2000","2030_2060","2070_2100"))

df <- tab.percent.continent %>% filter(Continent != "POLAR") 
ref <- df[which(df$model == "historical"),]
ref1 <- ref %>% mutate(model = "SSP245")
ref2 <- ref %>% mutate(model = "SSP370")
ref3 <- ref %>% mutate(model = "SSP585")

dfref <- rbind(subset(df, model != "historical"), ref1) %>% rbind(ref2) %>% rbind(ref3)  

dfref$Continent <- str_replace(dfref$Continent, "-", "\n")

ggplot(dfref, aes(x = period, y = diff.AI.mean.percent))+
  geom_boxplot(aes(col = Continent), outliers = F)+
  facet_grid(rows = vars(model), cols = vars(Continent), space = "free_y")+
  scale_color_viridis_d(end = 0.8)+
  labs(y = "Difference in aridity index compared to 1970-2000, %", x = "")+
  guides(color = F)+
  theme_minimal()+
  theme(legend.position = "bottom", axis.text.x = element_text(angle = 90))
  
```

## Point figure precipitation and temperature

```{r plot-evolution-temp-precip-continent, fig.dim = c(10,6)}
cmip6s$pr.percent <- with(cmip6s, diff.pr.mean/spr.ref.mean*100)

tab.percent.continent <- cmip6s %>% subset(!Continent %in% c("SOUTHERN","PACIFIC","ATLANTIC","INDIAN","ARCTIC") & lat > -55 & period != "1985_2015") %>%
  group_by(period, model, Continent) %>% select(c("diff.t.mean", "pr.percent")) %>%
  summarise(t.diff.mean = mean(diff.t.mean, na.rm = T), t.diff.sd= sd(diff.t.mean, na.rm = T), t.diff.cv = t.diff.sd/abs(t.diff.mean),
            pr.diff.mean = mean(pr.percent, na.rm = T), pr.diff.sd= sd(pr.percent, na.rm = T), pr.diff.cv = pr.diff.sd/abs(pr.diff.mean)) %>%
  ungroup()


df <- tab.percent.continent %>% filter(period != "1850_1880" & Continent != "POLAR") 
ref <- df[which(df$model == "historical"),]
ref1 <- ref %>% mutate(model = "SSP245")
ref2 <- ref %>% mutate(model = "SSP370")
ref3 <- ref %>% mutate(model = "SSP585")

dfref <- rbind(subset(df, model != "historical"), ref1) %>% rbind(ref2) %>% rbind(ref3)  

dfref$Continent <- str_replace(dfref$Continent, "-", "\n")
dfref$Continent <- factor(dfref$Continent, levels = c("NORTH\nAMERICA", "CENTRAL\nAMERICA", "SOUTH\nAMERICA", "EUROPE", "MED\nBASIN", "AFRICA", "ASIA", "OCEANIA"))


dfref$model[which(dfref$model == "SSP245")] <- "SSP 2-4.5"
dfref$model[which(dfref$model == "SSP370")] <- "SSP 3-7.0"
dfref$model[which(dfref$model == "SSP585")] <- "SSP 5-8.5"

dfref$period[which(dfref$period == "1970_2000")] <- "1970-2000"
dfref$period[which(dfref$period == "2030_2060")] <- "2030-2060"
dfref$period[which(dfref$period == "2070_2100")] <- "2070-2100"
    
ggplot(dfref, aes(x = period))+
  geom_point(aes(y = t.diff.mean*3, size = t.diff.cv, col = "Temperature, °C"), alpha = 0.5)+
  geom_line(aes(y = t.diff.mean*3, group = Continent, col = "Temperature, °C"), size = 0.1)+
  
  geom_point(aes(y = pr.diff.mean, size = pr.diff.cv, col = "Precipitation, %"), alpha = 0.5)+
  geom_line(aes(y = pr.diff.mean, group = Continent, col = "Precipitation, %"), size = 0.1)+
  scale_size_continuous(breaks = c(0,1,5,10,20), transform = "pseudo_log")+
  scale_color_manual(values = c("#3182bd", "#d6604d"))+
  facet_grid(rows = vars(model), cols = vars(Continent))+
  labs(y = "Precipitation anomalies, %", x = "", size = "Coefficient of variance", color = "")+
  ylim(-45,25)+
  theme_minimal()+
  theme(legend.position = "bottom", axis.text.x = element_text(angle = 90))+
  
  scale_y_continuous(sec.axis = sec_axis(trans = ~./3, name = "Temperature anomalies, °C"))
```

```{r table-evolution-continent-p-t}
knitr::kable(dfref) %>% kable_styling(bootstrap_options = "bordered")
```

```{r plot-evolution-subregion-temp-precip, fig.dim = c(10,50)}

tab.percent.continent <- cmip6s %>% subset(!Continent %in% c("SOUTHERN","PACIFIC","ATLANTIC","INDIAN","ARCTIC") & lat > -55 & period != "1985_2015") %>%
  group_by(period, model, Name) %>% select(c("diff.t.mean", "pr.percent")) %>%
  summarise(t.diff.mean = mean(diff.t.mean, na.rm = T), t.diff.sd= sd(diff.t.mean, na.rm = T),
            pr.diff.mean = mean(pr.percent, na.rm = T), pr.diff.sd= sd(pr.percent, na.rm = T)) %>%
  ungroup()


df <- tab.percent.continent %>% filter(period != "1850_1880") 
ref <- df[which(df$model == "historical"),]
ref1 <- ref %>% mutate(model = "SSP245")
ref2 <- ref %>% mutate(model = "SSP370")
ref3 <- ref %>% mutate(model = "SSP585")

dfref <- rbind(subset(df, model != "historical"), ref1) %>% rbind(ref2) %>% rbind(ref3)  

dfref$Name <- str_replace(dfref$Name, "-", "\n") 

ggplot(dfref, aes(x = period))+
  geom_point(aes(y = t.diff.mean*3, size = t.diff.sd, col = "Temperature, °C"), alpha = 0.5)+
  geom_line(aes(y = t.diff.mean*3, group = Name, col = "Temperature, °C"), size = 0.1)+
  
  geom_point(aes(y = pr.diff.mean, size = pr.diff.sd, col = "Precipitation, %"), alpha = 0.5)+
  geom_line(aes(y = pr.diff.mean, group = Name, col = "Precipitation, %"), size = 0.1)+
  
  scale_color_manual(values = c("#3182bd", "#d6604d"))+
  facet_grid(cols = vars(model), rows = vars(Name), switch = "both")+
  labs(y = "Precipitation anomalies, %", x = "", size = "Standard deviation", color = "")+
  ylim(-45,25)+
  theme_minimal()+
  theme(legend.position = "bottom", axis.text.x = element_text(angle = 90))+
  
  scale_y_continuous(sec.axis = sec_axis(trans = ~./3, name = "Temperature anomalies, °C"))
```

# Changes of precipitation and temperature over time and scenarios 

## Temperature 

### Absolute temperature 

```{r temperature}
quantile(cmip6s$t.mean, probs = seq(0,1,0.1)) 

t.breaks <- c(-50,-40,-30,-20,-10,0,10,20,30)
colscale <- c("#08519c", "#3182bd", "#6baed6", "#9ecae1", "#c6dbef", "#fddbc7", "#f4a582", "#d6604d", "#b2182b", "#67001f")
```

```{r map-historical-changes-quantiles-temperature, fig.dim = c(5,10)}
map_list <- list()

for(i in c("1850_1880", "1970_2000","1985_2015")){
  g <- ggplot() + geom_raster(data = subset(cmip6s, period == i & model == "historical"), aes(x=lon, y = lat,  fill = t.mean))+
    borders(colour = "grey60")+
    binned_scale(aesthetics = "fill", breaks = t.breaks, palette = function(x) colscale,
                 guide = guide_legend(label.theme = element_text(angle = 0)))+
    labs(title = i, fill = "Mean annual temperature, °C")+
    theme_void()+ylim(-55,90)+
    theme(legend.position = "right")
  map_list[[i]] <- g
}

ggpubr::ggarrange(plotlist = map_list, ncol = 1, nrow = 3, common.legend = T, legend = "bottom")

```

```{r map-future-changes-quantiles-temperature, fig.dim = c(10,10)}

map_list <- list()

for(i in c("SSP245","SSP370","SSP585")){
  for (j in c("2030_2060","2070_2100")){
    index <- paste(i, j, sep = " , ")
    g <- ggplot() + geom_raster(data = subset(cmip6s, period == j & model == i), aes(x=lon, y = lat,  fill = t.mean))+
      borders(colour = "grey60")+
      binned_scale(aesthetics = "fill", breaks = t.breaks, palette = function(x) colscale,
                   guide = guide_legend(label.theme = element_text(angle = 0)))+
      labs(fill = "Mean annual temperature, °C", title = index)+
      theme_void()+ylim(-55,90)+
      theme(legend.position = "bottom")
    map_list[[index]] <- g
  }}



ggpubr::ggarrange(plotlist = map_list,ncol = 2, nrow = 3, common.legend = T, legend = "bottom")

#labels = c("2030-2060", "2070-2100"
```

### Temperature anomalies

```{r temperature-anomalies}
quantile(cmip6s$diff.t.mean, probs = seq(0,1,0.1)) 

t.breaks <- c(-3,-0.5,0,1.5,2,3,4,5,10)
colscale <- c("#08519c", "#3182bd", "#6baed6", "#9ecae1", "#c6dbef", "#fddbc7", "#f4a582", "#d6604d", "#b2182b", "#67001f")
```

```{r map-historical-changes-quantiles-temperature-anomalies, fig.dim = c(5,10)}
map_list <- list()

for(i in c("1850_1880", "1970_2000","1985_2015")){
  g <- ggplot() + geom_raster(data = subset(cmip6s, period == i & model == "historical"), aes(x=lon, y = lat,  fill = diff.t.mean))+
    borders(colour = "grey60")+
   scale_fill_gradient2(low = "#08519c",mid = "white", high = "#b2182b", midpoint = 0, transform = "pseudo_log", breaks = c(-2, -1, 0, 1, 2, 3, 4))+
      labs(fill = "Precipitation anomalies in %", title = i)+
      theme_void()+ylim(-55,90)+
      theme(legend.position = "bottom", legend.text =  element_text(angle = 0))
  map_list[[i]] <- g
}

ggpubr::ggarrange(plotlist = map_list, ncol = 1, nrow = 3, common.legend = T, legend = "bottom")

```

```{r map-future-changes-quantiles-temperature-anomalies, fig.dim = c(10,10)}

map_list <- list()

for(i in c("SSP245","SSP370","SSP585")){
  for (j in c("2030_2060","2070_2100")){
    index <- paste(i, j, sep = " , ")
    g <- ggplot() + geom_tile(data = subset(cmip6s, period == j & model == i), aes(x=lon, y = lat,  fill = diff.t.mean))+
      borders(colour = "grey60")+
   scale_fill_gradient2(low = "#08519c",mid = "white", high = "#b2182b", midpoint = 0, transform = "pseudo_log", breaks = c(-2, -1, 0, 1, 2, 3, 4))+
      labs(fill = "Temperature anomalies in °C", title = index)+
      theme_void()+ylim(-55,90)+
      theme(legend.position = "bottom", legend.text =  element_text(angle = 0))
    map_list[[index]] <- g
  }}

ggpubr::ggarrange(plotlist = map_list,ncol = 2, nrow = 3, common.legend = T, legend = "bottom")

```
### Binned anomalies

```{r map-future-changes-quantiles-temperature-anomalies-binned, fig.dim = c(10,10)}
YOR <- brewer.pal(n = 11, name ="YlOrRd")
coltemp <- c("[0,1]" = YOR[1], "(1,2]" = YOR[2], "(2,3]" = YOR[3], "(3,4]" = YOR[4], "(4,5]" = YOR[5], "(5,6]" = YOR[6], "(6,7]" = YOR[7], "(7,8]" = YOR[8], "(8,9]" = YOR[9], "(9,10]" = "grey20")


map_list <- list()

for(i in c("SSP245","SSP370","SSP585")){
  for (j in c("2030_2060","2070_2100")){
    index <- paste(i, j, sep = " , ")
    g <- ggplot(data = subset(cmip6s, period == j & model == i), 
                aes(x=lon, y = lat,  fill = cut(diff.t.mean, breaks = c(0,1,2,3,4,5,6,7,8,9,10), include.lowest = T))) + 
      geom_tile(show.legend = T)+
      borders(colour = "grey60")+
      scale_fill_manual(values = coltemp, drop = F)+
      labs(fill = "Temperature\nanomalies in °C", title = index)+
      theme_void()+ylim(-55,90)+
     theme(legend.position = "bottom", legend.text =  element_text(angle = 0))+
    guides(fill = guide_legend(nrow = 2))
    map_list[[index]] <- g
  }}

ggpubr::ggarrange(plotlist = map_list,ncol = 2, nrow = 3, common.legend = T, legend = "bottom")
```

## Precipitation 

### Absolute values of precipitations

```{r precipitation}
quantile(cmip6s$pr.mean, probs = seq(0,1,0.1)) 

p.breaks <- c(10,50,100,200,300,400,500,600,700)
colscale <- c("#08519c", "#3182bd", "#6baed6", "#9ecae1", "#c6dbef", "#fddbc7", "#f4a582", "#d6604d", "#b2182b", "#67001f")
```

```{r map-historical-changes-quantiles-precipitation, fig.dim = c(5,10)}
map_list <- list()

for(i in c("1850_1880","1970_2000","1985_2015")){
  g <- ggplot() + geom_raster(data = subset(cmip6s, period == i & model == "historical"), aes(x=lon, y = lat,  fill = pr.mean))+
    borders(colour = "grey60")+
    binned_scale(aesthetics = "fill", breaks = p.breaks, palette = function(x) rev(colscale),
                 guide = guide_legend(label.theme = element_text(angle = 0)))+
    labs(title = i, fill = "Annual mean\nprecipitation, mm")+
    theme_void()+ylim(-55,90)+
    theme(legend.position = "right")
  map_list[[i]] <- g
}

ggpubr::ggarrange(plotlist = map_list, ncol = 1, nrow = 3, common.legend = T, legend = "bottom")

```

```{r map-future-changes-quantiles-precipitation, fig.dim = c(10,10)}

map_list <- list()

for(i in c("SSP245","SSP370","SSP585")){
  for (j in c("2030_2060","2070_2100")){
    index <- paste(i, j, sep = " , ")
    g <- ggplot() + geom_raster(data = subset(cmip6s, period == j & model == i), aes(x=lon, y = lat,  fill = pr.mean))+
      borders(colour = "grey60")+
      binned_scale(aesthetics = "fill", breaks = p.breaks, palette = function(x) rev(colscale),
                   guide = guide_legend(label.theme = element_text(angle = 0)))+
      labs(fill = "Annual mean\nprecipitation, mm", title = index)+
      theme_void()+ylim(-55,90)+
      theme(legend.position = "bottom")
    map_list[[index]] <- g
  }}

ggpubr::ggarrange(plotlist = map_list,ncol = 2, nrow = 3, common.legend = T, legend = "bottom")
```

### Precipitation anomalies


#### Gradient anomalies

```{r precipitation-anomalies}
cmip6s$pr.percent <- with(cmip6s, diff.pr.mean/spr.ref.mean*100)

quantile(cmip6s$pr.percent, probs = seq(0,1,0.1)) 

p.breaks <- c(-43,-3,0,3,5,9,13,18,24,148)
colscale <- c("#08519c", "#3182bd", "#6baed6", "#9ecae1", "#c6dbef", "#fddbc7", "#f4a582", "#d6604d", "#b2182b", "#67001f")
```

```{r map-historical-changes-quantiles-precipitation-anomalies, fig.dim = c(5,10)}
map_list <- list()

for(i in c("1850_1880","1970_2000","1985_2015")){
  g <- ggplot() + geom_raster(data = subset(cmip6s, period == i & model == "historical"), aes(x=lon, y = lat,  fill = pr.percent))+
    borders(colour = "grey60")+
   scale_fill_gradient2(high = "#08519c", mid = "white", low = "#b2182b", midpoint = 0, transform = "pseudo_log", breaks = c(-10, -5, -1, 0, 1, 5, 10, 15, 30))+
      labs(fill = "Precipitation anomalies in %", title = i)+
      theme_void()+ylim(-55,90)+
      theme(legend.position = "bottom", legend.text =  element_text(angle = 0))
  map_list[[i]] <- g
}

ggpubr::ggarrange(plotlist = map_list, ncol = 1, nrow = 3, common.legend = T, legend = "bottom")

```

```{r map-future-changes-quantiles-precipitation-anomalies, fig.dim = c(10,10)}

map_list <- list()

for(i in c("SSP245","SSP370","SSP585")){
  for (j in c("2030_2060","2070_2100")){
    index <- paste(i, j, sep = " , ")
    g <- ggplot() + geom_raster(data = subset(cmip6s, period == j & model == i), aes(x=lon, y = lat,  fill = pr.percent))+
      borders(colour = "grey60")+
   scale_fill_gradient2(high = "#08519c", mid = "white", low = "#b2182b", midpoint = 0, transform = "pseudo_log", breaks = c(-10, -5, -1, 0, 1, 5, 10, 15, 30))+
      labs(fill = "Precipitation anomalies in %", title = index)+
      theme_void()+ylim(-55,90)+
      theme(legend.position = "bottom", legend.text =  element_text(angle = 0))
    map_list[[index]] <- g
  }}

ggpubr::ggarrange(plotlist = map_list,ncol = 2, nrow = 3, common.legend = T, legend = "bottom")
```

#### Binned anomalies

```{r map-future-changes-quantiles-precipitation-anomalies-binned, fig.dim = c(10,10)}

map_list <- list()

for(i in c("SSP245","SSP370","SSP585")){
  for (j in c("2030_2060","2070_2100")){
    index <- paste(i, j, sep = " , ")
    g <- ggplot() + geom_raster(data = subset(cmip6s, period == j & model == i), aes(x=lon, y = lat,  fill = pr.percent))+
      borders(colour = "grey60")+
   scale_fill_binned_divergingx(mid = 0, palette = "RdBu", rev = F, n_interp = 7, breaks = c(-20, -10, 0, 10, 20, 30, 40, 50, 60))+
      labs(fill = "Precipitation anomalies in %", title = index)+
      theme_void()+ylim(-55,90)+
      theme(legend.position = "bottom", legend.text =  element_text(angle = 0))
    map_list[[index]] <- g
  }}

ggpubr::ggarrange(plotlist = map_list,ncol = 2, nrow = 3, common.legend = T, legend = "bottom")
```


```{r map-future-changes-quantiles-temperature-anomalies-binned-bis, fig.dim = c(10,10)}
BR <- brewer.pal(n = 11, name ="RdBu")
coltemp <- c("[-20,-10]" = BR[4], "(-10,0]" = BR[6], "(0,10]" = BR[6], "(10,20]" = BR[8], "(20,30]" = BR[9], "(30,40]" = BR[10], "(40,50]" = BR[11], "(50,60]" = "grey20")


map_list <- list()

for(i in c("SSP245","SSP370","SSP585")){
  for (j in c("2030_2060","2070_2100")){
    index <- paste(i, j, sep = " , ")
    g <- ggplot(data = subset(cmip6s, period == j & model == i), 
                aes(x=lon, y = lat,  fill = cut(pr.percent, breaks = c(-20, -10, 0, 10, 20, 30, 40, 50, 60), include.lowest = T))) + 
      geom_tile(show.legend = T)+
      borders(colour = "grey60")+
      scale_fill_manual(values = coltemp, drop = F, na.translate = T)+
      labs(fill = "Precipitation anomalies in %", title = index)+
      theme_void()+ylim(-55,90)+
     theme(legend.position = "bottom", legend.text =  element_text(angle = 0))+
    guides(fill = guide_legend(nrow = 2))
    map_list[[index]] <- g
  }}

ggpubr::ggarrange(plotlist = map_list,ncol = 2, nrow = 3, common.legend = T, legend = "bottom")
```

```{r map-future-changes-quantiles-temperature-anomalies-binned-ter, fig.dim = c(10,10)}
BR <- brewer.pal(n = 11, name ="RdBu")
coltemp <- c("[-20,-10]" = BR[4], "(-10,0]" = BR[5], "(0,10]" = BR[7], "(10,20]" = BR[8], "(20,30]" = BR[9], "(30,40]" = BR[10], "(40,50]" = BR[11], "(50,60]" = "grey20")


map_list <- list()

for(i in c("SSP245","SSP370","SSP585")){
  for (j in c("2030_2060","2070_2100")){
    index <- paste(i, j, sep = " , ")
    g <- ggplot(data = subset(cmip6s, period == j & model == i), 
                aes(x=lon, y = lat,  fill = cut(pr.percent, breaks = c(-20, -10, 0, 10, 20, 30, 40, 50, 60), include.lowest = T))) + 
      geom_tile(show.legend = T)+
      borders(colour = "grey60")+
      scale_fill_manual(values = coltemp, drop = F, na.translate = T)+
      labs(fill = "Precipitation anomalies in %", title = index)+
      theme_void()+ylim(-55,90)+
     theme(legend.position = "bottom", legend.text =  element_text(angle = 0))+
    guides(fill = guide_legend(nrow = 2))
    map_list[[index]] <- g
  }}

ggpubr::ggarrange(plotlist = map_list,ncol = 2, nrow = 3, common.legend = T, legend = "bottom")
```


# Seasonnal aridity

```{r compute-monthly-aridity, eval = F}
cmip6 <- read.table("cmip6_AI.txt")

cmip6$AI_01 <- with(cmip6, pr_01/ET0_01)
cmip6$AI_02 <- with(cmip6, pr_02/ET0_02)
cmip6$AI_03 <- with(cmip6, pr_03/ET0_03)
cmip6$AI_04 <- with(cmip6, pr_04/ET0_04)
cmip6$AI_05 <- with(cmip6, pr_05/ET0_05)
cmip6$AI_06 <- with(cmip6, pr_06/ET0_06)
cmip6$AI_07 <- with(cmip6, pr_07/ET0_07)
cmip6$AI_08 <- with(cmip6, pr_08/ET0_08)
cmip6$AI_09 <- with(cmip6, pr_09/ET0_09)
cmip6$AI_10 <- with(cmip6, pr_10/ET0_10)
cmip6$AI_11 <- with(cmip6, pr_11/ET0_11)
cmip6$AI_12 <- with(cmip6, pr_12/ET0_12)


```

```{r ref-monthly-aridity, eval = F}
cmip6.monthly.ref <- cmip6 %>% filter(period == "1970_2000") %>% 
  select(c("lon","lat","model","period","source", "Continent","Name",
           "AI_01","pr_01","ET0_01",
           "AI_02","pr_02","ET0_02",
           "AI_03","pr_03","ET0_03",
           "AI_04","pr_04","ET0_04",
           "AI_05","pr_05","ET0_05",
           "AI_06","pr_06","ET0_06",
           "AI_07","pr_07","ET0_07",
           "AI_08","pr_08","ET0_08",
           "AI_09","pr_09","ET0_09",
           "AI_10","pr_10","ET0_10",
           "AI_11","pr_11","ET0_11",
           "AI_12","pr_12","ET0_12"
           )) %>% 
  setNames(c("lon","lat","model","period","source","Continent","Name",
          "AI_01.ref","pr_01.ref","ET0_01.ref",
           "AI_02.ref","pr_02.ref","ET0_02.ref",
           "AI_03.ref","pr_03.ref","ET0_03.ref",
           "AI_04.ref","pr_04.ref","ET0_04.ref",
           "AI_05.ref","pr_05.ref","ET0_05.ref",
           "AI_06.ref","pr_06.ref","ET0_06.ref",
           "AI_07.ref","pr_07.ref","ET0_07.ref",
           "AI_08.ref","pr_08.ref","ET0_08.ref",
           "AI_09.ref","pr_09.ref","ET0_09.ref",
           "AI_10.ref","pr_10.ref","ET0_10.ref",
           "AI_11.ref","pr_11.ref","ET0_11.ref",
           "AI_12.ref","pr_12.ref","ET0_12.ref"
           ))

cmip6monthly <- merge(
  select(cmip6, c("lon","lat","source","model","period","source", "Continent","Name",
            "AI_01","pr_01","ET0_01",
           "AI_02","pr_02","ET0_02",
           "AI_03","pr_03","ET0_03",
           "AI_04","pr_04","ET0_04",
           "AI_05","pr_05","ET0_05",
           "AI_06","pr_06","ET0_06",
           "AI_07","pr_07","ET0_07",
           "AI_08","pr_08","ET0_08",
           "AI_09","pr_09","ET0_09",
           "AI_10","pr_10","ET0_10",
           "AI_11","pr_11","ET0_11",
           "AI_12","pr_12","ET0_12")),
  select(cmip6.monthly.ref, c("lon","lat","source",
           "AI_01.ref","pr_01.ref","ET0_01.ref",
           "AI_02.ref","pr_02.ref","ET0_02.ref",
           "AI_03.ref","pr_03.ref","ET0_03.ref",
           "AI_04.ref","pr_04.ref","ET0_04.ref",
           "AI_05.ref","pr_05.ref","ET0_05.ref",
           "AI_06.ref","pr_06.ref","ET0_06.ref",
           "AI_07.ref","pr_07.ref","ET0_07.ref",
           "AI_08.ref","pr_08.ref","ET0_08.ref",
           "AI_09.ref","pr_09.ref","ET0_09.ref",
           "AI_10.ref","pr_10.ref","ET0_10.ref",
           "AI_11.ref","pr_11.ref","ET0_11.ref",
           "AI_12.ref","pr_12.ref","ET0_12.ref")),
  by = c("lon","lat","source"), all.x = T) 

write.table(cmip6monthly,"cmip6monthly.txt")
```

```{r summary-table-monthly, eval = F}
cmip6monthly <- read.table("cmip6monthly.txt")

cmip6smonthly <- cmip6monthly %>% 
  group_by(lon,lat, Continent, Name, model, period) %>% 
  dplyr::summarise(AI_01_mean = mean(AI_01, na.rm = T), AI_01_sd = sd(AI_01, na.rm = T),
            AI_02_mean = mean(AI_02, na.rm = T), AI_02_sd = sd(AI_02, na.rm = T),
            AI_03_mean = mean(AI_03, na.rm = T), AI_03_sd = sd(AI_03, na.rm = T),
            AI_04_mean = mean(AI_04, na.rm = T), AI_04_sd = sd(AI_04, na.rm = T),
            AI_05_mean = mean(AI_05, na.rm = T), AI_05_sd = sd(AI_05, na.rm = T),
            AI_06_mean = mean(AI_06, na.rm = T), AI_06_sd = sd(AI_06, na.rm = T),
            AI_07_mean = mean(AI_07, na.rm = T), AI_07_sd = sd(AI_07, na.rm = T),
            AI_08_mean = mean(AI_08, na.rm = T), AI_08_sd = sd(AI_08, na.rm = T),
            AI_09_mean = mean(AI_09, na.rm = T), AI_09_sd = sd(AI_09, na.rm = T),
            AI_10_mean = mean(AI_10, na.rm = T), AI_10_sd = sd(AI_10, na.rm = T),
            AI_11_mean = mean(AI_11, na.rm = T), AI_11_sd = sd(AI_11, na.rm = T),
            AI_12_mean = mean(AI_12, na.rm = T), AI_11_sd = sd(AI_12, na.rm = T),
            ET0_01_mean = mean(ET0_01, na.rm = T), ET0_01_sd = sd(ET0_01, na.rm = T),
            ET0_02_mean = mean(ET0_02, na.rm = T), ET0_02_sd = sd(ET0_02, na.rm = T),
            ET0_03_mean = mean(ET0_03, na.rm = T), ET0_03_sd = sd(ET0_03, na.rm = T),
            ET0_04_mean = mean(ET0_04, na.rm = T), ET0_04_sd = sd(ET0_04, na.rm = T),
            ET0_05_mean = mean(ET0_05, na.rm = T), ET0_05_sd = sd(ET0_05, na.rm = T),
            ET0_06_mean = mean(ET0_06, na.rm = T), ET0_06_sd = sd(ET0_06, na.rm = T),
            ET0_07_mean = mean(ET0_07, na.rm = T), ET0_07_sd = sd(ET0_07, na.rm = T),
            ET0_08_mean = mean(ET0_08, na.rm = T), ET0_08_sd = sd(ET0_08, na.rm = T),
            ET0_09_mean = mean(ET0_09, na.rm = T), ET0_09_sd = sd(ET0_09, na.rm = T),
            ET0_10_mean = mean(ET0_10, na.rm = T), ET0_10_sd = sd(ET0_10, na.rm = T),
            ET0_11_mean = mean(ET0_11, na.rm = T), ET0_11_sd = sd(ET0_11, na.rm = T),
            ET0_12_mean = mean(ET0_12, na.rm = T), ET0_11_sd = sd(ET0_12, na.rm = T),
            pr_01_mean = mean(pr_01, na.rm = T), pr_01_sd = sd(pr_01, na.rm = T),
            pr_02_mean = mean(pr_02, na.rm = T), pr_02_sd = sd(pr_02, na.rm = T),
            pr_03_mean = mean(pr_03, na.rm = T), pr_03_sd = sd(pr_03, na.rm = T),
            pr_04_mean = mean(pr_04, na.rm = T), pr_04_sd = sd(pr_04, na.rm = T),
            pr_05_mean = mean(pr_05, na.rm = T), pr_05_sd = sd(pr_05, na.rm = T),
            pr_06_mean = mean(pr_06, na.rm = T), pr_06_sd = sd(pr_06, na.rm = T),
            pr_07_mean = mean(pr_07, na.rm = T), pr_07_sd = sd(pr_07, na.rm = T),
            pr_08_mean = mean(pr_08, na.rm = T), pr_08_sd = sd(pr_08, na.rm = T),
            pr_09_mean = mean(pr_09, na.rm = T), pr_09_sd = sd(pr_09, na.rm = T),
            pr_10_mean = mean(pr_10, na.rm = T), pr_10_sd = sd(pr_10, na.rm = T),
            pr_11_mean = mean(pr_11, na.rm = T), pr_11_sd = sd(pr_11, na.rm = T),
            pr_12_mean = mean(pr_12, na.rm = T), pr_11_sd = sd(pr_12, na.rm = T)
            ) %>%
 ungroup() %>% as.data.frame()

cmip6smonthly.ref <- cmip6smonthly %>% filter(period == "1970_2000") %>% 
  select(c("lon","lat","model","period", "Continent","Name",
           "AI_01_mean","ET0_01_mean","pr_01_mean",
           "AI_02_mean","ET0_02_mean","pr_02_mean",
           "AI_03_mean","ET0_03_mean","pr_03_mean",
           "AI_04_mean","ET0_04_mean","pr_04_mean",
           "AI_05_mean","ET0_05_mean","pr_05_mean",
           "AI_06_mean","ET0_06_mean","pr_06_mean",
           "AI_07_mean","ET0_07_mean","pr_07_mean",
           "AI_08_mean","ET0_08_mean","pr_08_mean",
           "AI_09_mean","ET0_09_mean","pr_09_mean",
           "AI_10_mean","ET0_10_mean","pr_10_mean",
           "AI_11_mean","ET0_11_mean","pr_11_mean",
           "AI_12_mean","ET0_12_mean""pr_12_mean",
           )) %>% 
  setNames(c("lon","lat","model","period","Continent","Name",
          "AI_01_mean.ref","ET0_01_mean.ref","pr_01_mean.ref",
           "AI_02_mean.ref","ET0_02_mean.ref","pr_02_mean.ref",
           "AI_03_mean.ref","ET0_03_mean.ref","pr_03_mean.ref",
           "AI_04_mean.ref","ET0_04_mean.ref","pr_04_mean.ref",
           "AI_05_mean.ref","ET0_05_mean.ref","pr_05_mean.ref",
           "AI_06_mean.ref","ET0_06_mean.ref","pr_06_mean.ref",
           "AI_07_mean.ref","ET0_07_mean.ref","pr_07_mean.ref",
           "AI_08_mean.ref","ET0_08_mean.ref","pr_08_mean.ref",
           "AI_09_mean.ref","ET0_09_mean.ref","pr_09_mean.ref",
           "AI_10_mean.ref","ET0_10_mean.ref","pr_10_mean.ref",
           "AI_11_mean.ref","ET0_11_mean.ref","pr_11_mean.ref",
           "AI_12_mean.ref","ET0_12_mean.ref","pr_12_mean.ref"
           ))

cmip6smonthly <- merge(cmip6smonthly,cmip6smonthly.ref, by = c("lon","lat","model","period", "Continent","Name"), all.x = T)

# CATEGORIES 01
cmip6smonthly$cat.AI_01 <- cmip6smonthly$AI_01_mean %>% cut(breaks = breaks.unesco) %>% revalue(cat.unesco)
cmip6smonthly$cat.AI_01[which(cmip6smonthly$ET0_01_mean < 400/360)] <- "Cold"

cmip6smonthly$cat.AI_01.ref <- cmip6smonthly$AI_01_mean.ref %>% cut(breaks = breaks.unesco) %>% revalue(cat.unesco)
cmip6smonthly$cat.AI_01.ref[which(cmip6smonthly$ET0_01_mean.ref < 400/360)] <- "Cold"

# CATEGORIES 02
cmip6smonthly$cat.AI_02 <- cmip6smonthly$AI_02_mean %>% cut(breaks = breaks.unesco) %>% revalue(cat.unesco)
cmip6smonthly$cat.AI_02[which(cmip6smonthly$ET0_02_mean < 400/360)] <- "Cold"

cmip6smonthly$cat.AI_02.ref <- cmip6smonthly$AI_02_mean.ref %>% cut(breaks = breaks.unesco) %>% revalue(cat.unesco)
cmip6smonthly$cat.AI_02.ref[which(cmip6smonthly$ET0_02_mean.ref < 400/360)] <- "Cold"

# CATEGORIES 03
cmip6smonthly$cat.AI_03 <- cmip6smonthly$AI_03_mean %>% cut(breaks = breaks.unesco) %>% revalue(cat.unesco)
cmip6smonthly$cat.AI_03[which(cmip6smonthly$ET0_03_mean < 400/360)] <- "Cold"

cmip6smonthly$cat.AI_03.ref <- cmip6smonthly$AI_03_mean.ref %>% cut(breaks = breaks.unesco) %>% revalue(cat.unesco)
cmip6smonthly$cat.AI_03.ref[which(cmip6smonthly$ET0_03_mean.ref < 400/360)] <- "Cold"

# CATEGORIES 04
cmip6smonthly$cat.AI_04 <- cmip6smonthly$AI_04_mean %>% cut(breaks = breaks.unesco) %>% revalue(cat.unesco)
cmip6smonthly$cat.AI_04[which(cmip6smonthly$ET0_04_mean < 400/360)] <- "Cold"

cmip6smonthly$cat.AI_04.ref <- cmip6smonthly$AI_05_mean.ref %>% cut(breaks = breaks.unesco) %>% revalue(cat.unesco)
cmip6smonthly$cat.AI_04.ref[which(cmip6smonthly$ET0_05_mean.ref < 400/360)] <- "Cold"

# CATEGORIES 05
cmip6smonthly$cat.AI_05 <- cmip6smonthly$AI_05_mean %>% cut(breaks = breaks.unesco) %>% revalue(cat.unesco)
cmip6smonthly$cat.AI_05[which(cmip6smonthly$ET0_05_mean < 400/360)] <- "Cold"

cmip6smonthly$cat.AI_05.ref <- cmip6smonthly$AI_05_mean.ref %>% cut(breaks = breaks.unesco) %>% revalue(cat.unesco)
cmip6smonthly$cat.AI_05.ref[which(cmip6smonthly$ET0_05_mean.ref < 400/360)] <- "Cold"

# CATEGORIES 06
cmip6smonthly$cat.AI_06 <- cmip6smonthly$AI_06_mean %>% cut(breaks = breaks.unesco) %>% revalue(cat.unesco)
cmip6smonthly$cat.AI_06[which(cmip6smonthly$ET0_06_mean < 400/360)] <- "Cold"

cmip6smonthly$cat.AI_06.ref <- cmip6smonthly$AI_06_mean.ref %>% cut(breaks = breaks.unesco) %>% revalue(cat.unesco)
cmip6smonthly$cat.AI_06.ref[which(cmip6smonthly$ET0_06_mean.ref < 400/360)] <- "Cold"

# CATEGORIES 07
cmip6smonthly$cat.AI_07 <- cmip6smonthly$AI_07_mean %>% cut(breaks = breaks.unesco) %>% revalue(cat.unesco)
cmip6smonthly$cat.AI_07[which(cmip6smonthly$ET0_07_mean < 400/360)] <- "Cold"

cmip6smonthly$cat.AI_07.ref <- cmip6smonthly$AI_07_mean.ref %>% cut(breaks = breaks.unesco) %>% revalue(cat.unesco)
cmip6smonthly$cat.AI_07.ref[which(cmip6smonthly$ET0_07_mean.ref < 400/360)] <- "Cold"

# CATEGORIES 08
cmip6smonthly$cat.AI_08 <- cmip6smonthly$AI_08_mean %>% cut(breaks = breaks.unesco) %>% revalue(cat.unesco)
cmip6smonthly$cat.AI_08[which(cmip6smonthly$ET0_08_mean < 400/360)] <- "Cold"

cmip6smonthly$cat.AI_08.ref <- cmip6smonthly$AI_08_mean.ref %>% cut(breaks = breaks.unesco) %>% revalue(cat.unesco)
cmip6smonthly$cat.AI_08.ref[which(cmip6smonthly$ET0_08_mean.ref < 400/360)] <- "Cold"

# CATEGORIES 09
cmip6smonthly$cat.AI_09 <- cmip6smonthly$AI_09_mean %>% cut(breaks = breaks.unesco) %>% revalue(cat.unesco)
cmip6smonthly$cat.AI_09[which(cmip6smonthly$ET0_09_mean < 400/360)] <- "Cold"

cmip6smonthly$cat.AI_09.ref <- cmip6smonthly$AI_09_mean.ref %>% cut(breaks = breaks.unesco) %>% revalue(cat.unesco)
cmip6smonthly$cat.AI_09.ref[which(cmip6smonthly$ET0_09_mean.ref < 400/360)] <- "Cold"

# CATEGORIES 10
cmip6smonthly$cat.AI_10 <- cmip6smonthly$AI_10_mean %>% cut(breaks = breaks.unesco) %>% revalue(cat.unesco)
cmip6smonthly$cat.AI_10[which(cmip6smonthly$ET0_10_mean < 400/360)] <- "Cold"

cmip6smonthly$cat.AI_10.ref <- cmip6smonthly$AI_10_mean.ref %>% cut(breaks = breaks.unesco) %>% revalue(cat.unesco)
cmip6smonthly$cat.AI_10.ref[which(cmip6smonthly$ET0_10_mean.ref < 400/360)] <- "Cold"

# CATEGORIES 11
cmip6smonthly$cat.AI_11 <- cmip6smonthly$AI_11_mean %>% cut(breaks = breaks.unesco) %>% revalue(cat.unesco)
cmip6smonthly$cat.AI_11[which(cmip6smonthly$ET0_11_mean < 400/360)] <- "Cold"

cmip6smonthly$cat.AI_11.ref <- cmip6smonthly$AI_11_mean.ref %>% cut(breaks = breaks.unesco) %>% revalue(cat.unesco)
cmip6smonthly$cat.AI_11.ref[which(cmip6smonthly$ET0_11_mean.ref < 400/360)] <- "Cold"

# CATEGORIES 12
cmip6smonthly$cat.AI_12 <- cmip6smonthly$AI_12_mean %>% cut(breaks = breaks.unesco) %>% revalue(cat.unesco)
cmip6smonthly$cat.AI_12[which(cmip6smonthly$ET0_12_mean < 400/360)] <- "Cold"

cmip6smonthly$cat.AI_12.ref <- cmip6smonthly$AI_12_mean.ref %>% cut(breaks = breaks.unesco) %>% revalue(cat.unesco)
cmip6smonthly$cat.AI_12.ref[which(cmip6smonthly$ET0_12_mean.ref < 400/360)] <- "Cold"

write.table(cmip6smonthly, "cmip6smonthly.txt")
```

```{r map-AI-monthly, fig.dim = c(12,16)}
cmip6smonthly <- read.table("cmip6smonthly.txt")

cmip6smonthly$cat.AI_01 <- factor(cmip6smonthly$cat.AI_01, levels = c("Hyperarid","Arid","Semi-arid","Dry subhumid","Humid","Cold"))

ggarrange(
ggplot() + 
    geom_tile(data = subset(cmip6smonthly, period == "1970_2000" & model == "historical"), aes(x=lon, y = lat,  fill = cat.AI_01))+
    borders(colour = "grey60")+
    scale_fill_manual(values =  col.cat, na.translate = F)+
    labs(title = "January", fill = "")+
    theme_void()+ylim(-55,90)+
    theme(legend.position = "bottom"),

ggplot() + 
    geom_tile(data = subset(cmip6smonthly, period == "1970_2000" & model == "historical"), aes(x=lon, y = lat,  fill = cat.AI_02))+
    borders(colour = "grey60")+
    scale_fill_manual(values =  col.cat, na.translate = F)+
    labs(title = "February", fill = "")+
    theme_void()+ylim(-55,90)+
    theme(legend.position = "bottom"),

ggplot() + 
    geom_tile(data = subset(cmip6smonthly, period == "1970_2000" & model == "historical"), aes(x=lon, y = lat,  fill = cat.AI_03))+
    borders(colour = "grey60")+
    scale_fill_manual(values =  col.cat, na.translate = F)+
    labs(title = "March", fill = "")+
    theme_void()+ylim(-55,90)+
    theme(legend.position = "bottom"),

ggplot() + 
    geom_tile(data = subset(cmip6smonthly, period == "1970_2000" & model == "historical"), aes(x=lon, y = lat,  fill = cat.AI_04))+
    borders(colour = "grey60")+
    scale_fill_manual(values =  col.cat, na.translate = F)+
    labs(title = "April", fill = "")+
    theme_void()+ylim(-55,90)+
    theme(legend.position = "bottom"),

ggplot() + 
    geom_tile(data = subset(cmip6smonthly, period == "1970_2000" & model == "historical"), aes(x=lon, y = lat,  fill = cat.AI_05))+
    borders(colour = "grey60")+
    scale_fill_manual(values =  col.cat, na.translate = F)+
    labs(title = "May", fill = "")+
    theme_void()+ylim(-55,90)+
    theme(legend.position = "bottom"),

ggplot() + 
    geom_tile(data = subset(cmip6smonthly, period == "1970_2000" & model == "historical"), aes(x=lon, y = lat,  fill = cat.AI_06))+
    borders(colour = "grey60")+
    scale_fill_manual(values =  col.cat, na.translate = F)+
    labs(title = "June", fill = "")+
    theme_void()+ylim(-55,90)+
    theme(legend.position = "bottom"),

ggplot() + 
    geom_tile(data = subset(cmip6smonthly, period == "1970_2000" & model == "historical"), aes(x=lon, y = lat,  fill = cat.AI_07))+
    borders(colour = "grey60")+
    scale_fill_manual(values =  col.cat, na.translate = F)+
    labs(title = "July", fill = "")+
    theme_void()+ylim(-55,90)+
    theme(legend.position = "bottom"),

ggplot() + 
    geom_tile(data = subset(cmip6smonthly, period == "1970_2000" & model == "historical"), aes(x=lon, y = lat,  fill = cat.AI_08))+
    borders(colour = "grey60")+
    scale_fill_manual(values =  col.cat, na.translate = F)+
    labs(title = "August", fill = "")+
    theme_void()+ylim(-55,90)+
    theme(legend.position = "bottom"),

ggplot() + 
    geom_tile(data = subset(cmip6smonthly, period == "1970_2000" & model == "historical"), aes(x=lon, y = lat,  fill = cat.AI_09))+
    borders(colour = "grey60")+
    scale_fill_manual(values =  col.cat, na.translate = F)+
    labs(title = "September", fill = "")+
    theme_void()+ylim(-55,90)+
    theme(legend.position = "bottom"),

ggplot() + 
    geom_tile(data = subset(cmip6smonthly, period == "1970_2000" & model == "historical"), aes(x=lon, y = lat,  fill = cat.AI_10))+
    borders(colour = "grey60")+
    scale_fill_manual(values =  col.cat, na.translate = F)+
    labs(title = "October", fill = "")+
    theme_void()+ylim(-55,90)+
    theme(legend.position = "bottom"),

ggplot() + 
    geom_tile(data = subset(cmip6smonthly, period == "1970_2000" & model == "historical"), aes(x=lon, y = lat,  fill = cat.AI_11))+
    borders(colour = "grey60")+
    scale_fill_manual(values =  col.cat, na.translate = F)+
    labs(title = "November", fill = "")+
    theme_void()+ylim(-55,90)+
    theme(legend.position = "bottom"),

ggplot() + 
    geom_tile(data = subset(cmip6smonthly, period == "1970_2000" & model == "historical"), aes(x=lon, y = lat,  fill = cat.AI_12))+
    borders(colour = "grey60")+
    scale_fill_manual(values =  col.cat, na.translate = F)+
    labs(title = "December", fill = "")+
    theme_void()+ylim(-55,90)+
    theme(legend.position = "bottom"),

ncol = 3, nrow = 4, common.legend = T, legend = "bottom"
)

```

## Evolution of monthly aridity by continent 

```{r AI-evol-by-continent}
cmip6smm <- cmip6smonthly %>% group_by(model, period, Continent) %>%
  summarise(AI_01 = median(AI_01_mean, na.rm = T), cov_AI_01 = sd(AI_01_mean, na.rm = T)/mean(AI_01_mean, na.rm = T),
            AI_02 = median(AI_02_mean, na.rm = T), cov_AI_02 = sd(AI_02_mean, na.rm = T)/mean(AI_02_mean, na.rm = T),
            AI_03 = median(AI_03_mean, na.rm = T), cov_AI_03 = sd(AI_03_mean, na.rm = T)/mean(AI_03_mean, na.rm = T),
            AI_04 = median(AI_04_mean, na.rm = T), cov_AI_04 = sd(AI_04_mean, na.rm = T)/mean(AI_04_mean, na.rm = T),
            AI_05 = median(AI_05_mean, na.rm = T), cov_AI_05 = sd(AI_05_mean, na.rm = T)/mean(AI_05_mean, na.rm = T),
            AI_06 = median(AI_06_mean, na.rm = T), cov_AI_06 = sd(AI_06_mean, na.rm = T)/mean(AI_06_mean, na.rm = T),
            AI_07 = median(AI_07_mean, na.rm = T), cov_AI_07 = sd(AI_07_mean, na.rm = T)/mean(AI_07_mean, na.rm = T),
            AI_08 = median(AI_08_mean, na.rm = T), cov_AI_08 = sd(AI_08_mean, na.rm = T)/mean(AI_08_mean, na.rm = T),
            AI_09 = median(AI_09_mean, na.rm = T), cov_AI_09 = sd(AI_09_mean, na.rm = T)/mean(AI_09_mean, na.rm = T),
            AI_10 = median(AI_10_mean, na.rm = T), cov_AI_10 = sd(AI_10_mean, na.rm = T)/mean(AI_10_mean, na.rm = T),
            AI_11 = median(AI_11_mean, na.rm = T), cov_AI_11 = sd(AI_11_mean, na.rm = T)/mean(AI_11_mean, na.rm = T),
            AI_12 = median(AI_12_mean, na.rm = T), cov_AI_12 = sd(AI_12_mean, na.rm = T)/mean(AI_12_mean, na.rm = T)) %>%
 ungroup() %>% as.data.frame()

# cmip6smm <- cmip6smonthly %>% group_by(model, period, Continent) %>% 
#   summarise(AI_01 = mean(AI_01_mean, na.rm = T),
#             AI_02 = mean(AI_02_mean, na.rm = T),
#             AI_03 = mean(AI_03_mean, na.rm = T),
#             AI_04 = mean(AI_04_mean, na.rm = T),
#             AI_05 = mean(AI_05_mean, na.rm = T),
#             AI_06 = mean(AI_06_mean, na.rm = T),
#             AI_07 = mean(AI_07_mean, na.rm = T),
#             AI_08 = mean(AI_08_mean, na.rm = T),
#             AI_09 = mean(AI_09_mean, na.rm = T),
#             AI_10 = mean(AI_10_mean, na.rm = T),
#             AI_11 = mean(AI_11_mean, na.rm = T),
#             AI_12 = mean(AI_12_mean, na.rm = T)) %>%
#  ungroup() %>% as.data.frame()

long_AI <- cmip6smm %>% select(c("model","period","Continent","AI_01","AI_02","AI_03","AI_04","AI_05","AI_06","AI_07","AI_08","AI_09","AI_10","AI_11","AI_12")) %>%
                                 tidyr::pivot_longer(
                                   cols = starts_with("AI_"),
                                   names_to= "Month",
                                   values_to = "AI")

long_AI$Month <- revalue(long_AI$Month, c("AI_01" = "January","AI_02" = "February", "AI_03" = "March", "AI_04" = "April","AI_05" ="May","AI_06" = "June", "AI_07" ="July","AI_08" = "August","AI_09" = "September","AI_10" = "October","AI_11" = "November","AI_12" = "December"))
                               
long_sdAI <- cmip6smm %>% 
  select(c("model","period","Continent","cov_AI_01","cov_AI_02","cov_AI_03","cov_AI_04","cov_AI_05","cov_AI_06","cov_AI_07","cov_AI_08","cov_AI_09","cov_AI_10","cov_AI_11","cov_AI_12")) %>%
  tidyr::pivot_longer(
  cols = starts_with("cov_AI_"),
  names_to= "Month",
  values_to = "cov_AI"
)

long_sdAI$Month <- revalue(long_sdAI$Month, c("cov_AI_01" = "January","cov_AI_02" = "February", "cov_AI_03" = "March", "cov_AI_04" = "April","cov_AI_05" ="May","cov_AI_06" = "June", "cov_AI_07" ="July","cov_AI_08" = "August","cov_AI_09" = "September","cov_AI_10" = "October","cov_AI_11" = "November","cov_AI_12" = "December"))

cmip6sml <- merge(long_AI,long_sdAI, by = c("model","period","Continent","Month"))

cmip6sml$Month <- factor(cmip6sml$Month, c("January","February","March","April","May","June","July","August","September","October","November","December"))

col.model <- c("historical" = "#4575B4", "SSP245" = "#FEE08B", "SSP370" = "#FDAE61", "SSP585" = "#D73027")

cmip6sml$NMonth <- cmip6sml$Month %>% revalue(c("January" = 1, "February" =2, "March"= 3, "April"= 4,"May" = 5, "June" = 6,"July" = 7, "August"= 8,"September" = 9, "October" = 10,"November" = 11, "December" = 12))



```

```{r plot-AI-month-cont}

for(i in  c("NORTH-AMERICA","CENTRAL-AMERICA","SOUTH-AMERICA","EUROPE","EUROPE-AFRICA","AFRICA","ASIA","OCEANIA")){
g <- ggplot(subset(cmip6sml, Continent == i & period %in% c("1970_2000","2070_2100")))+
  # geom_rect(aes(xmin = 1, xmax = 12, ymin= 0, ymax = 0.03), fill = "#D73027", alpha = 0.02)+
  # geom_rect(aes(xmin = 1, xmax = 12, ymin= 0.03, ymax = 0.2), fill = "#FDAE61", alpha = 0.02)+
  # geom_rect(aes(xmin = 1, xmax = 12, ymin= 0.2, ymax = 0.5), fill = "#FEE08B", alpha = 0.02)+
  # geom_rect(aes(xmin = 1, xmax = 12, ymin= 0.5, ymax = 0.65), fill = "#ABD9E9", alpha = 0.02)+
  # geom_rect(aes(xmin = 1, xmax = 12, ymin= 0.65, ymax = Inf), fill = "#4575B4", alpha = 0.02)+
  geom_point(aes(x=Month, y = AI, size = cov_AI, col = model))+
  geom_line(aes(x=Month, y = AI, col = model, group = model, linetype = model))+
  scale_color_manual(values = col.model)+
  scale_size_continuous(transform = "pseudo_log")+
  labs(y = "Median aridity index", size = "Coefficient of variance", col = "Model", linetype = "Model", title = i)+
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))
print(g)
}
```

## Evolution of monthly evapotranspiration by continent 

```{r ET0-evol-by-continent}
cmip6smm_et0 <- cmip6smonthly %>% group_by(model, period, Continent) %>%
  summarise(ET0_01 = median(ET0_01_mean, na.rm = T), cov_ET0_01 = sd(ET0_01_mean, na.rm = T)/mean(ET0_01_mean, na.rm = T),
            ET0_02 = median(ET0_02_mean, na.rm = T), cov_ET0_02 = sd(ET0_02_mean, na.rm = T)/mean(ET0_02_mean, na.rm = T),
            ET0_03 = median(ET0_03_mean, na.rm = T), cov_ET0_03 = sd(ET0_03_mean, na.rm = T)/mean(ET0_03_mean, na.rm = T),
            ET0_04 = median(ET0_04_mean, na.rm = T), cov_ET0_04 = sd(ET0_04_mean, na.rm = T)/mean(ET0_04_mean, na.rm = T),
            ET0_05 = median(ET0_05_mean, na.rm = T), cov_ET0_05 = sd(ET0_05_mean, na.rm = T)/mean(ET0_05_mean, na.rm = T),
            ET0_06 = median(ET0_06_mean, na.rm = T), cov_ET0_06 = sd(ET0_06_mean, na.rm = T)/mean(ET0_06_mean, na.rm = T),
            ET0_07 = median(ET0_07_mean, na.rm = T), cov_ET0_07 = sd(ET0_07_mean, na.rm = T)/mean(ET0_07_mean, na.rm = T),
            ET0_08 = median(ET0_08_mean, na.rm = T), cov_ET0_08 = sd(ET0_08_mean, na.rm = T)/mean(ET0_08_mean, na.rm = T),
            ET0_09 = median(ET0_09_mean, na.rm = T), cov_ET0_09 = sd(ET0_09_mean, na.rm = T)/mean(ET0_09_mean, na.rm = T),
            ET0_10 = median(ET0_10_mean, na.rm = T), cov_ET0_10 = sd(ET0_10_mean, na.rm = T)/mean(ET0_10_mean, na.rm = T),
            ET0_11 = median(ET0_11_mean, na.rm = T), cov_ET0_11 = sd(ET0_11_mean, na.rm = T)/mean(ET0_11_mean, na.rm = T),
            ET0_12 = median(ET0_12_mean, na.rm = T), cov_ET0_12 = sd(ET0_12_mean, na.rm = T)/mean(ET0_12_mean, na.rm = T)) %>%
 ungroup() %>% as.data.frame()

# cmip6smm <- cmip6smonthly %>% group_by(model, period, Continent) %>% 
#   summarise(ET0_01 = mean(ET0_01_mean, na.rm = T),
#             ET0_02 = mean(ET0_02_mean, na.rm = T),
#             ET0_03 = mean(ET0_03_mean, na.rm = T),
#             ET0_04 = mean(ET0_04_mean, na.rm = T),
#             ET0_05 = mean(ET0_05_mean, na.rm = T),
#             ET0_06 = mean(ET0_06_mean, na.rm = T),
#             ET0_07 = mean(ET0_07_mean, na.rm = T),
#             ET0_08 = mean(ET0_08_mean, na.rm = T),
#             ET0_09 = mean(ET0_09_mean, na.rm = T),
#             ET0_10 = mean(ET0_10_mean, na.rm = T),
#             ET0_11 = mean(ET0_11_mean, na.rm = T),
#             ET0_12 = mean(ET0_12_mean, na.rm = T)) %>%
#  ungroup() %>% as.data.frame()

long_ET0 <- cmip6smm_et0 %>% select(c("model","period","Continent","ET0_01","ET0_02","ET0_03","ET0_04","ET0_05","ET0_06","ET0_07","ET0_08","ET0_09","ET0_10","ET0_11","ET0_12")) %>%
                                 tidyr::pivot_longer(
                                   cols = starts_with("ET0_"),
                                   names_to= "Month",
                                   values_to = "AI")

long_ET0$Month <- revalue(long_AI$Month, c("ET0_01" = "January","ET0_02" = "February", "ET0_03" = "March", "ET0_04" = "April","ET0_05" ="May","ET0_06" = "June", "ET0_07" ="July","ET0_08" = "August","ET0_09" = "September","ET0_10" = "October","ET0_11" = "November","ET0_12" = "December"))
                               
long_sdET0 <- cmip6smm_et0 %>% 
  select(c("model","period","Continent","cov_ET0_01","cov_ET0_02","cov_ET0_03","cov_ET0_04","cov_ET0_05","cov_ET0_06","cov_ET0_07","cov_ET0_08","cov_ET0_09","cov_ET0_10","cov_ET0_11","cov_ET0_12")) %>%
  tidyr::pivot_longer(
  cols = starts_with("cov_ET0_"),
  names_to= "Month",
  values_to = "cov_AI"
)

long_sdET0$Month <- revalue(long_sdET0$Month, c("cov_ET0_01" = "January","cov_ET0_02" = "February", "cov_ET0_03" = "March", "cov_ET0_04" = "April","cov_ET0_05" ="May","cov_ET0_06" = "June", "cov_ET0_07" ="July","cov_ET0_08" = "August","cov_ET0_09" = "September","cov_ET0_10" = "October","cov_ET0_11" = "November","cov_ET0_12" = "December"))

cmip6sml_et0 <- merge(long_ET0,long_sdET0, by = c("model","period","Continent","Month"))

cmip6sml_et0$Month <- factor(cmip6sml_et0$Month, c("January","February","March","April","May","June","July","August","September","October","November","December"))

col.model <- c("historical" = "#4575B4", "SSP245" = "#FEE08B", "SSP370" = "#FDAE61", "SSP585" = "#D73027")

cmip6sml_et0$NMonth <- cmip6sml_et0$Month %>% revalue(c("January" = 1, "February" =2, "March"= 3, "April"= 4,"May" = 5, "June" = 6,"July" = 7, "August"= 8,"September" = 9, "October" = 10,"November" = 11, "December" = 12))



```

```{r plot-ET0-month-cont}

for(i in  c("NORTH-AMERICA","CENTRAL-AMERICA","SOUTH-AMERICA","EUROPE","EUROPE-AFRICA","AFRICA","ASIA","OCEANIA")){
g <- ggplot(subset(cmip6sml_et0, Continent == i & period %in% c("1970_2000","2070_2100")))+
  # geom_rect(aes(xmin = 1, xmax = 12, ymin= 0, ymax = 0.03), fill = "#D73027", alpha = 0.02)+
  # geom_rect(aes(xmin = 1, xmax = 12, ymin= 0.03, ymax = 0.2), fill = "#FDAE61", alpha = 0.02)+
  # geom_rect(aes(xmin = 1, xmax = 12, ymin= 0.2, ymax = 0.5), fill = "#FEE08B", alpha = 0.02)+
  # geom_rect(aes(xmin = 1, xmax = 12, ymin= 0.5, ymax = 0.65), fill = "#ABD9E9", alpha = 0.02)+
  # geom_rect(aes(xmin = 1, xmax = 12, ymin= 0.65, ymax = Inf), fill = "#4575B4", alpha = 0.02)+
  geom_point(aes(x=Month, y = AI, size = cov_AI, col = model))+
  geom_line(aes(x=Month, y = AI, col = model, group = model, linetype = model))+
  scale_color_manual(values = col.model)+
  scale_size_continuous(transform = "pseudo_log")+
  labs(y = "Median evapotranspiration", size = "Coefficient of variance", col = "Model", linetype = "Model", title = i)+
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))
print(g)
}
```

## Evolution of monthly precipitation by continent 

```{r pr-evol-by-continent}
cmip6smm_pr <- cmip6smonthly %>% group_by(model, period, Continent) %>%
  summarise(pr01 = median(pr_01_mean, na.rm = T), cov_pr_01 = sd(pr_01_mean, na.rm = T)/mean(pr_01_mean, na.rm = T),
            pr_02 = median(pr_02_mean, na.rm = T), cov_pr_02 = sd(pr_02_mean, na.rm = T)/mean(pr_02_mean, na.rm = T),
            pr_03 = median(pr_03_mean, na.rm = T), cov_pr_03 = sd(pr_03_mean, na.rm = T)/mean(pr_03_mean, na.rm = T),
            pr_04 = median(pr_04_mean, na.rm = T), cov_pr_04 = sd(pr_04_mean, na.rm = T)/mean(pr_04_mean, na.rm = T),
            pr_05 = median(pr_05_mean, na.rm = T), cov_pr_05 = sd(pr_05_mean, na.rm = T)/mean(pr_05_mean, na.rm = T),
            pr_06 = median(pr_06_mean, na.rm = T), cov_pr_06 = sd(pr_06_mean, na.rm = T)/mean(pr_06_mean, na.rm = T),
            pr_07 = median(pr_07_mean, na.rm = T), cov_pr_07 = sd(pr_07_mean, na.rm = T)/mean(pr_07_mean, na.rm = T),
            pr_08 = median(pr_08_mean, na.rm = T), cov_pr_08 = sd(pr_08_mean, na.rm = T)/mean(pr_08_mean, na.rm = T),
            pr_09 = median(pr_09_mean, na.rm = T), cov_pr_09 = sd(pr_09_mean, na.rm = T)/mean(pr_09_mean, na.rm = T),
            pr_10 = median(pr_10_mean, na.rm = T), cov_pr_10 = sd(pr_10_mean, na.rm = T)/mean(pr_10_mean, na.rm = T),
            pr_11 = median(pr_11_mean, na.rm = T), cov_pr_11 = sd(pr_11_mean, na.rm = T)/mean(pr_11_mean, na.rm = T),
            pr_12 = median(pr_12_mean, na.rm = T), cov_pr_12 = sd(pr_12_mean, na.rm = T)/mean(pr_12_mean, na.rm = T)) %>%
 ungroup() %>% as.data.frame()

# cmip6smm <- cmip6smonthly %>% group_by(model, period, Continent) %>% 
#   summarise(pr_01 = mean(pr_01_mean, na.rm = T),
#             pr_02 = mean(pr_02_mean, na.rm = T),
#             pr_03 = mean(pr_03_mean, na.rm = T),
#             pr_04 = mean(pr_04_mean, na.rm = T),
#             pr_05 = mean(pr_05_mean, na.rm = T),
#             pr_06 = mean(pr_06_mean, na.rm = T),
#             pr_07 = mean(pr_07_mean, na.rm = T),
#             pr_08 = mean(pr_08_mean, na.rm = T),
#             pr_09 = mean(pr_09_mean, na.rm = T),
#             pr_10 = mean(pr_10_mean, na.rm = T),
#             pr_11 = mean(pr_11_mean, na.rm = T),
#             pr_12 = mean(pr_12_mean, na.rm = T)) %>%
#  ungroup() %>% as.data.frame()

long_ET0 <- cmip6smm_et0 %>% select(c("model","period","Continent","pr_01","pr_02","pr_03","pr_04","pr_05","pr_06","pr_07","pr_08","pr_09","pr_10","pr_11","pr_12")) %>%
                                 tidyr::pivot_longer(
                                   cols = starts_with("pr_"),
                                   names_to= "Month",
                                   values_to = "AI")

long_pr$Month <- revalue(long_AI$Month, c("pr_01" = "January","pr_02" = "February", "pr_03" = "March", "pr_04" = "Apr_il","pr_05" ="May","pr_06" = "June", "pr_07" ="July","pr_08" = "August","pr_09" = "September","pr_10" = "October","pr_11" = "November","pr_12" = "December"))
                               
long_sdpr <- cmip6smm_pr %>% 
  select(c("model","period","Continent","cov_pr_01","cov_pr_02","cov_pr_03","cov_pr_04","cov_pr_05","cov_pr_06","cov_pr_07","cov_pr_08","cov_pr_09","cov_pr_10","cov_pr_11","cov_pr_12")) %>%
  tidyr::pivot_longer(
  cols = starts_with("cov_ET0_"),
  names_to= "Month",
  values_to = "cov_AI"
)

long_sdpr$Month <- revalue(long_pr$Month, c("cov_pr_01" = "January","cov_pr_02" = "February", "cov_pr_03" = "March", "cov_pr_04" = "April","cov_pr_05" ="May","cov_pr_06" = "June", "cov_pr_07" ="July","cov_pr_08" = "August","cov_pr_09" = "September","cov_pr_10" = "October","cov_pr_11" = "November","cov_pr_12" = "December"))

cmip6sml_pr <- merge(long_pr,long_sdpr, by = c("model","period","Continent","Month"))

cmip6sml_pr$Month <- factor(cmip6sml_pr$Month, c("January","February","March","April","May","June","July","August","September","October","November","December"))

col.model <- c("historical" = "#4575B4", "SSP245" = "#FEE08B", "SSP370" = "#FDAE61", "SSP585" = "#D73027")

cmip6sml_pr$NMonth <- cmip6sml_pr$Month %>% revalue(c("January" = 1, "February" =2, "March"= 3, "April"= 4,"May" = 5, "June" = 6,"July" = 7, "August"= 8,"September" = 9, "October" = 10,"November" = 11, "December" = 12))



```

```{r plot-pr-month-cont}

for(i in  c("NORTH-AMERICA","CENTRAL-AMERICA","SOUTH-AMERICA","EUROPE","EUROPE-AFRICA","AFRICA","ASIA","OCEANIA")){
g <- ggplot(subset(cmip6sml_pr, Continent == i & period %in% c("1970_2000","2070_2100")))+
  # geom_rect(aes(xmin = 1, xmax = 12, ymin= 0, ymax = 0.03), fill = "#D73027", alpha = 0.02)+
  # geom_rect(aes(xmin = 1, xmax = 12, ymin= 0.03, ymax = 0.2), fill = "#FDAE61", alpha = 0.02)+
  # geom_rect(aes(xmin = 1, xmax = 12, ymin= 0.2, ymax = 0.5), fill = "#FEE08B", alpha = 0.02)+
  # geom_rect(aes(xmin = 1, xmax = 12, ymin= 0.5, ymax = 0.65), fill = "#ABD9E9", alpha = 0.02)+
  # geom_rect(aes(xmin = 1, xmax = 12, ymin= 0.65, ymax = Inf), fill = "#4575B4", alpha = 0.02)+
  geom_point(aes(x=Month, y = AI, size = cov_AI, col = model))+
  geom_line(aes(x=Month, y = AI, col = model, group = model, linetype = model))+
  scale_color_manual(values = col.model)+
  scale_size_continuous(transform = "pseudo_log")+
  labs(y = "Median monthly precipitation, mm", size = "Coefficient of variance", col = "Model", linetype = "Model", title = i)+
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))
print(g)
}
```