---
title: "CMIP6 prep"
author: "Camille"
date: "2024-02-19"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: hide
---
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = F, warning = F, error = F, fig.align = "center",
                      collapse = T, cache = T)
```

```{r libraries}
library(raster)

library(plyr)
library(dplyr)

library(ggplot2)
library(viridisLite)
library(colorspace)
library(RColorBrewer)

library(sf)
library(stringr)

library(ggpubr)
library(kableExtra)

memory.limit(size = 160000)

#display.brewer.pal(n = 11, name ="RdYlGn")
colorvec <- brewer.pal(n = 11, name ="RdYlGn")

```

# Data

```{r land-masks}
land_mask <- raster("Masks/land_sea_mask_1degree.nc4") 
land_mask.df <- as.data.frame(land_mask, xy = T) %>% setNames(c("lon","lat","lm")) # 0 if ocean, 1 if land


elev <- raster("Worldclim/wc2.1_10m/wc2.1_10m_elev.tif") 
elev.df <- elev %>% projectRaster(to = land_mask) %>% as.data.frame(xy = T) %>% setNames(c("lon","lat", "z"))


ipcc_regions <- shapefile("Masks/IPCC-WGI-reference-regions-v4.shp") %>% spTransform(crs("EPSG:4326")) 
ipcc_regions.raster <- ipcc_regions %>% rasterize(land_mask)
ipcc_regions.df <- as.data.frame(ipcc_regions.raster, xy = T) %>% setNames(c("lon","lat","Continent","Type","Name","Acronym"))

```

```{r cmip6-data, eval = F}
tas.all_annual <- read.table("CMIP6/tas/tas.all_annual.txt")
pr.all_annual <- read.table("CMIP6/pr/pr.all_annual.txt")
sfcWind.all_annual <- read.table("CMIP6/sfcWind/sfcWind.all_annual.txt")
hfls.all_annual <- read.table("CMIP6/hfls/hfls.all_annual.txt")
hfss.all_annual <- read.table("CMIP6/hfss/hfss.all_annual.txt")
hurs.all_annual <- read.table("CMIP6/hurs/hurs.all_annual.txt")

cmip6_annual <- tas.all_annual %>% merge(pr.all_annual, by = c("lon","lat","model","period","lm", "Continent","Type","Name","Acronym","source")) %>%
  merge(sfcWind.all_annual, by = c("lon","lat","model","period","lm", "Continent","Type","Name","Acronym","source")) %>%
  merge(hfls.all_annual, by = c("lon","lat","model","period","lm", "Continent","Type","Name","Acronym","source")) %>%
  merge(hfss.all_annual, by = c("lon","lat","model","period","lm", "Continent","Type","Name","Acronym","source")) %>%
  merge(hurs.all_annual, by = c("lon","lat","model","period","lm", "Continent","Type","Name","Acronym","source")) %>% 
  merge(elev.df, by = c("lon", "lat"))

write.table(cmip6_annual, "cmip6_annual.txt")
```

```{r cmip6-january, eval = F}
tas.all_january <- read.table("CMIP6/tas/tas.all_january.txt")
pr.all_january <- read.table("CMIP6/pr/pr.all_january.txt")
sfcWind.all_january <- read.table("CMIP6/sfcWind/sfcWind.all_january.txt")
hfls.all_january <- read.table("CMIP6/hfls/hfls.all_january.txt")
hfss.all_january <- read.table("CMIP6/hfss/hfss.all_january.txt")
hurs.all_january <- read.table("CMIP6/hurs/hurs.all_january.txt")

cmip6_january <- tas.all_january %>% merge(pr.all_january, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>%
  merge(sfcWind.all_january, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>%
  merge(hfls.all_january, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>%
  merge(hfss.all_january, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>%
  merge(hurs.all_january, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>% 
  merge(elev.df, by = c("lon", "lat"))

write.table(cmip6_january, "cmip6_january.txt")
```

```{r cmip6-february, eval = F}
tas.all_february <- read.table("CMIP6/tas/tas.all_february.txt")
pr.all_february <- read.table("CMIP6/pr/pr.all_february.txt")
sfcWind.all_february <- read.table("CMIP6/sfcWind/sfcWind.all_february.txt")
hfls.all_february <- read.table("CMIP6/hfls/hfls.all_february.txt")
hfss.all_february <- read.table("CMIP6/hfss/hfss.all_february.txt")
hurs.all_february <- read.table("CMIP6/hurs/hurs.all_february.txt")

cmip6_february <- tas.all_february %>% merge(pr.all_february, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>%
  merge(sfcWind.all_february, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>%
  merge(hfls.all_february, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>%
  merge(hfss.all_february, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>%
  merge(hurs.all_february, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>% 
  merge(elev.df, by = c("lon", "lat"))

write.table(cmip6_february, "cmip6_february.txt")
```

```{r cmip6-march, eval = F}
tas.all_march <- read.table("CMIP6/tas/tas.all_march.txt")
pr.all_march <- read.table("CMIP6/pr/pr.all_march.txt")
sfcWind.all_march <- read.table("CMIP6/sfcWind/sfcWind.all_march.txt")
hfls.all_march <- read.table("CMIP6/hfls/hfls.all_march.txt")
hfss.all_march <- read.table("CMIP6/hfss/hfss.all_march.txt")
hurs.all_march <- read.table("CMIP6/hurs/hurs.all_march.txt")

cmip6_march <- tas.all_march %>% merge(pr.all_march, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>%
  merge(sfcWind.all_march, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>%
  merge(hfls.all_march, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>%
  merge(hfss.all_march, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>%
  merge(hurs.all_march, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>% 
  merge(elev.df, by = c("lon", "lat"))

write.table(cmip6_march, "cmip6_march.txt")
```

```{r cmip6-april, eval = F}
tas.all_april <- read.table("CMIP6/tas/tas.all_april.txt")
pr.all_april <- read.table("CMIP6/pr/pr.all_april.txt")
sfcWind.all_april <- read.table("CMIP6/sfcWind/sfcWind.all_april.txt")
hfls.all_april <- read.table("CMIP6/hfls/hfls.all_april.txt")
hfss.all_april <- read.table("CMIP6/hfss/hfss.all_april.txt")
hurs.all_april <- read.table("CMIP6/hurs/hurs.all_april.txt")

cmip6_april <- tas.all_april %>% merge(pr.all_april, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>%
  merge(sfcWind.all_april, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>%
  merge(hfls.all_april, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>%
  merge(hfss.all_april, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>%
  merge(hurs.all_april, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>% 
  merge(elev.df, by = c("lon", "lat"))

write.table(cmip6_april, "cmip6_april.txt")
```

```{r cmip6-may, eval = F}
tas.all_may <- read.table("CMIP6/tas/tas.all_may.txt")
pr.all_may <- read.table("CMIP6/pr/pr.all_may.txt")
sfcWind.all_may <- read.table("CMIP6/sfcWind/sfcWind.all_may.txt")
hfls.all_may <- read.table("CMIP6/hfls/hfls.all_may.txt")
hfss.all_may <- read.table("CMIP6/hfss/hfss.all_may.txt")
hurs.all_may <- read.table("CMIP6/hurs/hurs.all_may.txt")

cmip6_may <- tas.all_may %>% merge(pr.all_may, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>%
  merge(sfcWind.all_may, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>%
  merge(hfls.all_may, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>%
  merge(hfss.all_may, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>%
  merge(hurs.all_may, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>% 
  merge(elev.df, by = c("lon", "lat"))

write.table(cmip6_may, "cmip6_may.txt")
```

```{r cmip6-june, eval = F}
tas.all_june <- read.table("CMIP6/tas/tas.all_june.txt")
pr.all_june <- read.table("CMIP6/pr/pr.all_june.txt")
sfcWind.all_june <- read.table("CMIP6/sfcWind/sfcWind.all_june.txt")
hfls.all_june <- read.table("CMIP6/hfls/hfls.all_june.txt")
hfss.all_june <- read.table("CMIP6/hfss/hfss.all_june.txt")
hurs.all_june <- read.table("CMIP6/hurs/hurs.all_june.txt")

cmip6_june <- tas.all_june %>% merge(pr.all_june, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>%
  merge(sfcWind.all_june, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>%
  merge(hfls.all_june, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>%
  merge(hfss.all_june, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>%
  merge(hurs.all_june, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>% 
  merge(elev.df, by = c("lon", "lat"))

write.table(cmip6_june, "cmip6_june.txt")
```

```{r cmip6-july, eval = F}
tas.all_july <- read.table("CMIP6/tas/tas.all_july.txt")
pr.all_july <- read.table("CMIP6/pr/pr.all_july.txt")
sfcWind.all_july <- read.table("CMIP6/sfcWind/sfcWind.all_july.txt")
hfls.all_july <- read.table("CMIP6/hfls/hfls.all_july.txt")
hfss.all_july <- read.table("CMIP6/hfss/hfss.all_july.txt")
hurs.all_july <- read.table("CMIP6/hurs/hurs.all_july.txt")

cmip6_july <- tas.all_july %>% merge(pr.all_july, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>%
  merge(sfcWind.all_july, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>%
  merge(hfls.all_july, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>%
  merge(hfss.all_july, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>%
  merge(hurs.all_july, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>% 
  merge(elev.df, by = c("lon", "lat"))

write.table(cmip6_july, "cmip6_july.txt")
```

```{r cmip6-august, eval = F}
tas.all_august <- read.table("CMIP6/tas/tas.all_august.txt")
pr.all_august <- read.table("CMIP6/pr/pr.all_august.txt")
sfcWind.all_august <- read.table("CMIP6/sfcWind/sfcWind.all_august.txt")
hfls.all_august <- read.table("CMIP6/hfls/hfls.all_august.txt")
hfss.all_august <- read.table("CMIP6/hfss/hfss.all_august.txt")
hurs.all_august <- read.table("CMIP6/hurs/hurs.all_august.txt")

cmip6_august <- tas.all_august %>% merge(pr.all_august, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>%
  merge(sfcWind.all_august, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>%
  merge(hfls.all_august, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>%
  merge(hfss.all_august, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>%
  merge(hurs.all_august, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>% 
  merge(elev.df, by = c("lon", "lat"))

write.table(cmip6_august, "cmip6_august.txt")
```

```{r cmip6-september, eval = F}
tas.all_september <- read.table("CMIP6/tas/tas.all_september.txt")
pr.all_september <- read.table("CMIP6/pr/pr.all_september.txt")
sfcWind.all_september <- read.table("CMIP6/sfcWind/sfcWind.all_september.txt")
hfls.all_september <- read.table("CMIP6/hfls/hfls.all_september.txt")
hfss.all_september <- read.table("CMIP6/hfss/hfss.all_september.txt")
hurs.all_september <- read.table("CMIP6/hurs/hurs.all_september.txt")

cmip6_september <- tas.all_september %>% merge(pr.all_september, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>%
  merge(sfcWind.all_september, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>%
  merge(hfls.all_september, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>%
  merge(hfss.all_september, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>%
  merge(hurs.all_september, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>% 
  merge(elev.df, by = c("lon", "lat"))

write.table(cmip6_september, "cmip6_september.txt")
```

```{r cmip6-october, eval = F}
tas.all_october <- read.table("CMIP6/tas/tas.all_october.txt")
pr.all_october <- read.table("CMIP6/pr/pr.all_october.txt")
sfcWind.all_october <- read.table("CMIP6/sfcWind/sfcWind.all_october.txt")
hfls.all_october <- read.table("CMIP6/hfls/hfls.all_october.txt")
hfss.all_october <- read.table("CMIP6/hfss/hfss.all_october.txt")
hurs.all_october <- read.table("CMIP6/hurs/hurs.all_october.txt")

cmip6_october <- tas.all_october %>% merge(pr.all_october, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>%
  merge(sfcWind.all_october, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>%
  merge(hfls.all_october, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>%
  merge(hfss.all_october, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>%
  merge(hurs.all_october, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>% 
  merge(elev.df, by = c("lon", "lat"))

write.table(cmip6_october, "cmip6_october.txt")
```

```{r cmip6-november, eval = F}
tas.all_november <- read.table("CMIP6/tas/tas.all_november.txt")
pr.all_november <- read.table("CMIP6/pr/pr.all_november.txt")
sfcWind.all_november <- read.table("CMIP6/sfcWind/sfcWind.all_november.txt")
hfls.all_november <- read.table("CMIP6/hfls/hfls.all_november.txt")
hfss.all_november <- read.table("CMIP6/hfss/hfss.all_november.txt")
hurs.all_november <- read.table("CMIP6/hurs/hurs.all_november.txt")

cmip6_november <- tas.all_november %>% merge(pr.all_november, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>%
  merge(sfcWind.all_november, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>%
  merge(hfls.all_november, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>%
  merge(hfss.all_november, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>%
  merge(hurs.all_november, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>% 
  merge(elev.df, by = c("lon", "lat"))

write.table(cmip6_november, "cmip6_november.txt")
```

```{r cmip6-december, eval = F}
tas.all_december <- read.table("CMIP6/tas/tas.all_december.txt")
pr.all_december <- read.table("CMIP6/pr/pr.all_december.txt")
sfcWind.all_december <- read.table("CMIP6/sfcWind/sfcWind.all_december.txt")
hfls.all_december <- read.table("CMIP6/hfls/hfls.all_december.txt")
hfss.all_december <- read.table("CMIP6/hfss/hfss.all_december.txt")
hurs.all_december <- read.table("CMIP6/hurs/hurs.all_december.txt")

cmip6_december <- tas.all_december %>% merge(pr.all_december, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>%
  merge(sfcWind.all_december, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>%
  merge(hfls.all_december, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>%
  merge(hfss.all_december, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>%
  merge(hurs.all_december, by = c("lon","lat","model","period","month","lm", "Continent","Type","Name","Acronym","source")) %>% 
  merge(elev.df, by = c("lon", "lat"))

write.table(cmip6_december, "cmip6_december.txt")
```


# Aridity Index

## De Martonne

```{r aridity-index-martonne, fig.dim = c(15, 20)}
cmip6 <- read.table("cmip6_annual.txt")

breaks.martonne <- c(0,5,10,20,"30","40",Inf)
cat.martonne <- c("[0,5]" = "Desert", "(5,10]"= "Arid", "(10,20]" = "Semi-arid","(20,30]" = "Temperate", "(30,40]" = "Humid", "(40,Inf]" = "Forest")
col.martonne <- c("Desert" = colorvec[2], "Arid"= colorvec[4],"Semi-arid"= colorvec[5], "Temperate" = colorvec[8], "Humid"= colorvec[10],"Forest"= colorvec[11])

cmip6$AIm <- with(cmip6,pr*60*60*24*365/(tas-273.15+10)) # pr in mm/y, tas in Ceslsius


cmip6$cat.AIm <- cmip6$AIm %>% cut(breaks.martonne, include.lowest = T) %>% revalue(cat.martonne)


map_list <- list()

for(i in unique(cmip6$source)){
  g <- ggplot(data = subset(cmip6, period == "1970_2000" & source == i)) + geom_raster(aes(x=lon, y = lat,  fill = cat.AIm))+
    scale_fill_manual(values = col.martonne, na.translate = F)+
    labs(title = paste(i, "1970-2000", sep = ", "), fill = "")+
    theme_void()
  map_list[[i]] <- g
}

ggpubr::ggarrange(plotlist = map_list, ncol = 3, nrow = 5, common.legend = T, legend = "bottom")
```

## UNEP

ea from Rh:

ea = 6.108*(RH/100)*exp(17.27*T/(T+237.3))
Refs
https://www.nature.com/articles/s41598-023-40499-6 : Allen and Tetens

```{r aridity-index-unesco, eval = F}
id_vars <- c("lon","lat","model","period","lm", "Continent","Type","Name","Acronym","source")

cmip6_annual <- mutate(read.table("cmip6_annual.txt"),
                       t = tas - 273.15,
                       Rn = (hfls + hfss) * 86400 * 1e-6, # conversion from W/m2 to MJ/m2/day.
                       P = 101.3 * ((293-0.0065*z)/293)^(5.26),
                       es = 0.61078*exp(17.27*t/(t+237.3)),
                       ea = (hurs/100)*es,
                       delta = 4098*es/(t+237.3)^2, #slope vapor pressure
                       gamma = 0.00163*P/(2.501-52.6361e-3*t), #psychrometric constant
                       ET0_annual = ((0.408*delta*Rn+gamma*900/((t)+273)*0.748*sfcWind*(es-ea))/(delta+gamma*(1+0.34*0.748*sfcWind)))) 

cmip6_01 <- mutate(read.table("cmip6_january.txt"),
                   t = tas - 273.15,
                   pr_01 = pr*60*60*24, # mm/day
                   Rn = (hfls + hfss) * 86400 * 1e-6, # conversion from W/m2 to MJ/m2/day.
                   P = 101.3 * ((293-0.0065*z)/293)^(5.26),
                   es = 0.61078*exp(17.27*t/(t+237.3)),
                   ea = (hurs/100)*es,
                   delta = 4098*es/(t+237.3)^2, #slope vapor pressure
                   gamma = 0.00163*P/(2.501-52.6361e-3*t), #psychrometric constant
                   ET0_01 = ((0.408*delta*Rn+gamma*900/(t+273)*0.748*sfcWind*(es-ea))/(delta+gamma*(1+0.34*0.748*sfcWind)))) # Evapotranspiration, G considered to be negligible. in mm/day. Wind speed at 10 m is converted to wind speed at 2 m by multiplying by 0.748

cmip6_02 <- mutate(read.table("cmip6_february.txt"),
                   t = tas - 273.15,
                   pr_02 = pr*60*60*24, # mm/day
                   Rn = (hfls + hfss) * 86400 * 1e-6, # conversion from W/m2 to MJ/m2/day.
                   P = 101.3 * ((293-0.0065*z)/293)^(5.26),
                   es = 0.61078*exp(17.27*t/(t+237.3)),
                   ea = (hurs/100)*es,
                   delta = 4098*es/(t+237.3)^2, #slope vapor pressure
                   gamma = 0.00163*P/(2.501-52.6361e-3*t), #psychrometric constant
                   ET0_02 = ((0.408*delta*Rn+gamma*900/(t+273)*0.748*sfcWind*(es-ea))/(delta+gamma*(1+0.34*0.748*sfcWind))))

cmip6_03 <- mutate(read.table("cmip6_march.txt"),
                   t = tas - 273.15,
                   pr_03 = pr*60*60*24, # mm/day
                   Rn = (hfls + hfss) * 86400 * 1e-6, # conversion from W/m2 to MJ/m2/day.
                   P = 101.3 * ((293-0.0065*z)/293)^(5.26),
                   es = 0.61078*exp(17.27*t/(t+237.3)),
                   ea = (hurs/100)*es,
                   delta = 4098*es/(t+237.3)^2, #slope vapor pressure
                   gamma = 0.00163*P/(2.501-52.6361e-3*t), #psychrometric constant
                   ET0_03 = ((0.408*delta*Rn+gamma*900/(t+273)*0.748*sfcWind*(es-ea))/(delta+gamma*(1+0.34*0.748*sfcWind))))

cmip6_04 <- mutate(read.table("cmip6_april.txt"),
                   t = tas - 273.15,
                   pr_04 = pr*60*60*24, # mm/day
                   Rn = (hfls + hfss) * 86400 * 1e-6, # conversion from W/m2 to MJ/m2/day.
                   P = 101.3 * ((293-0.0065*z)/293)^(5.26),
                   es = 0.61078*exp(17.27*t/(t+237.3)),
                   ea = (hurs/100)*es,
                   delta = 4098*es/(t+237.3)^2, #slope vapor pressure
                   gamma = 0.00163*P/(2.501-52.6361e-3*t), #psychrometric constant
                   ET0_04 = ((0.408*delta*Rn+gamma*900/(t+273)*0.748*sfcWind*(es-ea))/(delta+gamma*(1+0.34*0.748*sfcWind))))

cmip6_05 <- mutate(read.table("cmip6_may.txt"),
                   t = tas - 273.15,
                   pr_05 = pr*60*60*24, # mm/day
                   Rn = (hfls + hfss) * 86400 * 1e-6, # conversion from W/m2 to MJ/m2/day.
                   P = 101.3 * ((293-0.0065*z)/293)^(5.26),
                   es = 0.61078*exp(17.27*t/(t+237.3)),
                   ea = (hurs/100)*es,
                   delta = 4098*es/(t+237.3)^2, #slope vapor pressure
                   gamma = 0.00163*P/(2.501-52.6361e-3*t), #psychrometric constant
                   ET0_05 = ((0.408*delta*Rn+gamma*900/(t+273)*0.748*sfcWind*(es-ea))/(delta+gamma*(1+0.34*0.748*sfcWind))))

cmip6_06 <- mutate(read.table("cmip6_june.txt"),
                   t = tas - 273.15,
                   pr_06 = pr*60*60*24, # mm/day
                   Rn = (hfls + hfss) * 86400 * 1e-6, # conversion from W/m2 to MJ/m2/day.
                   P = 101.3 * ((293-0.0065*z)/293)^(5.26),
                   es = 0.61078*exp(17.27*t/(t+237.3)),
                   ea = (hurs/100)*es,
                   delta = 4098*es/(t+237.3)^2, #slope vapor pressure
                   gamma = 0.00163*P/(2.501-52.6361e-3*t), #psychrometric constant
                   ET0_06 = ((0.408*delta*Rn+gamma*900/(t+273)*0.748*sfcWind*(es-ea))/(delta+gamma*(1+0.34*0.748*sfcWind))))

cmip6_07 <- mutate(read.table("cmip6_july.txt"),
                   t = tas - 273.15,
                   pr_07 = pr*60*60*24, # mm/day
                   Rn = (hfls + hfss) * 86400 * 1e-6, # conversion from W/m2 to MJ/m2/day.
                   P = 101.3 * ((293-0.0065*z)/293)^(5.26),
                   es = 0.61078*exp(17.27*t/(t+237.3)),
                   ea = (hurs/100)*es,
                   delta = 4098*es/(t+237.3)^2, #slope vapor pressure
                   gamma = 0.00163*P/(2.501-52.6361e-3*t), #psychrometric constant
                   ET0_07 = ((0.408*delta*Rn+gamma*900/(t+273)*0.748*sfcWind*(es-ea))/(delta+gamma*(1+0.34*0.748*sfcWind))))

cmip6_08 <- mutate(read.table("cmip6_august.txt"),
                   t = tas - 273.15,
                   pr_08 = pr*60*60*24, # mm/day
                   Rn = (hfls + hfss) * 86400 * 1e-6, # conversion from W/m2 to MJ/m2/day.
                   P = 101.3 * ((293-0.0065*z)/293)^(5.26),
                   es = 0.61078*exp(17.27*t/(t+237.3)),
                   ea = (hurs/100)*es,
                   delta = 4098*es/(t+237.3)^2, #slope vapor pressure
                   gamma = 0.00163*P/(2.501-52.6361e-3*t), #psychrometric constant
                   ET0_08 = ((0.408*delta*Rn+gamma*900/(t+273)*0.748*sfcWind*(es-ea))/(delta+gamma*(1+0.34*0.748*sfcWind))))

cmip6_09 <- mutate(read.table("cmip6_september.txt"),
                   t = tas - 273.15,
                   pr_09 = pr*60*60*24, # mm/day
                   Rn = (hfls + hfss) * 86400 * 1e-6, # conversion from W/m2 to MJ/m2/day.
                   P = 101.3 * ((293-0.0065*z)/293)^(5.26),
                   es = 0.61078*exp(17.27*t/(t+237.3)),
                   ea = (hurs/100)*es,
                   delta = 4098*es/(t+237.3)^2, #slope vapor pressure
                   gamma = 0.00163*P/(2.501-52.6361e-3*t), #psychrometric constant
                   ET0_09 = ((0.408*delta*Rn+gamma*900/(t+273)*0.748*sfcWind*(es-ea))/(delta+gamma*(1+0.34*0.748*sfcWind))))

cmip6_10 <- mutate(read.table("cmip6_october.txt"),
                   t = tas - 273.15,
                   pr_10 = pr*60*60*24, # mm/day
                   Rn = (hfls + hfss) * 86400 * 1e-6, # conversion from W/m2 to MJ/m2/day.
                   P = 101.3 * ((293-0.0065*z)/293)^(5.26),
                   es = 0.61078*exp(17.27*t/(t+237.3)),
                   ea = (hurs/100)*es,
                   delta = 4098*es/(t+237.3)^2, #slope vapor pressure
                   gamma = 0.00163*P/(2.501-52.6361e-3*t), #psychrometric constant
                   ET0_10 = ((0.408*delta*Rn+gamma*900/(t+273)*0.748*sfcWind*(es-ea))/(delta+gamma*(1+0.34*0.748*sfcWind))))

cmip6_11 <- mutate(read.table("cmip6_november.txt"),
                   t = tas - 273.15,
                   pr_11 = pr*60*60*24, # mm/day
                   Rn = (hfls + hfss) * 86400 * 1e-6, # conversion from W/m2 to MJ/m2/day.
                   P = 101.3 * ((293-0.0065*z)/293)^(5.26),
                   es = 0.61078*exp(17.27*t/(t+237.3)),
                   ea = (hurs/100)*es,
                   delta = 4098*es/(t+237.3)^2, #slope vapor pressure
                   gamma = 0.00163*P/(2.501-52.6361e-3*t), #psychrometric constant
                   ET0_11 = ((0.408*delta*Rn+gamma*900/(t+273)*0.748*sfcWind*(es-ea))/(delta+gamma*(1+0.34*0.748*sfcWind))))

cmip6_12 <- mutate(read.table("cmip6_december.txt"),
                   t = tas - 273.15,
                   pr_12 = pr*60*60*24, # mm/day
                   Rn = (hfls + hfss) * 86400 * 1e-6, # conversion from W/m2 to MJ/m2/day.
                   P = 101.3 * ((293-0.0065*z)/293)^(5.26),
                   es = 0.61078*exp(17.27*t/(t+237.3)),
                   ea = (hurs/100)*es,
                   delta = 4098*es/(t+237.3)^2, #slope vapor pressure
                   gamma = 0.00163*P/(2.501-52.6361e-3*t), #psychrometric constant
                   ET0_12 = ((0.408*delta*Rn+gamma*900/(t+273)*0.748*sfcWind*(es-ea))/(delta+gamma*(1+0.34*0.748*sfcWind))))

et0_monthly <- select(cmip6_annual, id_vars) %>%
  merge(select(cmip6_01, c(id_vars, "ET0_01", "pr_01")), by = id_vars) %>%
  merge(select(cmip6_02, c(id_vars, "ET0_02", "pr_02")), by = id_vars) %>%
  merge(select(cmip6_03, c(id_vars, "ET0_03", "pr_03")), by = id_vars) %>%
  merge(select(cmip6_04, c(id_vars, "ET0_04", "pr_04")), by = id_vars) %>%
  merge(select(cmip6_05, c(id_vars, "ET0_05", "pr_05")), by = id_vars) %>%
  merge(select(cmip6_06, c(id_vars, "ET0_06", "pr_06")), by = id_vars) %>%
  merge(select(cmip6_07, c(id_vars, "ET0_07", "pr_07")), by = id_vars) %>%
  merge(select(cmip6_08, c(id_vars, "ET0_08", "pr_08")), by = id_vars) %>%
  merge(select(cmip6_09, c(id_vars, "ET0_09", "pr_09")), by = id_vars) %>%
  merge(select(cmip6_10, c(id_vars, "ET0_10", "pr_10")), by = id_vars) %>%   
  merge(select(cmip6_11, c(id_vars, "ET0_11", "pr_11")), by = id_vars) %>%
  merge(select(cmip6_12, c(id_vars, "ET0_12", "pr_12")), by = id_vars)

et0_monthly$spr <- with(et0_monthly, pr_01*31 + pr_02 * 28 + pr_03 * 31 + pr_04 * 30 + pr_05 * 31 + pr_06 * 30 +
                          pr_07 * 31 + pr_08 * 31 + pr_09 * 30 + pr_10 * 31 + pr_11 * 30 + pr_12 * 31)
et0_monthly$sET0 <- with(et0_monthly, ET0_01*31 + ET0_02 * 28 + ET0_03 * 31 + ET0_04 * 30 + ET0_05 * 31 + ET0_06 * 30 +
                           ET0_07 * 31 + ET0_08 * 31 + ET0_09 * 30 + ET0_10 * 31 + ET0_11 * 30 + ET0_12 * 31)

cmip6 <- merge(cmip6_annual, et0_monthly, by = id_vars)
cmip6$AI <- with(cmip6, abs(spr/sET0))  
cmip6$AI_annual <- with(cmip6, abs((pr*60*60*24*365)/(ET0_annual*365)))

breaks.unesco <- c(-Inf,0,0.03,0.2,0.5,0.65,Inf)
cat.unesco <- c("(0,0.03]"= "Hyperarid", "(0.03,0.2]" = "Arid", "(0.2,0.5]" = "Semi-arid", "(0.5,0.65]" = "Dry subhumid","(0.65, Inf]" = "Humid", "(-Inf,0]" = "Cold")
col.cat <- c("Hyperarid" = colorvec[2], "Arid"= colorvec[4],"Semi-arid"= colorvec[5], "Dry subhumid" = colorvec[8], "Humid"= colorvec[10], "Cold" = "powderblue")

cmip6$cat.AI <- cmip6$AI %>% cut(breaks = breaks.unesco) %>% revalue(cat.unesco)
cmip6$cat.AI[which(cmip6$sET0 < 400)] <- "Cold"
cmip6$cat.AI %>% unique()

cmip6$cat.AI <- cmip6$AI %>% cut(breaks = breaks.unesco) %>% 
  revalue(cat.unesco)


cmip6$cat.AI_annual <- cmip6$AI_annual %>% cut(breaks = breaks.unesco) %>% 
  revalue(cat.unesco)

cmip6$cat.AI_annual[which(cmip6$ET0_annual*365 < 400)] <- "Cold"

write.table(cmip6, "cmip6_AI.txt")
```

```{r summarises-ref, eval = F}

cmip6 <- read.table("cmip6_AI.txt")

cmip6_AI <- cmip6 %>% subset(period == "1970_2000") %>%
  group_by(lon,lat,Continent, Type, Name, Acronym, lm, model, z) %>% 
  dplyr::summarise(tas = mean(tas, na.rm = T) - 273.15, pr = mean(pr, na.rm = T) * 60 * 60 * 24 * 365,
                   sfcWind = mean(sfcWind, na.rm = T), Rn = mean(Rn, na.rm = T),
                   ea = mean(ea, na.rm = T), ET0 = mean(sET0, na.rm = T), AI = mean(AI, na.rm = T)) %>%
  ungroup() %>%  
  mutate(source = "CMIP6")%>%
  select(c("lon","lat","Continent","Type","Name","Acronym","lm","model","source","pr","tas","sfcWind","Rn", "ea","z","ET0","AI"))


write.table(cmip6_AI, "cmip6ref.txt")
```

# Maps AI by model 

```{r read-data, cache.extra = tools::md5sum("cmip6_AI.txt")}
cmip6 <- read.table("cmip6_AI.txt")

breaks.unesco <- c(-Inf,0.03,0.2,0.5,0.65,Inf)
cat.unesco <- c("(-Inf,0.03]"= "Hyperarid", "(0.03,0.2]" = "Arid", "(0.2,0.5]" = "Semi-arid", "(0.5,0.65]" = "Dry subhumid","(0.65, Inf]" = "Humid")

col.cat <- c("Hyperarid" = colorvec[2], "Arid"= colorvec[4],"Semi-arid"= colorvec[5], "Dry subhumid" = colorvec[8], "Humid"= colorvec[10], "Cold" = "powderblue")
```

## 1850-1880

```{r maps-hist-1850-1880, fig.dim = c(15, 20)}
map_list <- list()

for(i in unique(cmip6$source)){
  g <- ggplot(data = subset(cmip6, period == "1850_1880" & source == i)) + geom_raster(aes(x=lon, y = lat,  fill = cat.AI))+
    scale_fill_manual(values = col.cat, na.translate = F)+
    labs(title = paste(i, "1850-1880", sep = ", "), fill = "")+
    theme_void()
  map_list[[i]] <- g
}

ggpubr::ggarrange(plotlist = map_list, ncol = 3, nrow = 5, common.legend = T, legend = "bottom")
```

```{r table-of-percent-1850}

sdmm <- cmip6 %>% subset(period == "1850_1880" & !Continent %in% c("SOUTHERN","PACIFIC","ATLANTIC","INDIAN","ARCTIC")) %>%
  group_by(source, cat.AI) %>%
  summarise(count = n()) %>%
  ungroup() %>% group_by(source) %>% mutate(percent = round(count/sum(count)*100, 1), count = NULL) %>%
  ungroup() %>% group_by(cat.AI) %>% summarise(mean = round(mean(percent, na.rm = T),1), sd = round(sd(percent, na.rm = T),1)) %>%
  t() %>% as.data.frame() %>% setNames(c(.[1,1:6], "NA")) %>% .[-1, ] %>% sapply(as.numeric) %>% as.data.frame() %>%
  mutate("Sum drylands" = rowSums(.[c("Hyperarid","Arid","Semi-arid","Dry subhumid")]), "source" = c("mean","sd")) %>%
  select(c("source","Hyperarid","Arid","Semi-arid","Dry subhumid","Sum drylands","Humid","Cold","NA"))


tab <- cmip6 %>% subset(period == "1850_1880" & !Continent %in% c("SOUTHERN","PACIFIC","ATLANTIC","INDIAN","ARCTIC")) %>%
  group_by(source, cat.AI) %>%
  summarise(count = n()) %>%
  ungroup() %>% group_by(source) %>% mutate(percent = round(count/sum(count)*100, 1), count = NULL) %>%
  reshape2::dcast(source~cat.AI) %>%
  mutate("Sum drylands" = rowSums(.[c("Hyperarid","Arid","Semi-arid","Dry subhumid")])) %>%
  select(c("source","Hyperarid","Arid","Semi-arid","Dry subhumid","Sum drylands","Humid","Cold","NA")) %>% 
  rbind(sdmm)


knitr::kable(tab, digits = 2, escape = F)  %>%
  kable_styling(bootstrap_options = "bordered") %>% 
  column_spec(c(6,9), italic = T, include_thead = T) %>% row_spec(c(14,15), bold = T) 
```

## 1970-2000

```{r maps-hist-1970-2000, fig.dim = c(15, 20)}
map_list <- list()

for(i in unique(cmip6$source)){
  g <- ggplot(data = subset(cmip6, period == "1970_2000" & source == i)) + geom_raster(aes(x=lon, y = lat,  fill = cat.AI))+
    scale_fill_manual(values = col.cat, na.translate = F)+
    labs(title = paste(i, "1850-1880", sep = ", "), fill = "")+
    theme_void()
  map_list[[i]] <- g
}

ggpubr::ggarrange(plotlist = map_list, ncol = 3, nrow = 5, common.legend = T, legend = "bottom")
```

```{r percent-of-each-class-by-by-source, fig.dim = c(16,16)}
library(ggrepel) 

pie <- cmip6 %>% subset(period == "1970_2000" & !Continent %in% c("SOUTHERN","PACIFIC","POLAR","ATLANTIC","INDIAN","ARCTIC")) %>%
  select(c("source", "cat.AI")) %>%
  reshape2::melt(id.vars = c("source")) %>%
  group_by(source, value) %>%
  summarise(count = n()) %>%
  mutate(perc = round(count/sum(count)*100, 1),
         lab.y = rev(cumsum(rev(count))) - count*0.5) %>% 
  ungroup() 

pie_list <- list()

for(i in unique(pie$source)){
  p <- ggplot(subset(pie, source == i))+
    geom_bar(aes(x = "", y = count, fill = value), width = 1, stat = "identity")+
    coord_polar("y", start = 0)+
    scale_fill_manual(values = col.cat)+
    geom_label_repel(aes(x = "", y = lab.y, label = perc), nudge_x = 0.5)+
    labs(fill = "", title = i)+
    theme_void()+theme(legend.position = "right")
  pie_list[[i]] <- p
}

ggpubr::ggarrange(plotlist = pie_list, nrow = 4, ncol = 4, common.legend = T, legend = "bottom")
```

```{r table-of-percent-ref}

sdmm <- cmip6 %>% subset(period == "1970_2000" & !Continent %in% c("SOUTHERN","PACIFIC","ATLANTIC","INDIAN","ARCTIC")) %>%
  group_by(source, cat.AI) %>%
  summarise(count = n()) %>%
  ungroup() %>% group_by(source) %>% mutate(percent = round(count/sum(count)*100, 1), count = NULL) %>%
  ungroup() %>% group_by(cat.AI) %>% summarise(mean = round(mean(percent, na.rm = T),1), sd = round(sd(percent, na.rm = T),1)) %>%
  t() %>% as.data.frame() %>% setNames(c(.[1,1:6], "NA")) %>% .[-1, ] %>% sapply(as.numeric) %>% as.data.frame() %>%
  mutate("Sum drylands" = rowSums(.[c("Hyperarid","Arid","Semi-arid","Dry subhumid")]), "source" = c("mean","sd")) %>%
  select(c("source","Hyperarid","Arid","Semi-arid","Dry subhumid","Sum drylands","Humid","Cold","NA"))


tab <- cmip6 %>% subset(period == "1970_2000" & !Continent %in% c("SOUTHERN","PACIFIC","ATLANTIC","INDIAN","ARCTIC")) %>%
  group_by(source, cat.AI) %>%
  summarise(count = n()) %>%
  ungroup() %>% group_by(source) %>% mutate(percent = round(count/sum(count)*100, 1), count = NULL) %>%
  reshape2::dcast(source~cat.AI) %>%
  mutate("Sum drylands" = rowSums(.[c("Hyperarid","Arid","Semi-arid","Dry subhumid")])) %>%
  select(c("source","Hyperarid","Arid","Semi-arid","Dry subhumid","Sum drylands","Humid","Cold","NA")) %>% 
  rbind(sdmm)


knitr::kable(tab, digits = 2, escape = F)  %>%
  kable_styling(bootstrap_options = "bordered") %>% 
  column_spec(c(6,9), italic = T, include_thead = T) %>% row_spec(c(14,15), bold = T) 
```

## 1985-2010

```{r maps-hist-1985-2010, fig.dim = c(15,20)}
map_list <- list()

for(i in unique(cmip6$source)){
  g <- ggplot(data = subset(cmip6, period == "1985_2015" & source == i)) + geom_raster(aes(x=lon, y = lat,  fill = cat.AI))+
    scale_fill_manual(values = col.cat, na.translate = F)+
    labs(title = paste(i, "1985-2010", sep = ", "), fill = "")+
    theme_void()
  map_list[[i]] <- g
}

ggpubr::ggarrange(plotlist = map_list, ncol = 3, nrow = 5, common.legend = T, legend = "bottom")
```

```{r table-of-percent-1985}

sdmm <- cmip6 %>% subset(period == "1985_2015" & !Continent %in% c("SOUTHERN","PACIFIC","ATLANTIC","INDIAN","ARCTIC")) %>%
  group_by(source, cat.AI) %>%
  summarise(count = n()) %>%
  ungroup() %>% group_by(source) %>% mutate(percent = round(count/sum(count)*100, 1), count = NULL) %>%
  ungroup() %>% group_by(cat.AI) %>% summarise(mean = round(mean(percent, na.rm = T),1), sd = round(sd(percent, na.rm = T),1)) %>%
  t() %>% as.data.frame() %>% setNames(c(.[1,1:6], "NA")) %>% .[-1, ] %>% sapply(as.numeric) %>% as.data.frame() %>%
  mutate("Sum drylands" = rowSums(.[c("Hyperarid","Arid","Semi-arid","Dry subhumid")]), "source" = c("mean","sd")) %>%
  select(c("source","Hyperarid","Arid","Semi-arid","Dry subhumid","Sum drylands","Humid","Cold","NA"))


tab <- cmip6 %>% subset(period == "1985_2015" & !Continent %in% c("SOUTHERN","PACIFIC","ATLANTIC","INDIAN","ARCTIC")) %>%
  group_by(source, cat.AI) %>%
  summarise(count = n()) %>%
  ungroup() %>% group_by(source) %>% mutate(percent = round(count/sum(count)*100, 1), count = NULL) %>%
  reshape2::dcast(source~cat.AI) %>%
  mutate("Sum drylands" = rowSums(.[c("Hyperarid","Arid","Semi-arid","Dry subhumid")])) %>%
  select(c("source","Hyperarid","Arid","Semi-arid","Dry subhumid","Sum drylands","Humid","Cold","NA")) %>% 
  rbind(sdmm)


knitr::kable(tab, digits = 2, escape = F)  %>%
  kable_styling(bootstrap_options = "bordered") %>% 
  column_spec(c(6,9), italic = T, include_thead = T) %>% row_spec(c(14,15), bold = T) 
```

## SSP 2-4.5 

###  2030-2060

```{r maps-ssp245-2030-2060, fig.dim = c(15,20)}
map_list <- list()

for(i in unique(cmip6$source)){
  g <- ggplot(data = subset(cmip6, period == "2030_2060" & model == "SSP245" & source == i)) + geom_raster(aes(x=lon, y = lat,  fill = cat.AI))+
    scale_fill_manual(values = col.cat, na.translate = F)+
    labs(title = paste(i, "2030-2060", sep = ", "), fill = "")+
    theme_void()
  map_list[[i]] <- g
}

ggpubr::ggarrange(plotlist = map_list, ncol = 3, nrow = 5, common.legend = T, legend = "bottom")
```

```{r table-of-percent-SSP245-2030}

sdmm <- cmip6 %>% subset(period == "2030_2060" & model == "SSP245" &!Continent %in% c("SOUTHERN","PACIFIC","ATLANTIC","INDIAN","ARCTIC")) %>%
  group_by(source, cat.AI) %>%
  summarise(count = n()) %>%
  ungroup() %>% group_by(source) %>% mutate(percent = round(count/sum(count)*100, 1), count = NULL) %>%
  ungroup() %>% group_by(cat.AI) %>% summarise(mean = round(mean(percent, na.rm = T),1), sd = round(sd(percent, na.rm = T),1)) %>%
  t() %>% as.data.frame() %>% setNames(c(.[1,1:6], "NA")) %>% .[-1, ] %>% sapply(as.numeric) %>% as.data.frame() %>%
  mutate("Sum drylands" = rowSums(.[c("Hyperarid","Arid","Semi-arid","Dry subhumid")]), "source" = c("mean","sd")) %>%
  select(c("source","Hyperarid","Arid","Semi-arid","Dry subhumid","Sum drylands","Humid","Cold","NA"))

tab <- cmip6 %>% subset(period == "2030_2060" & model == "SSP245" & !Continent %in% c("SOUTHERN","PACIFIC","ATLANTIC","INDIAN","ARCTIC")) %>%
  group_by(source, cat.AI) %>%
  summarise(count = n()) %>%
  ungroup() %>% group_by(source) %>% mutate(percent = round(count/sum(count)*100, 1), count = NULL) %>%
  reshape2::dcast(source~cat.AI) %>%
  mutate("Sum drylands" = rowSums(.[c("Hyperarid","Arid","Semi-arid","Dry subhumid")])) %>%
  select(c("source","Hyperarid","Arid","Semi-arid","Dry subhumid","Sum drylands","Humid","Cold","NA")) %>% 
  rbind(sdmm)

knitr::kable(tab, digits = 2, escape = F)  %>%
  kable_styling(bootstrap_options = "bordered") %>% 
  column_spec(c(6,9), italic = T, include_thead = T) %>% row_spec(c(14,15), bold = T) 
```

### 2070-2100

```{r maps-ssp245-2070-2100, fig.dim = c(15,20)}
map_list <- list()

for(i in unique(cmip6$source)){
  g <- ggplot(data = subset(cmip6, period == "2070_2100" & model == "SSP245" & source == i)) + geom_raster(aes(x=lon, y = lat,  fill = cat.AI))+
    scale_fill_manual(values = col.cat, na.translate = F)+
    labs(title = paste(i, "2070-2100", sep = ", "), fill = "")+
    theme_void()
  map_list[[i]] <- g
}

ggpubr::ggarrange(plotlist = map_list, ncol = 3, nrow = 5, common.legend = T, legend = "bottom")
```

```{r table-of-percent-SSP245-2070}
sdmm <- cmip6 %>% subset(period == "2070_2100" & model == "SSP245" &!Continent %in% c("SOUTHERN","PACIFIC","ATLANTIC","INDIAN","ARCTIC")) %>%
  group_by(source, cat.AI) %>%
  summarise(count = n()) %>%
  ungroup() %>% group_by(source) %>% mutate(percent = round(count/sum(count)*100, 1), count = NULL) %>%
  ungroup() %>% group_by(cat.AI) %>% summarise(mean = round(mean(percent, na.rm = T),1), sd = round(sd(percent, na.rm = T),1)) %>%
  t() %>% as.data.frame() %>% setNames(c(.[1,1:6], "NA")) %>% .[-1, ] %>% sapply(as.numeric) %>% as.data.frame() %>%
  mutate("Sum drylands" = rowSums(.[c("Hyperarid","Arid","Semi-arid","Dry subhumid")]), "source" = c("mean","sd")) %>%
  select(c("source","Hyperarid","Arid","Semi-arid","Dry subhumid","Sum drylands","Humid","Cold","NA"))

tab <- cmip6 %>% subset(period == "2070_2100" & model == "SSP245" & !Continent %in% c("SOUTHERN","PACIFIC","ATLANTIC","INDIAN","ARCTIC")) %>%
  group_by(source, cat.AI) %>%
  summarise(count = n()) %>%
  ungroup() %>% group_by(source) %>% mutate(percent = round(count/sum(count)*100, 1), count = NULL) %>%
  reshape2::dcast(source~cat.AI) %>%
  mutate("Sum drylands" = rowSums(.[c("Hyperarid","Arid","Semi-arid","Dry subhumid")])) %>%
  select(c("source","Hyperarid","Arid","Semi-arid","Dry subhumid","Sum drylands","Humid","Cold","NA")) %>% 
  rbind(sdmm)

knitr::kable(tab, digits = 2, escape = F)  %>%
  kable_styling(bootstrap_options = "bordered") %>% 
  column_spec(c(6,9), italic = T, include_thead = T) %>% row_spec(c(14,15), bold = T) 
```

## SSP 3-7.0

### 2030-2060

```{r maps-ssp370-2030-2060, fig.dim = c(15,20)}
map_list <- list()

for(i in unique(cmip6$source)){
  g <- ggplot(data = subset(cmip6, period == "2030_2060" & model == "SSP370" & source == i)) + geom_raster(aes(x=lon, y = lat,  fill = cat.AI))+
    scale_fill_manual(values = col.cat, na.translate = F)+
    labs(title = paste(i, "2030-2060", sep = ", "), fill = "")+
    theme_void()
  map_list[[i]] <- g
}

ggpubr::ggarrange(plotlist = map_list, ncol = 3, nrow = 5, common.legend = T, legend = "bottom")
```

```{r table-of-percent-SSP370-2030}
sdmm <- cmip6 %>% subset(period == "2030_2060" & model == "SSP370" &!Continent %in% c("SOUTHERN","PACIFIC","ATLANTIC","INDIAN","ARCTIC")) %>%
  group_by(source, cat.AI) %>%
  summarise(count = n()) %>%
  ungroup() %>% group_by(source) %>% mutate(percent = round(count/sum(count)*100, 1), count = NULL) %>%
  ungroup() %>% group_by(cat.AI) %>% summarise(mean = round(mean(percent, na.rm = T),1), sd = round(sd(percent, na.rm = T),1)) %>%
  t() %>% as.data.frame() %>% setNames(c(.[1,1:6], "NA")) %>% .[-1, ] %>% sapply(as.numeric) %>% as.data.frame() %>%
  mutate("Sum drylands" = rowSums(.[c("Hyperarid","Arid","Semi-arid","Dry subhumid")]), "source" = c("mean","sd")) %>%
  select(c("source","Hyperarid","Arid","Semi-arid","Dry subhumid","Sum drylands","Humid","Cold","NA"))

tab <- cmip6 %>% subset(period == "2030_2060" & model == "SSP370" & !Continent %in% c("SOUTHERN","PACIFIC","ATLANTIC","INDIAN","ARCTIC")) %>%
  group_by(source, cat.AI) %>%
  summarise(count = n()) %>%
  ungroup() %>% group_by(source) %>% mutate(percent = round(count/sum(count)*100, 1), count = NULL) %>%
  reshape2::dcast(source~cat.AI) %>%
  mutate("Sum drylands" = rowSums(.[c("Hyperarid","Arid","Semi-arid","Dry subhumid")])) %>%
  select(c("source","Hyperarid","Arid","Semi-arid","Dry subhumid","Sum drylands","Humid","Cold","NA")) %>% 
  rbind(sdmm)

knitr::kable(tab, digits = 2, escape = F)  %>%
  kable_styling(bootstrap_options = "bordered") %>% 
  column_spec(c(6,9), italic = T, include_thead = T) %>% row_spec(c(14,15), bold = T) 
```

### 2070-2100

```{r maps-ssp370-2070-2100, fig.dim = c(15,20)}
map_list <- list()

for(i in unique(cmip6$source)){
  g <- ggplot(data = subset(cmip6, period == "2070_2100" & model == "SSP370" & source == i)) + geom_raster(aes(x=lon, y = lat,  fill = cat.AI))+
    scale_fill_manual(values = col.cat, na.translate = F)+
    labs(title = paste(i, "2070-2100", sep = ", "), fill = "")+
    theme_void()
  map_list[[i]] <- g
}

ggpubr::ggarrange(plotlist = map_list, ncol = 3, nrow = 5, common.legend = T, legend = "bottom")
```

```{r table-of-percent-SSP370-2070}
sdmm <- cmip6 %>% subset(period == "2070_2100" & model == "SSP370" &!Continent %in% c("SOUTHERN","PACIFIC","ATLANTIC","INDIAN","ARCTIC")) %>%
  group_by(source, cat.AI) %>%
  summarise(count = n()) %>%
  ungroup() %>% group_by(source) %>% mutate(percent = round(count/sum(count)*100, 1), count = NULL) %>%
  ungroup() %>% group_by(cat.AI) %>% summarise(mean = round(mean(percent, na.rm = T),1), sd = round(sd(percent, na.rm = T),1)) %>%
  t() %>% as.data.frame() %>% setNames(c(.[1,1:6], "NA")) %>% .[-1, ] %>% sapply(as.numeric) %>% as.data.frame() %>%
  mutate("Sum drylands" = rowSums(.[c("Hyperarid","Arid","Semi-arid","Dry subhumid")]), "source" = c("mean","sd")) %>%
  select(c("source","Hyperarid","Arid","Semi-arid","Dry subhumid","Sum drylands","Humid","Cold","NA"))

tab <- cmip6 %>% subset(period == "2070_2100" & model == "SSP370" & !Continent %in% c("SOUTHERN","PACIFIC","ATLANTIC","INDIAN","ARCTIC")) %>%
  group_by(source, cat.AI) %>%
  summarise(count = n()) %>%
  ungroup() %>% group_by(source) %>% mutate(percent = round(count/sum(count)*100, 1), count = NULL) %>%
  reshape2::dcast(source~cat.AI) %>%
  mutate("Sum drylands" = rowSums(.[c("Hyperarid","Arid","Semi-arid","Dry subhumid")])) %>%
  select(c("source","Hyperarid","Arid","Semi-arid","Dry subhumid","Sum drylands","Humid","Cold","NA")) %>% 
  rbind(sdmm)

knitr::kable(tab, digits = 2, escape = F)  %>%
  kable_styling(bootstrap_options = "bordered") %>% 
  column_spec(c(6,9), italic = T, include_thead = T) %>% row_spec(c(14,15), bold = T) 
```

## SSP 5-8.5

### 2030-2060

```{r maps-ssp585-2030-2060, fig.dim = c(15,20)}
map_list <- list()

for(i in unique(cmip6$source)){
  g <- ggplot(data = subset(cmip6, period == "2030_2060" & model == "SSP585" & source == i)) + geom_raster(aes(x=lon, y = lat,  fill = cat.AI))+
    scale_fill_manual(values = col.cat, na.translate = F)+
    labs(title = paste(i, "2030-2060", sep = ", "), fill = "")+
    theme_void()
  map_list[[i]] <- g
}

ggpubr::ggarrange(plotlist = map_list, ncol = 3, nrow = 5, common.legend = T, legend = "bottom")
```

```{r table-of-percent-SSP585-2030}
sdmm <- cmip6 %>% subset(period == "2030_2060" & model == "SSP585" &!Continent %in% c("SOUTHERN","PACIFIC","ATLANTIC","INDIAN","ARCTIC")) %>%
  group_by(source, cat.AI) %>%
  summarise(count = n()) %>%
  ungroup() %>% group_by(source) %>% mutate(percent = round(count/sum(count)*100, 1), count = NULL) %>%
  ungroup() %>% group_by(cat.AI) %>% summarise(mean = round(mean(percent, na.rm = T),1), sd = round(sd(percent, na.rm = T),1)) %>%
  t() %>% as.data.frame() %>% setNames(c(.[1,1:6], "NA")) %>% .[-1, ] %>% sapply(as.numeric) %>% as.data.frame() %>%
  mutate("Sum drylands" = rowSums(.[c("Hyperarid","Arid","Semi-arid","Dry subhumid")]), "source" = c("mean","sd")) %>%
  select(c("source","Hyperarid","Arid","Semi-arid","Dry subhumid","Sum drylands","Humid","Cold","NA"))

tab <- cmip6 %>% subset(period == "2030_2060" & model == "SSP585" & !Continent %in% c("SOUTHERN","PACIFIC","ATLANTIC","INDIAN","ARCTIC")) %>%
  group_by(source, cat.AI) %>%
  summarise(count = n()) %>%
  ungroup() %>% group_by(source) %>% mutate(percent = round(count/sum(count)*100, 1), count = NULL) %>%
  reshape2::dcast(source~cat.AI) %>%
  mutate("Sum drylands" = rowSums(.[c("Hyperarid","Arid","Semi-arid","Dry subhumid")])) %>%
  select(c("source","Hyperarid","Arid","Semi-arid","Dry subhumid","Sum drylands","Humid","Cold","NA")) %>% 
  rbind(sdmm)

knitr::kable(tab, digits = 2, escape = F)  %>%
  kable_styling(bootstrap_options = "bordered") %>% 
  column_spec(c(6,9), italic = T, include_thead = T) %>% row_spec(c(14,15), bold = T) 
```

### 2070-2100

```{r maps-ssp585-2070-2100, fig.dim = c(15,20)}
map_list <- list()

for(i in unique(cmip6$source)){
  g <- ggplot(data = subset(cmip6, period == "2070_2100" & model == "SSP585" & source == i)) + geom_raster(aes(x=lon, y = lat,  fill = cat.AI))+
    scale_fill_manual(values = col.cat, na.translate = F)+
    labs(title = paste(i, "2070-2100", sep = ", "), fill = "")+
    theme_void()
  map_list[[i]] <- g
}

ggpubr::ggarrange(plotlist = map_list, ncol = 3, nrow = 5, common.legend = T, legend = "bottom")
```

```{r table-of-percent-SSP585-2070}
sdmm <- cmip6 %>% subset(period == "2070_2100" & model == "SSP585" &!Continent %in% c("SOUTHERN","PACIFIC","ATLANTIC","INDIAN","ARCTIC")) %>%
  group_by(source, cat.AI) %>%
  summarise(count = n()) %>%
  ungroup() %>% group_by(source) %>% mutate(percent = round(count/sum(count)*100, 1), count = NULL) %>%
  ungroup() %>% group_by(cat.AI) %>% summarise(mean = round(mean(percent, na.rm = T),1), sd = round(sd(percent, na.rm = T),1)) %>%
  t() %>% as.data.frame() %>% setNames(c(.[1,1:6], "NA")) %>% .[-1, ] %>% sapply(as.numeric) %>% as.data.frame() %>%
  mutate("Sum drylands" = rowSums(.[c("Hyperarid","Arid","Semi-arid","Dry subhumid")]), "source" = c("mean","sd")) %>%
  select(c("source","Hyperarid","Arid","Semi-arid","Dry subhumid","Sum drylands","Humid","Cold","NA"))

tab <- cmip6 %>% subset(period == "2070_2100" & model == "SSP585" & !Continent %in% c("SOUTHERN","PACIFIC","ATLANTIC","INDIAN","ARCTIC")) %>%
  group_by(source, cat.AI) %>%
  summarise(count = n()) %>%
  ungroup() %>% group_by(source) %>% mutate(percent = round(count/sum(count)*100, 1), count = NULL) %>%
  reshape2::dcast(source~cat.AI) %>%
  mutate("Sum drylands" = rowSums(.[c("Hyperarid","Arid","Semi-arid","Dry subhumid")])) %>%
  select(c("source","Hyperarid","Arid","Semi-arid","Dry subhumid","Sum drylands","Humid","Cold","NA")) %>% 
  rbind(sdmm)

knitr::kable(tab, digits = 2, escape = F)  %>%
  kable_styling(bootstrap_options = "bordered") %>% 
  column_spec(c(6,9), italic = T, include_thead = T) %>% row_spec(c(13,14), bold = T) 
```

# Agreement between models

## Summary dataframe

### Difference with AI ref for all SSP 

```{r diff-AI-cmip6, eval = F}
cmip6 <- read.table("cmip6_AI.txt")

tab <- cmip6

list_ref <- list()

for(i in unique(tab$source)){
  tab.i <- filter(tab, source == i) 
  
  if(i == "CESM"){
    tab.Ai.ref <- filter(tab.i, period == "1970_2000") %>% pull(AI) %>% rep(8)
    tab.i$AI.ref <- tab.Ai.ref
    
    tab.pr.ref <- filter(tab.i, period == "1970_2000") %>% pull(spr) %>% rep(8)
    tab.i$pr.ref <- tab.pr.ref
    
    tab.tmean.ref <- filter(tab.i, period == "1970_2000") %>% pull(t) %>% rep(8)
    tab.i$t.ref <- tab.tmean.ref
    
  }else{
    tab.AI.ref <- filter(tab.i, period == "1970_2000") %>% pull(AI) %>% rep(9)
    tab.i$AI.ref <- tab.AI.ref
    
    tab.pr.ref <- filter(tab.i, period == "1970_2000") %>% pull(spr) %>% rep(9)
    tab.i$pr.ref <- tab.pr.ref
    
    tab.tmean.ref <- filter(tab.i, period == "1970_2000") %>% pull(t) %>% rep(9)
    tab.i$t.ref <- tab.tmean.ref
  }
  list_ref[[i]] <- tab.i  
}


cmip6ext <- bind_rows(list_ref, .id = "column_label") %>%
  mutate(diff.AI = AI - AI.ref, diff.pr = pr - pr.ref, diff.t = t - t.ref)
write.table(cmip6ext, "cmip6ext.txt")


```


```{r agreement-between-models, eval = F}
cmip6 <- read.table("cmip6ext.txt")
cmip6s <- cmip6 %>% 
  group_by(lon,lat,Continent, Type, Name, Acronym, lm, model, period, z) %>% 
  dplyr::summarise(t.mean = mean(t, na.rm = T), t.sd = sd(t, na.rm = T), # t in C
                   pr.mean = mean(spr, na.rm = T) , pr.sd = sd(spr, na.rm = T),  # pr in mm/y
                   sfcWind.mean = mean(sfcWind, na.rm = T), sfcWind.sd = sd(sfcWind, na.rm = T),
                   Rn.mean = mean(Rn, na.rm = T), Rn.sd = sd(Rn, na.rm = T), # in MJ/m2/day
                   ea.mean = mean(ea, na.rm = T), ea.sd = sd(ea, na.rm = T), # in kPa
                   ET0 = mean(sET0, na.rm = T), ET0.sd = sd(sET0, na.rm = T), # in mm/y
                   AI.mean = mean(AI, na.rm = T), AI.sd = sd(AI, na.rm = T),
                   diff.AI.mean = mean(diff.AI, na.rm = T), diff.AI.sd = sd(diff.AI, na.rm = T),
                   hyperarid = sum(cat.AI == "Hyperarid"),
                   arid = sum(cat.AI == "Arid"),
                   semiarid = sum(cat.AI == "Semi-arid"),
                   drysubhumid = sum(cat.AI == "Dry subhumid") ,
                   humid = sum(cat.AI == "Humid"),
                   cold = sum(cat.AI == "Cold"),
                   na = sum(is.na(cat.AI)),
                   nb.models = n(),
                   nb.maj = max(hyperarid,arid,semiarid,drysubhumid,humid,cold,na),
                   prop.maj = nb.maj/nb.models,
                   plus.AI = table(diff.AI > 0)["TRUE"],
                   minus.AI = table(diff.AI < 0)["TRUE"]) %>% 
  ungroup() %>% as.data.frame()

cmip6s$cat.maj <- colnames(select(cmip6s, c("hyperarid","arid",'semiarid',"drysubhumid",'humid',"cold","na")))[max.col(select(cmip6s, c("hyperarid","arid",'semiarid',"drysubhumid",'humid',"cold","na")),ties.method = "first")]

write.table(cmip6s, "cmip6s.txt")

#test <- cmip6 %>% subset(lon == -10.5 & lat == 6.5 & Continent == "AFRICA" & model == "SSP245" & period == "2030_2060")
```

```{r summary-periods, eval = F}
cmip6s <- read.table("cmip6s.txt")

breaks.unesco <- c(-Inf,0,0.03,0.2,0.5,0.65,Inf)
cat.unesco <- c("(0,0.03]"= "Hyperarid", "(0.03,0.2]" = "Arid", "(0.2,0.5]" = "Semi-arid", "(0.5,0.65]" = "Dry subhumid","(0.65, Inf]" = "Humid", "(-Inf,0]" = "Cold")

cmip6s$cat.AI <- cmip6s$AI.mean %>% cut(breaks = breaks.unesco) %>% revalue(cat.unesco)
cmip6s$cat.AI[which(cmip6s$ET0 < 400)] <- "Cold"

tab <- cmip6s
list_ref <- list()

for(i in unique(tab$lon)){
  tab.i <- filter(tab, lon == i)
  for(j in unique(tab$lat)){
    index <- paste(i,j,sep = ",")
    tab.ij <- filter(tab.i, lat == j)
    tab.ij$AIref <- filter(tab.ij, period == "1970_2000") %>% pull(AI.mean) %>% rep(9) #3 historical scenario + 2 x 3 future scenario
    tab.ij$catAIref <- filter(tab.ij, period == "1970_2000") %>% pull(cat.AI) %>% rep(9)
    tab.ij$catmajref <- filter(tab.ij, period == "1970_2000") %>% pull(cat.maj) %>% rep(9)
    list_ref[[index]] <- tab.ij 
  }
}

cmip6ss <- bind_rows(list_ref)

cmip6ss$anomalies <- with(cmip6ss, AI.mean - AIref)

cmip6ss$same.catAI <- with(cmip6ss, ifelse(cat.maj == cat.AI, "same", "diff"))

write.table(cmip6ss, "cmip6ss.txt")
```

```{r read-summary-cmip6, cache.extra = tools::md5sum("cmip6s2.txt")}
cmip6s <- read.table("cmip6ss.txt")

# attributes a change from the studied period compared to the reference period 1970-2000. Read as: grid cell x is humid in 2030-2060 and Arid in 1970-2000, so it is WETTER

#cmip6s$change <-  with(cmip6s, paste(catAIref, cat.AI, collapse = " ", sep = " to "))

cmip6s$dryer <- with(cmip6s, 
          ifelse(catAIref == "Arid" & cat.AI %in% c("Cold","Humid","Dry subhumid","Semi-arid"), "wetter", 
          ifelse(catAIref == "Arid" & cat.AI == "Hyperarid", "dryer",
          ifelse(catAIref == "Semi-arid" & cat.AI %in% c("Cold","Humid","Dry subhumid"), "wetter",
          ifelse(catAIref == "Semi-arid" & cat.AI %in% c("Hyperarid","Arid"), "dryer",
          ifelse(catAIref == "Dry subhumid" & cat.AI %in% c("Cold","Humid"), "wetter",
          ifelse(catAIref == "Dry subhumid" & cat.AI %in% c("Semi-arid","Arid","Hyperarid"), "dryer",
          ifelse(catAIref == "Humid" & cat.AI == "Cold","wetter",
          ifelse(catAIref == "Humid" & cat.AI %in% c("Dry subhumid","Semi-arid","Arid","Hyperarid"), "dryer",
          ifelse(catAIref == "Cold" & cat.AI %in% c("Humid","Dry subhumid","Semi-arid","Arid","Hyperarid"), "dryer",
          ifelse(catAIref == "Hyperarid" & cat.AI %in% c("Cold","Humid","Dry subhumid","Semi-arid","Arid"), "wetter",
          ifelse(catAIref == cat.AI, "same",
          NA))))))))))))

cmip6s$catmajref <- revalue(cmip6s$catmajref, c("hyperarid" = "Hyperarid", "arid" = "Arid", "semiarid" = "Semi-arid", "drysubhumid" = "Dry subhumid", "humid" = "Humid", "cold" = "Cold"))

cmip6s$cat.maj <- revalue(cmip6s$cat.maj, c("hyperarid" = "Hyperarid", "arid" = "Arid", "semiarid" = "Semi-arid", "drysubhumid" = "Dry subhumid", "humid" = "Humid", "cold" = "Cold"))



cmip6s$dryer.maj <- with(cmip6s, 
        ifelse(catmajref == "Arid" & cat.maj %in% c("Cold","Humid","Dry subhumid","Semi-arid"), "wetter", 
        ifelse(catmajref == "Arid" & cat.maj == "Hyperarid", "dryer",
        ifelse(catmajref == "Semi-arid" & cat.maj %in% c("Cold","Humid","Dry subhumid"), "wetter",
        ifelse(catmajref == "Semi-arid" & cat.maj %in% c("Hyperarid","Arid"), "dryer",
        ifelse(catmajref == "Dry subhumid" & cat.maj %in% c("Cold","Humid"), "wetter",
        ifelse(catmajref == "Dry subhumid" & cat.maj %in% c("Semi-arid","Arid","Hyperarid"), "dryer",
        ifelse(catmajref == "Humid" & cat.maj == "Cold","wetter",
        ifelse(catmajref == "Humid" & cat.maj %in% c("Dry subhumid","Semi-arid","Arid","Hyperarid"), "dryer",
        ifelse(catmajref == "Cold" & cat.maj %in% c("Humid","Dry subhumid","Semi-arid","Arid","Hyperarid"), "dryer",
        ifelse(catmajref == "Hyperarid" & cat.maj %in% c("Cold","Humid","Dry subhumid","Semi-arid","Arid"), "wetter",
        ifelse(catmajref == cat.maj, "same",
        NA))))))))))))


```


## Compare cat mean and cat maj

cat mean = category of aridity obtained by using the multimodel mean AI
cat maj = category of aridity obtained by using the category majoritarily obtained in the 13 models

### Where are cat mean and cat maj different

```{r table-comparing-catmaj-catmean, fig.dim = c(8,6)}

cmip6s %>% group_by(model, period) %>% summarise(maj.equal.mean = table(same.catAI)[2], maj.diff.mean = table(same.catAI)[1], prop.same = round(maj.equal.mean/n()*100,1)) %>% knitr::kable(col.names = c("Model","Period","Number of gridcells in which cat.maj = cat.mean", "Number of gridcells in which cat.maj /= cat.mean", "Proportion of cells with the same category (%)")) %>% kableExtra::kable_styling(bootstrap_options = "bordered")

cmip6s$cat.AI <- factor(cmip6s$cat.AI, levels =  c("Hyperarid","Arid","Semi-arid","Dry subhumid", "Humid","Cold"))
cmip6s$cat.maj <- factor(cmip6s$cat.maj, levels =  c("Hyperarid","Arid","Semi-arid","Dry subhumid", "Humid","Cold"))

ggplot(subset(cmip6s, cat.AI != "Cold"))+
  geom_col(aes(x= period, y = cat.AI, fill = cat.AI), just = -0.2, width = 0.3)+
  geom_col(aes(x = period, y = cat.maj, fill = cat.maj), just = 1.2, width = 0.3)+
  scale_fill_manual(values = col.cat)+
  labs(fill = "", y ="")+
  facet_grid(rows = vars(model), switch = "y")+
  theme_minimal()+theme(axis.text.y = element_blank())



tab <- rbind(table(cmip6s$cat.AI)/length(cmip6s$cat.AI)*100, table(cmip6s$cat.maj)/length(cmip6s$cat.maj)*100) %>% t() %>% as.data.frame() %>% setNames(c("Cat mean", "Cat maj"))
knitr::kable(tab, digits =2, caption = "Proportion of gridcells in each aridity category") %>% kable_styling(bootstrap_options = "bordered")

```

```{r maps-of-cat-differences, fig.dim = c(10,27)}
map_list <- list()

for(i in unique(cmip6s$model)){
  df <- subset(cmip6s, model == i)
  for(j in unique(df$period)){
    index <- paste(i, j, sep = "")
    g <- ggplot()+geom_raster(data = subset(df, period == j & same.catAI == "diff"), aes(x = lon, y = lat, fill = cat.AI))+
      borders(colour = "grey60")+
      scale_fill_manual(values =  col.cat, na.translate = F)+
      labs(title = paste(i, j , sep = " "), fill = "")+
      theme_void()+ylim(-55,90)+
      theme(legend.position = "bottom")
    map_list[[index]] <- g
  }
}

col1 <- ggarrange(plotlist = map_list, nrow = 9, ncol = 1, common.legend = T, legend = "bottom") %>% annotate_figure(top = "cat.mean for gridcells in which\ncat.maj and cat.mean are different")

map_list2 <- list()

for(i in unique(cmip6s$model)){
  df <- subset(cmip6s, model == i)
  for(j in unique(df$period)){
    index <- paste(i, j, sep = "")
    g <- ggplot()+geom_raster(data = subset(df, period == j & same.catAI == "diff"), aes(x = lon, y = lat, fill = cat.maj))+
      borders(colour = "grey60")+
      scale_fill_manual(values =  col.cat, na.translate = F)+
      labs(title = paste(i, j , sep = " "), fill = "")+
      theme_void()+ylim(-55,90)+
      theme(legend.position = "bottom")
    map_list2[[index]] <- g
  }
}

col2 <- ggarrange(plotlist = map_list2, nrow = 9, ncol = 1, common.legend = T, legend = "bottom") %>% annotate_figure(top = "cat.maj for gridcells in which\ncat.maj and cat.mean are different")

ggarrange(col1,col2,ncol = 2, common.legend = T)

```

### Plot cat maj and cat mean for reference period

```{r ref-cat-maj, fig.dim = c(10,4), eval = F}
ggpubr::ggarrange(plotlist = list(
  
  ggplot() + 
    geom_raster(data = subset(cmip6s, period == "1970_2000"), aes(x=lon, y = lat,  fill = cat.maj))+
    geom_point(data = subset(cmip6s, period == "1970_2000" & nb.maj > 10), aes(x =lon, y = lat), shape = "'", size = 1, col = "grey20")+
    scale_fill_manual(values = col.cat)+
    labs(title = "Category chosen in majority by models", fill = "")+
    ylim(-55,90)+
    theme_void(base_size = 15)+
    theme(legend.position = "none"),
  
  ggplot() + 
    geom_raster(data = subset(cmip6s, period == "1970_2000"), aes(x=lon, y = lat,  fill = cat.AI))+
    geom_point(data = subset(cmip6s, period == "1970_2000" & nb.maj > 10), aes(x =lon, y = lat), shape = "'", size = 1, col = "grey20")+
    scale_fill_manual(values = col.cat)+
    labs(title = "Multimodel average categories of AI", fill = "")+
    ylim(-55,90)+
    theme_void(base_size = 15)+
    theme(legend.position = "none")),
  
  ncol = 2)

```


```{r is-cat-maj-same-as-recomputed-cat, fig.dim = c(10,10)}
map_list <- list()

for(i in c("1850_1880","1970_2000","1985_2015")){
  g <- ggplot() + geom_raster(data = subset(cmip6s, period == i & model == "historical"), aes(x=lon, y = lat,  fill = cat.maj))+
    geom_point(data = subset(cmip6s, period == i & model == "historical" & nb.maj > 10), aes(x= lon, y = lat), shape = "'", col = "black")+
    scale_fill_manual(values =  col.cat, na.translate = F)+
    labs(title = i, fill = "")+
    theme_void()+ylim(-55,90)+
    theme(legend.position = "bottom")
  map_list[[i]] <- g
}


map_list2 <- list()

for(i in c("1850_1880","1970_2000","1985_2015")){
  g <- ggplot() + geom_raster(data = subset(cmip6s, period == i & model == "historical"), aes(x=lon, y = lat,  fill = cat.AI))+
    geom_point(data = subset(cmip6s, period == i & model == "historical" & nb.maj > 10), aes(x= lon, y = lat), shape = "'", col = "black")+
    scale_fill_manual(values =  col.cat, na.translate = F)+
    labs(title = i, fill = "")+
    theme_void()+ylim(-55,90)+
    theme(legend.position = "bottom")
  map_list2[[i]] <- g
}


ggarrange(
  annotate_figure(ggarrange(plotlist = map_list, nrow = 3, common.legend = T, legend = "bottom"), top = "Majority category among models"),
  annotate_figure(ggarrange(plotlist = map_list2, nrow = 3, common.legend = T, legend = "bottom"), top = "Multimodel average category"),
  ncol = 2, legend = "bottom")
```

```{r is-cat-maj-same-as-recomputed-cat-future, fig.dim = c(10,10)}
map_list <- list()

for(i in c("SSP245","SSP370","SSP585")){
  for (j in c("2030_2060","2070_2100")){
    index <- paste(i, j, sep = " , ")
    g <- ggplot() + geom_raster(data = subset(cmip6s, period == j & model == i), aes(x=lon, y = lat,  fill = cat.maj))+
      #   geom_point(data = subset(cmip6s, period == i & model == "historical" & change.cat == 1 & nb.maj > 10), aes(x= lon, y = lat), shape = "'")+
      scale_fill_manual(values = col.cat, na.translate = F)+
      labs(fill = "Cat AI maj", title = paste(i, j, sep = ", "))+
      theme_void()+ylim(-55,90)+
      theme(legend.position = "bottom")
    map_list[[index]] <- g
  }}


map_list2 <- list()

for(i in c("SSP245","SSP370","SSP585")){
  for (j in c("2030_2060","2070_2100")){
    index <- paste(i, j, sep = " , ")
    g <- ggplot() + geom_raster(data = subset(cmip6s, period == j & model == i), aes(x=lon, y = lat,  fill = cat.AI))+
      #   geom_point(data = subset(cmip6s, period == i & model == "historical" & change.cat == 1 & nb.maj > 10), aes(x= lon, y = lat), shape = "'")+
      scale_fill_manual(values = col.cat, na.translate = F)+
      labs(fill = "Cat AI mean", title = paste(i, j, sep = ", "))+
      theme_void()+ylim(-55,90)+
      theme(legend.position = "bottom")
    map_list2[[index]] <- g
  }}

ggarrange(plotlist = map_list, nrow = 3, ncol = 2, common.legend = T, legend = "bottom")
ggarrange(plotlist = map_list2, nrow = 3, ncol = 2, common.legend = T, legend = "bottom")

```



## Multimodel maps of cat mean

```{r map-historical-changes-AI-catmean, fig.dim = c(4, 10)}

cmip6s$cat.AI <- factor(cmip6s$cat.AI, levels = c("Hyperarid","Arid","Semi-arid","Dry subhumid","Humid","Cold"))

map_list <- list()

for(i in c("1850_1880", "1970_2000","1985_2015")){
  g <- ggplot() + 
    geom_tile(data = subset(cmip6s, period == i & model == "historical"), aes(x=lon, y = lat,  fill = cat.AI))+
    geom_point(data = subset(cmip6s,nb.maj > 10), aes(x = lon, y = lat), shape = "'", col = "grey60")+
    borders(colour = "grey60")+
    scale_fill_manual(values =  col.cat, na.translate = F)+
    labs(title = paste("Category AI",i , sep = " "), fill = "")+
    theme_void()+ylim(-55,90)+
    theme(legend.position = "bottom")
  map_list[[i]] <- g
}

ggarrange(plotlist = map_list, ncol = 1, nrow = 3, common.legend = T, legend = "bottom")

```

```{r map-future-categories-AI, fig.dim = c(10,10)}

map_list <- list()

for(i in c("SSP245","SSP370","SSP585")){
  for (j in c("2030_2060","2070_2100")){
    index <- paste(i, j, sep = " , ")
    g <- ggplot() + 
      geom_tile(data = subset(cmip6s, period == j & model == i), aes(x=lon, y = lat,  fill = cat.AI))+
      geom_point(data = subset(cmip6s,nb.maj > 10), aes(x = lon, y = lat), shape = ".", col = "grey20", size = 0.1)+
      borders(colour = "grey60")+
      scale_fill_manual(values =  col.cat, na.translate = F)+
      labs(title = paste("Category AI",index, sep = " "), fill = "")+
      theme_void()+ylim(-55,90)+
      theme(legend.position = "bottom")
    map_list[[index]] <- g
  }}

ggpubr::ggarrange(plotlist = map_list,ncol = 2, nrow = 3, common.legend = T, legend = "bottom")
```

# Maps multimodel AI

```{r aridity-index}
quantile(cmip6s$AI.mean, probs = seq(0,1,0.1), na.rm = T) 

ai.breaks <- c(0,0.03,0.2,0.5,0.65,1,10,20,30)
colscale <- colorvec[-c(5,6)]
```

```{r map-historical-changes-quantiles-AI, fig.dim = c(5,10)}
map_list <- list()

for(i in c("1850_1880", "1970_2000","1985_2015")){
  g <- ggplot() + geom_raster(data = subset(cmip6s, period == i & model == "historical"), aes(x=lon, y = lat,  fill = AI.mean))+
    borders(colour = "grey60")+
    binned_scale(aesthetics = "fill", breaks = ai.breaks, palette = function(x) colscale,
                 guide = guide_legend(label.theme = element_text(angle = 0)))+
    labs(title = i, fill = "Aridity index")+
    theme_void()+ylim(-55,90)+
    theme(legend.position = "right")
  map_list[[i]] <- g
}

ggpubr::ggarrange(plotlist = map_list, ncol = 1, nrow = 3, common.legend = T, legend = "bottom")

```

```{r map-future-changes-quantiles-AI, fig.dim = c(10,10)}

map_list <- list()

for(i in c("SSP245","SSP370","SSP585")){
  for (j in c("2030_2060","2070_2100")){
    index <- paste(i, j, sep = " , ")
    g <- ggplot() + geom_raster(data = subset(cmip6s, period == j & model == i), aes(x=lon, y = lat,  fill = AI.mean))+
      borders(colour = "grey60")+
      binned_scale(aesthetics = "fill", breaks = ai.breaks, palette = function(x) colscale,
                   guide = guide_legend(label.theme = element_text(angle = 0)))+
      labs(fill = "Aridity index", title = index)+
      theme_void()+ylim(-55,90)+
      theme(legend.position = "bottom")
    map_list[[index]] <- g
  }}

ggpubr::ggarrange(plotlist = map_list,ncol = 2, nrow = 3, common.legend = T, legend = "bottom")
```

# AI changes over time and scenarios

```{r validation}
v<-NULL
v2<-NULL
for (N in 5:30) {
  v<-c(v,sum(sapply(floor(2*N/3):N,function(i) choose(N,i)*(1/2)^N*2)))# proba d'avoir 2/3 de mme signe  partir de N modles, lorsque N change
  v2<-c(v2,sum(sapply(floor(2*N/3):N,function(i) choose(N,i)*(1/2)^N)))# proba d'avoir 2/3 de signe >0  partir de N modles, lorsque N change
}


plot(5:30,v,type="l",ylab="Proba(>2/3 models...)",xlab="Nb of models")
lines(5:30,v2,col="red")

```

# Changes AI for summary CMIP6

## Boxplot of changes AI

```{r boxplot-diff, fig.dim = c(8,5)}
ggpubr::ggarrange(plotlist = list(
  ggplot(subset(cmip6s, anomalies < 1 & anomalies > -1 & model == "historical"))+
    geom_boxplot(aes(x=period, y = - anomalies, col = period), outlier.size = 1)+
    scale_color_manual(values = c("#1A9850","#66BD63","#A6D96A"))+
    labs(x = "", y = "difference of AI between each period and\nthe reference period 1970-2000")+
    facet_grid(cols = vars(model))+
    theme_minimal() + theme(legend.position = "none"),
  ggplot(subset(cmip6s, anomalies < 1 & anomalies > -1 & model != "historical"))+
    geom_boxplot(aes(x=period, y = anomalies, col = period), outlier.size = 1)+
    scale_color_manual(values = c("#FDAE61","#F46D43"))+
    scale_y_continuous(position = "right")+
    labs(y = "", x = "")+
    facet_grid(cols = vars(model))+
    theme_minimal() + theme(legend.position = "none")),
  widths = c(1.5,3),
  ncol = 2, nrow = 1
)

```

## Strictly wetter or dryer (reference period: 1970-2000)

```{r questcequinemarchepas}
test <- cmip6s %>% select(c("lon","lat","model","period","dryer","cat.AI","catAIref","AI.mean", "AIref", "anomalies","plus.AI", "minus.AI"))

View(subset(test, dryer == "dryer"))

ggplot(subset(cmip6s, lat > -55 & !is.na(dryer))) + 
  geom_boxplot(aes(y = cat.AI, x = diff.AI.mean, col = catAIref), outliers = F) +
  scale_color_manual(values = col.cat)+
  facet_grid(cols = vars(dryer))+
  labs(y = "Aridity category", x = "Difference between AI and AI ref", col = "Reference aridity category")+
  theme_bw()

ggplot(subset(cmip6s, lat > -55 & !is.na(dryer))) + 
  geom_boxplot(aes(y = cat.AI, x = anomalies, col = catAIref), outliers = F) +
  scale_color_manual(values = col.cat)+
  facet_grid(cols = vars(dryer))+
  labs(y = "Aridity category", x = "Difference between AI and AI ref", col = "Reference aridity category")+
  theme_bw()


```

```{r map-historical-changes-strict, fig.dim = c(8,3)}
map_list <- list()

for(i in c("1850_1880","1985_2015")){
  g <- ggplot() + geom_tile(data = subset(cmip6s, period == i & model == "historical" & anomalies > 0), aes(x=lon, y = lat,  fill = "wetter"))+
    geom_tile(data = subset(cmip6s, period == i & model == "historical" & anomalies < 0), aes(x=lon, y = lat,  fill = "dryer"))+
    geom_point(data = subset(cmip6s, period == i & model == "historical" & plus.AI > 10), aes(x=lon, y = lat), shape = ".")+
    borders(colour = "grey60")+
    scale_fill_manual(values =  c("#F46D43", "#74ADD1"), na.translate = F)+
    labs(title = paste("AI changes between",i , "and 1970-2000", sep = " "), fill = "")+
    theme_void()+ylim(-55,90)+
    theme(legend.position = "bottom")
  map_list[[i]] <- g
}

ggarrange(plotlist = map_list, ncol = 2)

```

```{r map-future-changes-strict, fig.dim = c(10,10)}

map_list <- list()

for(i in c("SSP245","SSP370","SSP585")){
  for (j in c("2030_2060","2070_2100")){
    index <- paste(i, j, sep = " , ")
    g <- ggplot() + geom_tile(data = subset(cmip6s, period == j & model == i & anomalies > 0), aes(x=lon, y = lat,  fill = "wetter"))+
      geom_tile(data = subset(cmip6s, period == j & model == i & anomalies < 0), aes(x=lon, y = lat,  fill = "dryer"))+
       geom_point(data = subset(cmip6s, period == j & model == i & anomalies > 0 & plus.AI > 10), aes(x=lon, y = lat), shape = "+", col = "grey40", size = 0.1)+
       geom_point(data = subset(cmip6s, period == j & model == i & anomalies < 0 & minus.AI > 10), aes(x=lon, y = lat), shape = "+", col = "grey40", size = 0.1)+
      borders(colour = "grey60")+
      scale_fill_manual(values = c("#F46D43", "#74ADD1"), na.translate = F)+
      labs(fill = "Compared to 1970-2000", title = i)+
      theme_void()+ylim(-55,90)+
      theme(legend.position = "bottom")
    map_list[[index]] <- g
  }}



ggpubr::ggarrange(plotlist = map_list,ncol = 2, nrow = 3, common.legend = T, legend = "bottom")
```

## Wetter or dryer, binned (reference period: 1970-2000)

```{r summary-of-AI-changes}
ano.quantiles <- cmip6s %>% group_by(model, period) %>% summarise(q10 = quantile(anomalies, probs = 0.1, na.rm =T), 
                                                                  q25 = quantile(anomalies, probs = 0.25, na.rm =T), 
                                                                  q50 = quantile(anomalies, probs = 0.5, na.rm =T), 
                                                                  q75 = quantile(anomalies, probs = 0.75, na.rm =T), 
                                                                  q90 = quantile(anomalies, probs = 0.9, na.rm =T)) 

ano.quantiles <- cmip6s %>% summarise(q10 = quantile(anomalies, probs = 0.1, na.rm =T), 
                                      q25 = quantile(anomalies, probs = 0.25, na.rm =T), 
                                      q50 = quantile(anomalies, probs = 0.5, na.rm =T), 
                                      q75 = quantile(anomalies, probs = 0.75, na.rm =T), 
                                      q90 = quantile(anomalies, probs = 0.9, na.rm =T)) 
qbreaks <- c(-5, -1,-0.1, -0.01, 0, 0.01, 0.1, 1,5)
colscale <- c("#08519c", "#3182bd", "#6baed6", "#9ecae1", "#c6dbef", "#fddbc7", "#f4a582", "#d6604d", "#b2182b", "#67001f")
```

"Anomalies" is the multimodel AI average minus the reference AI (period 1970-2000). More blue: AI mean is superior to AI ref, hence wetter than the ref. More red: the anomalie is negative, hence AI mean is inferior to AI ref, hence dryer. 

AS OF NOW THE SAME QUANTILES ARE USED FOR ALL MAPS.

```{r map-historical-changes-quantiles, fig.dim = c(8,3)}
map_list <- list()

for(i in c("1850_1880","1985_2015")){
  #quant = subset(ano.quantiles, model == "historical" & period == i)
  #b = quant[1,c(3:7)] %>% as.numeric
  b = qbreaks
  g <- ggplot() + geom_raster(data = subset(cmip6s, period == i & model == "historical"), aes(x=lon, y = lat,  fill = anomalies))+
    borders(colour = "grey60")+
    binned_scale(aesthetics = "fill", breaks = b, palette = function(x) rev(colscale),
                 guide = guide_legend(label.theme = element_text(angle = 0)))+
    labs(title = i, fill = "")+
    theme_void()+ylim(-55,90)+
    theme(legend.position = "right")
  map_list[[i]] <- g
}

ggpubr::ggarrange(plotlist = map_list, ncol = 2, common.legend = T, legend = "bottom")

```

```{r map-future-changes-quantiles, fig.dim = c(10,10)}

map_list <- list()

for(i in c("SSP245","SSP370","SSP585")){
  for (j in c("2030_2060","2070_2100")){
    index <- paste(i, j, sep = " , ")
    b = qbreaks
    g <- ggplot() + geom_raster(data = subset(cmip6s, period == j & model == i), aes(x=lon, y = lat,  fill = anomalies))+
      borders(colour = "grey60")+
      binned_scale(aesthetics = "fill", breaks = b, palette = function(x) rev(colscale),
                   guide = guide_legend(label.theme = element_text(angle = 0)))+
      labs(fill = "Compared to 1970-2000", title = index)+
      theme_void()+ylim(-55,90)+
      theme(legend.position = "bottom")
    map_list[[index]] <- g
  }}



ggpubr::ggarrange(plotlist = map_list,ncol = 2, nrow = 3, common.legend = T, legend = "bottom")
```

## Wetter or dryer, binned, % (reference period: 1970-2000)

```{r summary-of-AI-changes-percent}
cmip6s$anomalies.percent <- with(cmip6s, (AI.mean-AI.ref)/AI.ref*100)


pbreaks <- c(-40, -30,-20, -10, 0, 10, 20, 30, 40)
colscale <- c("#08519c", "#3182bd", "#6baed6", "#9ecae1", "#c6dbef", "#fddbc7", "#f4a582", "#d6604d", "#b2182b", "#67001f")
```

"Anomalies" is the multimodel AI average minus the reference AI (period 1970-2000). More blue: AI mean is superior to AI ref, hence wetter than the ref. More red: the anomalie is negative, hence AI mean is inferior to AI ref, hence dryer. 

```{r map-historical-changes-percent, fig.dim = c(5,8)}
map_list <- list()

for(i in c("1850_1880","1985_2015")){
  #quant = subset(ano.quantiles, model == "historical" & period == i)
  #b = quant[1,c(3:7)] %>% as.numeric
  b = pbreaks
  g <- ggplot() + geom_raster(data = subset(cmip6s, period == i & model == "historical"), aes(x=lon, y = lat,  fill = anomalies.percent))+
    borders(colour = "grey60")+
    binned_scale(aesthetics = "fill", breaks = b, palette = function(x) rev(colscale),
                 guide = guide_legend(label.theme = element_text(angle = 0)))+
    labs(title = i, fill = "")+
    theme_void()+ylim(-55,90)+
    theme(legend.position = "right")
  map_list[[i]] <- g
}

ggpubr::ggarrange(plotlist = map_list, ncol = 1, nrow = 2, common.legend = T, legend = "bottom") %>% annotate_figure(top = "% change of AI index compared to 1970-2000")
```

```{r map-future-changes-percent, fig.dim = c(10,10)}

map_list <- list()

for(i in c("SSP245","SSP370","SSP585")){
  for (j in c("2030_2060","2070_2100")){
    index <- paste(i, j, sep = " , ")
    b = pbreaks
    g <- ggplot() + geom_tile(data = subset(cmip6s, period == j & model == i), aes(x=lon, y = lat,  fill = anomalies.percent))+
      borders(colour = "grey60")+
      binned_scale(aesthetics = "fill", breaks = b, palette = function(x) rev(colscale),
                   guide = guide_legend(label.theme = element_text(angle = 0)))+
      labs(fill = "% change of AI index\ncompared to 1970-2000", title = index)+
      theme_void()+ylim(-55,90)+
      theme(legend.position = "bottom")
    map_list[[index]] <- g
  }}


ggpubr::ggarrange(plotlist = map_list,ncol = 2, nrow = 3, common.legend = T, legend = "bottom")
```


# Difference between future and reference category of AI, using cat maj

## Strictly dryer or wetter category cat maj

```{r map-historical-changes-dryerwetter-catmaj-2, fig.dim = c(8, 4)}

cmip6s$change.catmaj <- with(cmip6s, ifelse(cat.maj == catmajref, 0, 1))

map_list <- list()

for(i in c("1850_1880","1985_2015")){
  g <- ggplot() + geom_raster(data = subset(cmip6s, period == i & model == "historical" & change.catmaj == 1), aes(x=lon, y = lat,  fill = dryer.maj))+
    borders(colour = "grey60")+
    scale_fill_manual(values =  c("#F46D43", "#74ADD1"), na.translate = F)+
    labs(title = paste("AI changes between", i , "and 1970-2000", sep = " "), fill = "")+
    theme_void()+ylim(-55,90)+
    theme(legend.position = "bottom")
  map_list[[i]] <- g
}

ggarrange(plotlist = map_list, ncol = 2, common.legend = T)

```

```{r map-future-changes-dryerwetter-catmaj, fig.dim = c(10,10)}

map_list <- list()

for(i in c("SSP245","SSP370","SSP585")){
  for (j in c("2030_2060","2070_2100")){
    index <- paste(i, j, sep = " , ")
    g <- ggplot() + geom_raster(data = subset(cmip6s, period == j & model == i), aes(x=lon, y = lat,  fill = dryer.maj))+
      geom_raster(data = subset(cmip6s, period == j & model == "historical" & change.catmaj == 1 & nb.maj > 10), aes(x= lon, y = lat, fill = "'"))+
      borders(colour = "grey60")+
      scale_fill_manual(values = c("#F46D43", "#74ADD1"), na.translate = F)+
      labs(fill = "Compared to 1970-2000", title = index)+
      theme_void()+ylim(-55,90)+
      theme(legend.position = "bottom")
    map_list[[index]] <- g
  }}



ggpubr::ggarrange(plotlist = map_list,ncol = 2, nrow = 3, common.legend = T, legend = "bottom")
```

## Detailed differences cat maj

### Detailed changes and colors cat maj

```{r detailed-cat-change-catmaj}
cmip6s$change.catmaj <- with(cmip6s, ifelse(cat.maj == catmajref, 0, 1))

cmip6s$change.cat.maj <- with(cmip6s, paste(catmajref, cat.maj, sep = " to "))

cmip6s$change.cat.maj <- with(cmip6s, ifelse(catmajref == "Arid" & cat.maj %in% c("Cold","Humid"),"Arid to cold or humid",
         ifelse(catmajref == "Arid" & cat.maj %in% c("Semi-Arid","Dry subhumid"), "Arid to semi-arid or dry subhumid",
         ifelse(catmajref == "Arid" & cat.maj == "Hyperarid", "Arid to hyperarid",
         ifelse(catmajref == "Semi-arid" & cat.maj == "Dry subhumid", "Semi-arid to dry subhumid",
         ifelse(catmajref == "Semi-arid" & cat.maj == "Arid", "Semi-arid to arid",
         ifelse(catmajref == "Semi-arid" & cat.maj == "Hyperarid", "Semi-arid to hyperarid",
         ifelse(catmajref == "Semi-arid" & cat.maj %in% c("Humid","Cold"), "Semi-arid to humid or cold",
         ifelse(catmajref == "Dry subhumid" & cat.maj %in% c("Semi-arid","Arid","Hyperarid"), "Dry subhumid to arid, semi-arid or hyperarid",
         ifelse(catmajref == "Dry subhumid" & cat.maj %in% c("Humid","Cold"), "Dry subhumid to humid or cold",
         ifelse(catmajref == "Humid" & cat.maj == "Cold","Humid to cold",
         ifelse(catmajref == "Humid" & cat.maj %in% c("Dry subhumid", "Semi-arid"), "Humid to dry-subhumid or semi-arid",
         ifelse(catmajref == "Humid" & cat.maj %in% c("Arid", "Hyperarid"), "Humid to arid or hyperarid",
         ifelse(catmajref == "Cold" & cat.maj == "Humid", "Cold to humid", 
         ifelse(catmajref == "Cold" & cat.maj %in% c("Dry subhumid","Semi-arid","Arid","Hyperarid"), "Cold to dryland",
         ifelse(catmajref == "Hyperarid" & cat.maj == "Arid", "Hyperarid to arid",
         ifelse(catmajref == "Hyperarid" & cat.maj %in% c("Semi-arid","Dry-subhumid"),"Hyperarid to semi arid or dry subhumid",
         ifelse(catmajref == "Hyperarid" & cat.maj %in% c("Humid", "Cold"), "Hyperarid to humid or cold", NA))))))))))))))))))

cmip6s$change.cat.maj <- factor(cmip6s$change.cat.maj, 
                                levels = c("Arid to hyperarid","Semi-arid to arid", "Dry subhumid to arid, semi-arid or hyperarid", "Humid to dry-subhumid or semi-arid", "Cold to humid","Hyperarid to arid", "Arid to semi-arid or dry subhumid", "Dry subhumid to humid or cold", "Semi-arid to dry subhumid", "Semi-arid to humid or cold"))
table(cmip6s$change.cat.maj)

colors_changes <- c("Hyperarid to arid" = "#E7D7B6", 
                    "Arid to semi-arid or dry subhumid" = colorvec[7],
                    "Dry subhumid to humid or cold" = colorvec[8],
                    "Semi-arid to dry subhumid" = colorvec[9],
                    "Semi-arid to humid or cold" = colorvec[10],
                    "Humid to cold" = colorvec[11],
                    "Cold to humid" = "#E7D7B6", 
                    "Humid to dry-subhumid or semi-arid" = colorvec[5],
                    "Dry subhumid to arid, semi-arid or hyperarid" = colorvec[4],
                    "Semi-arid to arid" = colorvec[3],
                    "Arid to hyperarid" = colorvec[1])
```

### Changes towards dryer categories  cat maj

#### Historical

```{r map-historical-changes-dry-catmaj, fig.dim = c(8, 4)}
map_list <- list()

for(i in c("1850_1880","1985_2015")){
  g <- ggplot() + 
    geom_raster(data = subset(cmip6s, period == i & model == "historical" & change.catmaj == 1 & dryer.maj == "dryer"), 
                aes(x=lon, y = lat,  fill = change.cat.maj))+
    borders(colour = "grey60")+
    scale_fill_manual(values =  colors_changes, na.translate = F)+
    labs(title = i, fill = "Change of category\ncompared to\n1970-2000")+
    theme_void()+ylim(-55,90)+
    theme(legend.position = "bottom")+guides(fill = guide_legend(nrow = 2))
  map_list[[i]] <- g
}


ggarrange(plotlist = map_list, ncol = 2, common.legend = T, legend = "bottom")

```

##### Future

```{r map-future-changes-dry-catmaj, fig.dim = c(10,10)}

map_list <- list()

for(i in c("SSP245","SSP370","SSP585")){
  for (j in c("2030_2060","2070_2100")){
    index <- paste(i, j, sep = " , ")
    g <- ggplot() + geom_tile(data = subset(cmip6s, period == j & model == i & change.catmaj == 1 & dryer.maj == "dryer"), 
                              aes(x=lon, y = lat,  fill = change.cat.maj))+
      #   geom_raster(data = subset(cmip6s, period == i & model == "historical" & change.catmaj == 1 & nb.maj > 10), aes(x= lon, y = lat, fill = "'"))+
      borders(colour = "grey60")+
      scale_fill_manual(values = colors_changes, na.translate = F)+
      #  scale_fill_viridis_d()+
      labs(fill = "Change of category\ncompared to\n1970-2000", title = index)+
      theme_void()+ylim(-55,90)+
      theme(legend.position = "bottom")+guides(fill = guide_legend(nrow = 2))
    map_list[[index]] <- g
  }}

ggarrange(plotlist = map_list,ncol = 2, nrow = 3, common.legend = T, legend = "bottom")
```

### Changes towards wetter categories  cat maj

#### Historical

```{r map-historical-changes-wet-catmaj, fig.dim = c(8, 4)}
map_list <- list()

for(i in c("1850_1880","1985_2015")){
  g <- ggplot() + geom_raster(data = subset(cmip6s, period == i & model == "historical" & change.catmaj == 1 & dryer.maj == "wetter"), 
                              aes(x=lon, y = lat,  fill = change.cat.maj))+
    borders(colour = "grey60", ylim = c(-40,40))+
    scale_fill_manual(values =  colors_changes, na.translate = F)+
    labs(title = i, fill = "Change of category\ncompared to\n1970-2000")+
    theme_void()+
    ylim(-50,50)+
    theme(legend.position = "bottom")+guides(fill = guide_legend(nrow = 2))
  map_list[[i]] <- g
}

ggarrange(plotlist = map_list, ncol = 2, common.legend = T, legend = "bottom")

```

#### Future

```{r map-future-changes-wet-catmaj, fig.dim = c(16,8)}

map_list <- list()

for(i in c("SSP245","SSP370","SSP585")){
  for (j in c("2030_2060","2070_2100")){
    index <- paste(i, j, sep = " , ")
    g <- ggplot() + geom_tile(data = subset(cmip6s, period == j & model == i & change.catmaj == 1 & dryer.maj == "wetter"), 
                              aes(x=lon, y = lat,  fill = change.cat.maj))+
      borders(colour = "grey60", ylim = c(-40,40))+
      scale_fill_manual(values = colors_changes, na.translate = F)+
      labs(fill = "Change of category\ncompared to\n1970-2000", title = index)+
      theme_void()+ylim(-50,50)+
      theme(legend.position = "bottom")+guides(fill = guide_legend(nrow = 2))
    map_list[[index]] <- g
  }}

ggpubr::ggarrange(plotlist = map_list, nrow = 3, ncol = 2, common.legend = T, legend = "bottom")
```

# Difference between future and reference category of AI, using cat mean

## Stricly dryer or wetter category cat mean

```{r map-historical-changes-dryerwetter-catmean, fig.dim = c(8, 4)}

cmip6s$change.catmean <- with(cmip6s, ifelse(cat.maj == catmajref, 0, 1))

map_list <- list()

for(i in c("1850_1880","1985_2015")){
  g <- ggplot() + geom_raster(data = subset(cmip6s, period == i & model == "historical" & change.catmean == 1), aes(x=lon, y = lat,  fill = dryer))+
    borders(colour = "grey60")+
    scale_fill_manual(values =  c("#F46D43", "#74ADD1"), na.translate = F)+
    labs(title = paste("AI changes between",i , "and 1970-2000", sep = " "), fill = "")+
    theme_void()+ylim(-55,90)+
    theme(legend.position = "bottom")
  map_list[[i]] <- g
}

ggarrange(plotlist = map_list, ncol = 2, common.legend = T, legend = "bottom")

```

```{r map-future-changes-dryerwetter-catmean, fig.dim = c(10,10)}

map_list <- list()

for(i in c("SSP245","SSP370","SSP585")){
  for (j in c("2030_2060","2070_2100")){
    index <- paste(i, j, sep = " , ")
    g <- ggplot() + geom_raster(data = subset(cmip6s, period == j & model == i  & change.catmean == 1), aes(x=lon, y = lat,  fill = dryer))+
      borders(colour = "grey60")+
      scale_fill_manual(values = c("#F46D43", "#74ADD1"), na.translate = F)+
      labs(fill = "Compared to 1970-2000", title = index)+
      theme_void()+ylim(-55,90)+
      theme(legend.position = "bottom")
    map_list[[index]] <- g
  }}

ggpubr::ggarrange(plotlist = map_list,ncol = 2, nrow = 3, common.legend = T, legend = "bottom")
```

## Detailed differences cat mean

### Detailed changes and colors 

```{r detailed-cat-mean-change}
cmip6s$change.catmean <- with(cmip6s, ifelse(cat.AI == catAIref, 0, 1))

cmip6s$change.cat.mean <- with(cmip6s, paste(catAIref, cat.AI, sep = " to "))
unique(cmip6s$change.cat)

# cmip6s$change.cat.mean <- with(cmip6s, 
#       ifelse(catAIref == "Arid" & cat.AI %in% c("Cold","Humid"),"Arid to cold or humid",
#       ifelse(catAIref == "Arid" & cat.AI %in% c("Semi-Arid","Dry subhumid"), "Arid to semi-arid or dry subhumid",
#       ifelse(catAIref == "Arid" & cat.AI == "Hyperarid", "Arid to hyperarid",
#       ifelse(catAIref == "Semi-arid" & cat.AI == "Dry subhumid", "Semi-arid to dry subhumid",
#       ifelse(catAIref == "Semi-arid" & cat.AI == "Arid", "Semi-arid to arid",
#       ifelse(catAIref == "Semi-arid" & cat.AI == "Hyperarid", "Semi-arid to hyperarid",
#       ifelse(catAIref == "Semi-arid" & cat.AI %in% c("Humid","Cold"), "Semi-arid to humid or cold",
#       ifelse(catAIref == "Dry subhumid" & cat.AI %in% c("Semi-arid","Arid","Hyperarid"), "Dry subhumid to arid, semi-arid or hyperarid",
#       ifelse(catAIref == "Dry subhumid" & cat.AI %in% c("Humid","Cold"), "Dry subhumid to humid or cold",
#       ifelse(catAIref == "Humid" & cat.AI == "Cold","Humid to cold",
#       ifelse(catAIref == "Humid" & cat.AI %in% c("Dry subhumid","Semi-arid"), "Humid to dry-subhumid or semi-arid",
#       ifelse(catAIref == "Humid" & cat.AI %in% c("Arid", "Hyperarid"), "Humid to arid or hyperarid",
#       ifelse(catAIref == "Cold" & cat.AI == "Humid", "Cold to humid", 
#       ifelse(catAIref == "Cold" & cat.AI %in% c("Dry subhumid","Semi-arid","Arid","Hyperarid"), "Cold to dryland",
#       ifelse(catAIref == "Hyperarid" & cat.AI == "Arid", "Hyperarid to arid",
#       ifelse(catAIref == "Hyperarid" & cat.AI %in% c("Semi-arid","Dry-subhumid"),"Hyperarid to semi arid or dry subhumid",
#       ifelse(catAIref == "Hyperarid" & cat.AI %in% c("Humid", "Cold"), "Hyperarid to humid or cold", 
#       NA))))))))))))))))))

table(cmip6s$change.cat.mean)
cmip6s$change.cat.mean <- factor(cmip6s$change.cat.mean, 
                                  levels = c("Arid to Hyperarid","Semi-arid to Arid", "Dry subhumid to Semi-arid", "Humid to Semi-arid", "Humid to Dry subhumid", "Cold to Humid","Hyperarid to arid", "Arid to Semi-arid", "Dry subhumid to Humid", "Semi-arid to Dry subhumid", "Semi-arid to Humid"))

colors_changes <- c("Hyperarid to arid" = "#E7D7B6", 
                     "Arid to Semi-arid" = colorvec[7],
                     "Dry subhumid to Humid" = colorvec[8],
                     "Semi-arid to Dry subhumid" = colorvec[9],
                     "Semi-arid to Humid" = colorvec[10],
                     "Humid to Cold" = colorvec[11],
                     "Cold to Humid" = "#E7D7B6",
                     "Humid to Dry subhumid" = colorvec[6],
                     "Humid to Semi-arid" = colorvec[5],
                     "Dry subhumid to Semi-arid" = colorvec[4],
                     "Semi-arid to arid" = colorvec[3],
                     "Arid to hyperarid" = colorvec[1])
```
### Changes towards dryer categories cat mean
#### Historical

```{r map-historical-changes-dry-catmean, fig.dim = c(8, 4)}
map_list <- list()

for(i in c("1850_1880","1985_2015")){
  g <- ggplot() + 
    geom_raster(data = subset(cmip6s, period == i & model == "historical" & change.catmean == 1 & dryer == "dryer"), 
                aes(x=lon, y = lat,  fill = change.cat.mean))+
    borders(colour = "grey60")+
    scale_fill_manual(values =  colorvec, na.translate = F)+
    labs(title = i, fill = "Change of category\ncompared to\n1970-2000")+
    theme_void()+ylim(-55,90)+
    theme(legend.position = "bottom")+guides(fill = guide_legend(nrow = 2))
  map_list[[i]] <- g
}

ggarrange(plotlist = map_list, ncol = 2, common.legend = T, legend = "bottom")
```

#### Future

```{r map-future-changes-dry-catmean, fig.dim = c(10,10)}

map_list <- list()

for(i in c("SSP245","SSP370","SSP585")){
  for (j in c("2030_2060","2070_2100")){
    index <- paste(i, j, sep = " , ")
    g <- ggplot() + geom_tile(data = subset(cmip6s, period == j & model == i & change.catmean == 1 & dryer == "dryer"), 
                              aes(x=lon, y = lat,  fill = change.cat.mean))+
      borders(colour = "grey60")+
      scale_fill_manual(values = colorvec)+
      labs(fill = "Change of category\ncompared to\n1970-2000", title = index)+
      theme_void()+ylim(-55,90)+
      theme(legend.position = "bottom")+guides(fill = guide_legend(nrow = 2))
    map_list[[index]] <- g
  }}

ggarrange(plotlist = map_list, ncol = 2, nrow = 3, common.legend = T, legend = "bottom")
```

### Changes towards wetter categories cat mean

#### Historical

```{r map-historical-changes-wet-, fig.dim = c(8, 3)}
map_list <- list()

for(i in c("1850_1880","1985_2015")){
  g <- ggplot() + geom_tile(data = subset(cmip6s, period == i & model == "historical" & change.catmaj == 1 & dryer == "wetter"), 
                            aes(x=lon, y = lat,  fill = change.cat.mean))+
    borders(colour = "grey60", ylim = c(-40,40))+
    scale_fill_manual(values =  colors_changes, na.translate = F)+
    labs(title = i, fill = "Change of category\ncompared to\n1970-2000")+
    theme_void()+ylim(-50,50)+
    theme(legend.position = "bottom")+guides(fill = guide_legend(nrow = 2))
  map_list[[i]] <- g
}

ggarrange(plotlist = map_list, ncol = 2, common.legend = T, legend = "bottom")

```

#### Future

```{r map-future-changes-wet-catmean, fig.dim = c(10,8)}

map_list <- list()

for(i in c("SSP245","SSP370","SSP585")){
  for (j in c("2030_2060","2070_2100")){
    index <- paste(i, j, sep = " , ")
    g <- ggplot() + geom_tile(data = subset(cmip6s, period == j & model == i & change.catmean == 1 & dryer == "wetter"), 
                              aes(x=lon, y = lat,  fill = change.cat.mean))+
      borders(colour = "grey60", ylim = c(-40,40))+
      scale_fill_manual(values = colors_changes, na.translate = F)+
      labs(fill = "Change of category\ncompared to\n1970-2000", title = index)+
      theme_void()+
      theme(legend.position = "bottom")+guides(fill = guide_legend(nrow = 2))
    map_list[[index]] <- g
  }}

ggpubr::ggarrange(plotlist = map_list,ncol = 2, nrow = 3, common.legend = T, legend = "bottom")
```



# Tables

## Changes of aridity category using cat mean

```{r changes-in-table}
tab <- cmip6s %>% group_by(model, period, Continent) %>% summarise(Cold = table(cat.AI)[1], Hyperarid = table(cat.AI)[2], Arid = table(cat.AI)[3], Semiarid = table(cat.AI)[4], Drysubhumid = table(cat.AI)[5], Humid = table(cat.AI)[6], total = n()) 
write.table(tab, "tab.catAI.txt")

tab.future <- cmip6s %>% subset(!Continent %in% c("SOUTHERN","PACIFIC","ATLANTIC","INDIAN","ARCTIC")) %>%
  group_by(period, model, cat.AI) %>%
  summarise(count = n()) %>%
  ungroup() %>% group_by(period, model) %>% mutate(percent = round(count/sum(count)*100, 1), count = NULL) %>%
  reshape2::dcast(period + model ~cat.AI) %>%
  mutate("Sum drylands" = rowSums(.[c("Hyperarid","Arid","Semi-arid","Dry subhumid")])) %>%
  select(c("model","period","Hyperarid","Arid","Semi-arid","Dry subhumid","Sum drylands","Humid","Cold","NA"))

kable(tab.future[order(tab.future$model, decreasing = F),]) %>% kable_styling(bootstrap_options = "bordered") %>%
  column_spec(c(7,10), italic = T, include_thead = T) %>% row_spec(2, bold = T) 

# ggplot(tab.future, aes(x = period))+
#   geom_point(aes(y = Hyperarid, col = "Hyperarid", shape = model))+
#   geom_line(aes(y=Hyperarid, col = "Hyperarid", group = model, lty = model))+
#   geom_point(aes(y = Arid, col = "Arid", shape = model))+
#   geom_line(aes(y=Arid, col = "Arid", group = model, lty = model))+
#   geom_point(aes(y = get('Semi-arid'), col = "Semi-arid", shape = model))+
#   geom_line(aes(y=get('Semi-arid'), col = "Semi-arid", group = model, lty = model))+
#   geom_point(aes(y = get('Dry subhumid'), col = "Dry subhumid", shape = model))+
#   geom_line(aes(y=get('Dry subhumid'), col = "Dry subhumid", group = model, lty = model))+
#   geom_point(aes(y = Humid, col = "Humid", shape = model))+
#   geom_line(aes(y= Humid, col = "Humid", group = model, lty = model))+
#   scale_color_manual(values = col.cat)+
#   theme_minimal()
```

```{r tab-percent-catmean, eval = F}

cmip6$cat.AI <- factor(cmip6$cat.AI, levels = c("Hyperarid", "Arid", "Semi-arid", "Dry subhumid", "Humid", "Cold"))
tab.percent <- cmip6 %>% subset(!Continent %in% c("SOUTHERN","PACIFIC","POLAR","ATLANTIC","INDIAN","ARCTIC")) %>%
  group_by(period, model, cat.AI, source) %>%
  summarise(count = n()) %>%
  ungroup() %>% 
  group_by(period, model, cat.AI) %>% 
  summarise(mmmean = mean(count, na.rm = T), mmsd = sd(count, na.rm = T)) %>%
  mutate(percent = round(mmmean/sum(mmmean)*100, 1), sd.percent = round(mmsd/sum(mmmean)*100,1)) 
write.table(tab.percent, "tab.percent.txt")
```

```{r colplots-futurecatmean, fig.dim = c(10,5)}
tab.percent <- read.table("tab.percent.txt")

df245 <- rbind(subset(tab.percent, model == "historical" & period %in% c("1850_1880", "1970_2000")), subset(tab.percent, model == "SSP245")) %>%
  group_by(period, model) %>% 
  mutate(lab.y = (rev(cumsum(rev(percent)))) - percent*0.5) %>% 
  subset(cat.AI %in% c("Hyperarid", "Arid", "Semi-arid", "Dry subhumid", "Humid","Cold"))

df245$cat.AI <- factor(df245$cat.AI, levels = c("Hyperarid", "Arid", "Semi-arid", "Dry subhumid", "Humid", "Cold"))

b245 <- ggplot(data = df245, aes(x = period, y = percent))+
  geom_col(aes(group = cat.AI, col = cat.AI, fill = cat.AI))+
  geom_label(aes(y = lab.y, label = paste(percent, "  ", sd.percent, sep = "")), size = 3)+
  scale_color_manual(values = col.cat, aesthetics = c("col", "fill"), na.translate = F)+
  labs(title = "SSP 2-4.5", y = "%", x = "", col = "", fill = "")+
  theme_minimal()

df370 <- rbind(subset(tab.percent, model == "historical" & period %in% c("1850_1880", "1970_2000")), subset(tab.percent, model == "SSP370")) %>%
  group_by(period, model) %>% 
  mutate(lab.y = (rev(cumsum(rev(percent)))) - percent*0.5) %>% 
  subset(cat.AI %in% c("Hyperarid", "Arid", "Semi-arid", "Dry subhumid", "Humid","Cold"))

df370$cat.AI <- factor(df370$cat.AI, levels = c("Hyperarid", "Arid", "Semi-arid", "Dry subhumid", "Humid", "Cold"))

b370 <- ggplot(data = df370, aes(x = period, y = percent))+
  geom_col(aes(group = cat.AI, col = cat.AI, fill = cat.AI))+
  geom_label(aes(y = lab.y, label = paste(percent, "  ", sd.percent, sep = "")), size = 3)+
  scale_color_manual(values = col.cat, aesthetics = c("col", "fill"), na.translate = F)+
  labs(title = "SSP 3-7.0", y = "%", x = "", col = "", fill = "")+
  theme_minimal()


df585 <- rbind(subset(tab.percent, model == "historical" & period %in% c("1850_1880", "1970_2000")), subset(tab.percent, model == "SSP585")) %>%
  group_by(period, model) %>% 
  mutate(lab.y = (rev(cumsum(rev(percent)))) - percent*0.5) %>% 
  subset(cat.AI %in% c("Hyperarid", "Arid", "Semi-arid", "Dry subhumid", "Humid","Cold"))

df585$cat.AI <- factor(df585$cat.AI, levels = c("Hyperarid", "Arid", "Semi-arid", "Dry subhumid", "Humid", "Cold"))

b585 <- ggplot(data = df585, aes(x = period, y = percent))+
  geom_col(aes(group = cat.AI, col = cat.AI, fill = cat.AI))+
  geom_label(aes(y = lab.y, label = paste(percent, "  ", sd.percent, sep = "")), size = 3)+
  scale_color_manual(values = col.cat, aesthetics = c("col", "fill"), na.translate = F)+
  labs(title = "SSP 5-8.5", y = "%", x = "", col = "", fill = "")+
  theme_minimal()


ggarrange(plotlist = list(b245, b370, b585), common.legend = T, legend = "bottom", ncol = 3)
```



## Changes of aridity category using cat maj

```{r changes-in-table-maj}
tab.maj <- cmip6s %>% group_by(model, period, Continent) %>% summarise(Cold = table(cat.maj)[1], Hyperarid = table(cat.maj)[2], Arid = table(cat.maj)[3], Semiarid = table(cat.maj)[4], Drysubhumid = table(cat.maj)[5], Humid = table(cat.maj)[6], total = n()) 
write.table(tab.maj, "tab.catmaj.txt")

tab.future.maj <- cmip6s %>% subset(!Continent %in% c("SOUTHERN","PACIFIC","ATLANTIC","INDIAN","ARCTIC")) %>%
  group_by(period, model, cat.maj) %>%
  summarise(count = n()) %>%
  ungroup() %>% group_by(period, model) %>% mutate(percent = round(count/sum(count)*100, 1), count = NULL) %>%
  reshape2::dcast(period + model ~cat.maj) %>%
  mutate("Sum drylands" = rowSums(.[c("Hyperarid","Arid","Semi-arid","Dry subhumid")])) %>%
  select(c("model","period","Hyperarid","Arid","Semi-arid","Dry subhumid","Sum drylands","Humid","Cold","NA"))

kable(tab.future.maj[order(tab.future.maj$model, decreasing = F),]) %>% kable_styling(bootstrap_options = "bordered") %>%
  column_spec(c(7,10), italic = T, include_thead = T) %>% row_spec(2, bold = T) 
```

```{r colplots-futurecat, fig.dim = c(10,5)}
tab.flow.catmaj <- cmip6s %>% subset(!Continent %in% c("SOUTHERN","PACIFIC","POLAR","ATLANTIC","INDIAN","ARCTIC")) %>%
  group_by(period, model, cat.maj) %>%
  summarise(count = n()) %>%
  ungroup() %>% group_by(period, model) %>% mutate(percent = round(count/sum(count)*100, 1), count = NULL)

df245 <- rbind(subset(tab.flow.catmaj, model == "historical" & period %in% c("1850_1880", "1970_2000")), subset(tab.flow.catmaj, model == "SSP245")) %>%
  mutate(lab.y = (rev(cumsum(rev(percent)))) - percent*0.5) %>% 
  subset(cat.maj %in% c("Hyperarid", "Arid", "Semi-arid", "Dry subhumid", "Humid","Cold"))

b245 <- ggplot(data = df245, aes(x = period, y = percent))+
  geom_col(aes(group = cat.maj, col = cat.maj, fill = cat.maj))+
  geom_label(aes(y = lab.y, label = percent))+
  scale_color_manual(values = col.cat, aesthetics = c("col", "fill"), na.translate = F)+
  labs(title = "SSP 2-4.5", y = "%", x = "")+
  theme_minimal()

df370 <- rbind(subset(tab.flow.catmaj, model == "historical" & period %in% c("1850_1880", "1970_2000")), subset(tab.flow.catmaj, model == "SSP370")) %>% mutate(lab.y = (rev(cumsum(rev(percent)))) - percent*0.5) %>% 
  subset(cat.maj %in% c("Hyperarid", "Arid", "Semi-arid", "Dry subhumid", "Humid","Cold"))

b370 <- ggplot(data = df370, aes(x = period, y = percent))+
  geom_col(aes(group = cat.maj, col = cat.maj, fill = cat.maj))+
  geom_label(aes(y = lab.y, label = percent))+
  scale_color_manual(values = col.cat, aesthetics = c("col", "fill"), na.translate = F)+
  labs(title = "SSP 3-7.0", y = "%", x = "")+
  theme_minimal()


df585 <- rbind(subset(tab.flow.catmaj, model == "historical" & period %in% c("1850_1880", "1970_2000")), subset(tab.flow.catmaj, model == "SSP585")) %>% mutate(lab.y = (rev(cumsum(rev(percent)))) - percent*0.5) %>% 
  subset(cat.maj %in% c("Hyperarid", "Arid", "Semi-arid", "Dry subhumid", "Humid","Cold"))

b585 <- ggplot(data = df585, aes(x = period, y = percent))+
  geom_col(aes(group = cat.maj, col = cat.maj, fill = cat.maj))+
  geom_label(aes(y = lab.y, label = percent))+
  scale_color_manual(values = col.cat, aesthetics = c("col", "fill"), na.translate = F)+
  labs(title = "SSP 5-8.5", y = "%", x = "")+
  theme_minimal()


ggarrange(plotlist = list(b245, b370, b585), common.legend = T, legend = "bottom", ncol = 3)
```



## Changes by continent, cat mean

```{r tab-percent-continent-catmean, eval = F}

cmip6$cat.AI <- factor(cmip6$cat.AI, levels = c("Hyperarid", "Arid", "Semi-arid", "Dry subhumid", "Humid", "Cold"))

tab.percent.cont <- cmip6 %>% subset(!Continent %in% c("SOUTHERN","PACIFIC","POLAR","ATLANTIC","INDIAN","ARCTIC")) %>%
  group_by(period, model, Continent, cat.AI, source) %>%
  summarise(count = n()) %>%
  ungroup() %>% 
  group_by(period, model, Continent, cat.AI) %>% 
  summarise(mmmean = mean(count, na.rm = T), mmsd = sd(count, na.rm = T)) %>%
  mutate(percent = round(mmmean/sum(mmmean)*100, 1), sd.percent = round(mmsd/sum(mmmean)*100,1)) 
write.table(tab.percent.cont, "tab.percent.continent.txt")

tab.cont <- cmip6 %>% subset(!Continent %in% c("SOUTHERN","PACIFIC","POLAR","ATLANTIC","INDIAN","ARCTIC")) %>%
  group_by(period, model, Continent, cat.AI, source) %>%
  summarise(count = n()) %>%
  ungroup() 

list_ref <- list()

for(i in unique(tab.cont$source)){
  tab.cont.i <- filter(tab.cont, source == i)
  tab.cont.i.ref <- filter(tab.cont.i, period == "1970_2000")
  
  for(j in unique(tab.cont.i$Continent)){
    tab.cont.i.j <- filter(tab.cont.i, Continent == j)
    tab.cont.i.j.ref <- filter(tab.cont.i.ref, Continent == j)
    
    for(k in unique(tab.cont.i$cat.AI)){
      
      index <- paste(i, j, k, sep = "_")
      tab.cont.i.j.k <- filter(tab.cont.i.j, cat.AI == k)
      
      lines <- ifelse(dim(tab.cont.i.j.k)[1] > 0, dim(tab.cont.i.j.k)[1] - 1, 1)
      
      tab.cont.i.j.k.ref <- subset(tab.cont.i.j.ref, cat.AI == k)
      
      if(dim(tab.cont.i.j.k.ref)[1] == 0){
        names1 <- names(tab.cont.i.j.k.ref)
        line1 <- c("1970_2000","historical", i, j, k, 0)
        tab.cont.i.j.k.ref <- rbind(tab.cont.i.j.k.ref, line1) %>%
          setNames(names1)
        
      }else{
        tab.cont.i.j.k.ref <- tab.cont.i.j.k.ref
      }
      
      tab.cont.i.j.k.ref <- tab.cont.i.j.k.ref %>% rbind(tab.cont.i.j.k.ref[rep(1,lines),])
      tab.cont.i.j.k$diff <- tab.cont.i.j.k$count - as.numeric(tab.cont.i.j.k.ref$count)  
      
      list_ref[[index]] <- tab.cont.i.j.k
    }
  }
}


tab.cont.diff <- bind_rows(list_ref, .id = "column_label") %>%
  group_by(period, model, Continent, cat.AI) %>% 
  summarise(diff.mean = mean(diff, na.rm = T), diff.sd = sd(diff, na.rm = T))
write.table(tab.cont.diff, "tab.cont.diff.txt")

```

```{r changes-in-table-by-continent-mean, results = "asis"}
tab.percent.cont <- read.table("tab.percent.continent.txt")

tab_list_percent <- list()
tab_list_sd <- list()

for(i in unique(tab.percent.cont$Continent)){
  tab.i <- subset(tab.percent.cont, Continent == i)
  
  df.percent <- tab.i %>% reshape2::dcast(period + model ~ cat.AI, value.var = "percent")
  
  df.sd <- tab.i %>% reshape2::dcast(period + model ~ cat.AI, value.var = "sd.percent")
  # sd for a sum: square root of sum of squared sd
  
  drycats <- which(names(df.percent) %in% c("Hyperarid","Arid","Semi-arid","Dry subhumid"))
  
  df.percent <- df.percent %>% mutate("Sum drylands" = rowSums(.[drycats], na.rm = T))
  df.sd <- df.sd %>% mutate("Sum drylands" = round(sqrt(rowSums((.[drycats])^2, na.rm = T)),2))
  
  missing <- setdiff(c("model","period","Hyperarid","Arid","Semi-arid","Dry subhumid","Sum drylands","Humid","Cold","NA"), names(df.percent))
  
  df.percent[, missing] <- 0
  df.sd[,missing] <- 0
  
  df.percent <- df.percent %>% select(c("model","period","Hyperarid","Arid","Semi-arid","Dry subhumid","Sum drylands","Humid","Cold","NA"))
  df.percent[is.na(df.percent)] <- 0
  
  tab_list_percent[[i]] <- df.percent
  
  df.sd <- df.sd %>% select(c("model","period","Hyperarid","Arid","Semi-arid","Dry subhumid","Sum drylands","Humid","Cold","NA"))
  df.sd[is.na(df.sd)] <- 0
  
  tab_list_sd[[i]] <- df.sd
}

k_list <- list()
kdiff_list <- list()

for(i in names(tab_list_percent)){
  
  df <- tab_list_percent[[i]] %>% select(c("model", "period"))
  
  prop.ref <- tab_list_percent[[i]] %>% filter(period == "1970_2000")
  prop.ref <- prop.ref %>% rbind(prop.ref[rep(1,8),])
  df2 <- tab_list_percent[[i]][,c(3:10)] - prop.ref[,c(3:10)] %>% round(1)
  diff.prop.percent <- cbind(df, df2)
  
  sd.ref <- tab_list_sd[[i]] %>% filter(period == "1970_2000")
  sd.ref <- sd.ref %>% rbind(sd.ref[rep(1,8),])
  df3 <- sqrt(tab_list_sd[[i]][,c(3:10)]^2 + sd.ref[,c(3:10)]^2) %>% round(1)
  diff.sd.percent <- cbind(df, df3)
  
  df$Hyperarid <- paste(tab_list_percent[[i]]$Hyperarid, tab_list_sd[[i]]$Hyperarid, sep = "  ")
  df$Arid <- paste(tab_list_percent[[i]]$Arid, tab_list_sd[[i]]$Arid, sep = "  ")
  df$'Semi-Arid' <- paste(tab_list_percent[[i]][,5], tab_list_sd[[i]][,5], sep = "  ")
  df$'Dry subhumid' <- paste(tab_list_percent[[i]][,6], tab_list_sd[[i]][,6], sep = "  ")
  df$'Sum drylands' <- paste(tab_list_percent[[i]][,7], tab_list_sd[[i]][,7], sep = "  ")
  df$Humid <- paste(tab_list_percent[[i]]$Humid, tab_list_sd[[i]]$Humid, sep = "  ")
  df$Cold <- paste(tab_list_percent[[i]]$Cold, tab_list_sd[[i]]$Cold, sep = "  ")
  
  df.diff <- df
  df.diff$Hyperarid <- paste(round(diff.prop.percent$Hyperarid, 1), round(diff.sd.percent$Hyperarid, 1), sep = "  ")
  df.diff$Arid <- paste(round(diff.prop.percent$Arid, 1), round(diff.sd.percent$Arid, 1), sep = "  ")
  df.diff$'Semi-Arid' <- paste(round(diff.prop.percent[,5], 1), round(diff.sd.percent[,5], 1), sep = "  ")
  df.diff$'Dry subhumid' <- paste(round(diff.prop.percent[,6], 1), round(diff.sd.percent[,6], 1), sep = "  ")
  df.diff$'Sum drylands' <- paste(round(diff.prop.percent[,7],1), round(diff.sd.percent[,7], 1), sep = "  ")
  df.diff$Humid <- paste(round(diff.prop.percent$Humid, 1), round(diff.sd.percent$Humid, 1), sep = "  ")
  df.diff$Cold <- paste(round(diff.prop.percent$Cold, 1), round(diff.sd.percent$Cold, 1), sep = "  ")
  
  k_list[[i]] <- kable(df[order(df$model, decreasing = F),], caption = i) %>% kable_styling(bootstrap_options = "bordered") %>%
    column_spec(8, italic = T, background = colorvec[5]) %>% row_spec(2, bold = T) 
  kdiff_list[[i]] <- kable(df.diff[order(df.diff$model, decreasing = F),], caption = i) %>% kable_styling(bootstrap_options = "bordered") %>% 
    column_spec(8, italic = T, background = colorvec[5]) %>% row_spec(2, bold = T) 
  
}

```

### Proportion of aridity categories by continent

```{r prop-aridity-category, results = "asis"}
k_list
```

### Changes in proportion of aridity categories by continent

```{r prop-change-aridity-category, results = "asis"}
kdiff_list
```

### Changes in proportion of aridity categories, by continent, calculated by model

```{r changes-in-table-by-continent-diff, results = "asis"}
tab.cont.diff <- read.table("tab.cont.diff.txt")

tab_list_percent <- list()
tab_list_sd <- list()

for(i in unique(tab.cont.diff$Continent)){
  tab.i <- subset(tab.cont.diff, Continent == i)
  
  df.percent <- tab.i %>% reshape2::dcast(period + model ~ cat.AI, value.var = "percent")
  
  df.sd <- tab.i %>% reshape2::dcast(period + model ~ cat.AI, value.var = "sd.percent")
  # sd for a sum: square root of sum of squared sd
  
  drycats <- which(names(df.percent) %in% c("Hyperarid","Arid","Semi-arid","Dry subhumid"))
  
  df.percent <- df.percent %>% mutate("Sum drylands" = rowSums(.[drycats], na.rm = T))
  df.sd <- df.sd %>% mutate("Sum drylands" = round(sqrt(rowSums((.[drycats])^2, na.rm = T)),2))
  
  missing <- setdiff(c("model","period","Hyperarid","Arid","Semi-arid","Dry subhumid","Sum drylands","Humid","Cold","NA"), names(df.percent))
  
  df.percent[, missing] <- 0
  df.sd[,missing] <- 0
  
  df.percent <- df.percent %>% select(c("model","period","Hyperarid","Arid","Semi-arid","Dry subhumid","Sum drylands","Humid","Cold","NA"))
  df.percent[is.na(df.percent)] <- 0
  
  tab_list_percent[[i]] <- df.percent
  
  df.sd <- df.sd %>% select(c("model","period","Hyperarid","Arid","Semi-arid","Dry subhumid","Sum drylands","Humid","Cold","NA"))
  df.sd[is.na(df.sd)] <- 0
  
  tab_list_sd[[i]] <- df.sd
}

k_list <- list()

for(i in names(tab_list_percent)){
  
  df <- tab_list_percent[[i]] %>% select(c("model", "period"))
  
  df$Hyperarid <- paste(tab_list_percent[[i]]$Hyperarid, tab_list_sd[[i]]$Hyperarid, sep = "  ")
  df$Arid <- paste(tab_list_percent[[i]]$Arid, tab_list_sd[[i]]$Arid, sep = "  ")
  df$'Semi-Arid' <- paste(tab_list_percent[[i]][,5], tab_list_sd[[i]][,5], sep = "  ")
  df$'Dry subhumid' <- paste(tab_list_percent[[i]][,6], tab_list_sd[[i]][,6], sep = "  ")
  df$'Sum drylands' <- paste(tab_list_percent[[i]][,7], tab_list_sd[[i]][,7], sep = "  ")
  df$Humid <- paste(tab_list_percent[[i]]$Humid, tab_list_sd[[i]]$Humid, sep = "  ")
  df$Cold <- paste(tab_list_percent[[i]]$Cold, tab_list_sd[[i]]$Cold, sep = "  ")
  
  k_list[[i]] <- kable(df[order(df$model, decreasing = F),], caption = i) %>% kable_styling(bootstrap_options = "bordered") %>%
    column_spec(8, italic = T, background = colorvec[5]) %>% row_spec(2, bold = T) 
  
}

k_list

```

### Changes in proportion of aridity categories, figure

```{r cat-changes-by-continent-in-graph-catmean, fig.dim = c(8,10)}
list_plots <- list()

for(i in names(tab_list_percent)){
  df.percent <- tab_list_percent[[i]]
  df.sd <- tab_list_sd[[i]]
  
  df <- merge(df.percent, df.sd, by = c("model", "period"))
  
  g <- ggplot(data = df)+geom_point(aes(x=period,y= get("Sum drylands.x"), col = model), shape = "\u2605", size = 5, position = position_dodge(width = 0.5))+
    geom_errorbar(aes(x = period, ymin = get("Sum drylands.x")-get("Sum drylands.y"), ymax = get("Sum drylands.x")+get("Sum drylands.y"), col = model), position = position_dodge(width = 0.5))+
    scale_color_manual(values = colorvec[c(1,3,9,11)])+
    ylim(0,100)+
    labs(x="", y = "Percent of drylands", title = i)+
    theme_minimal()
  list_plots[[i]] <- g
}

ggarrange(plotlist = list_plots, ncol = 2, nrow = 4, common.legend = T, legend = "bottom")
```

```{r colplots-futurecat-continent, fig.dim = c(15,25)}
tab.flow <- cmip6s %>% subset(!Continent %in% c("SOUTHERN","PACIFIC","POLAR","ATLANTIC","INDIAN","ARCTIC")) %>%
  group_by(period, model, Continent, cat.AI) %>%
  summarise(count = n()) %>%
  ungroup() %>% group_by(period, model, Continent) %>% mutate(percent = round(count/sum(count)*100, 1), count = NULL)

df245 <- rbind(subset(tab.flow, model == "historical" & period %in% c("1850_1880", "1970_2000")), subset(tab.flow, model == "SSP245")) %>% mutate(lab.y = (rev(cumsum(rev(percent)))) - percent*0.5) %>% 
  subset(cat.AI %in% c("Hyperarid", "Arid", "Semi-arid", "Dry subhumid", "Humid","Cold"))

b245 <- ggplot(data = df245, aes(x = period, y = percent))+
  geom_col(aes(group = cat.AI, col = cat.AI, fill = cat.AI))+
  geom_label(aes(y = lab.y, label = percent))+
  scale_color_manual(values = col.cat, aesthetics = c("col", "fill"), na.translate = F)+
  facet_grid(rows = vars(Continent), switch = "y")+
  scale_y_continuous(position = "right")+
  labs(title = "SSP 2-4.5", y = "%", x = "")+
  theme_minimal()

df370 <- rbind(subset(tab.flow, model == "historical" & period %in% c("1850_1880", "1970_2000")), subset(tab.flow, model == "SSP370")) %>% mutate(lab.y = (rev(cumsum(rev(percent)))) - percent*0.5) %>% 
  subset(cat.AI %in% c("Hyperarid", "Arid", "Semi-arid", "Dry subhumid", "Humid","Cold"))

b370 <- ggplot(data = df370, aes(x = period, y = percent))+
  geom_col(aes(group = cat.AI, col = cat.AI, fill = cat.AI))+
  geom_label(aes(y = lab.y, label = percent))+
  scale_color_manual(values = col.cat, aesthetics = c("col", "fill"), na.translate = F)+
  facet_grid(rows = vars(Continent), switch = "y")+
  scale_y_continuous(position = "right")+  labs(title = "SSP 3-7.0", y = "%", x = "")+
  theme_minimal()


df585 <- rbind(subset(tab.flow, model == "historical" & period %in% c("1850_1880", "1970_2000")), subset(tab.flow, model == "SSP585")) %>% mutate(lab.y = (rev(cumsum(rev(percent)))) - percent*0.5) %>% 
  subset(cat.AI %in% c("Hyperarid", "Arid", "Semi-arid", "Dry subhumid", "Humid","Cold"))

b585 <- ggplot(data = df585, aes(x = period, y = percent))+
  geom_col(aes(group = cat.AI, col = cat.AI, fill = cat.AI))+
  geom_label(aes(y = lab.y, label = percent))+
  scale_color_manual(values = col.cat, aesthetics = c("col", "fill"), na.translate = F)+
  facet_grid(rows = vars(Continent), switch = "y")+
  scale_y_continuous(position = "right")+  labs(title = "SSP 5-8.5", y = "%", x = "")+
  theme_minimal()


ggarrange(plotlist = list(b245, b370, b585), common.legend = T, legend = "bottom", ncol = 3)
```



# Changes of variables over time and scenarios 


## Temperature 

```{r temperature}
quantile(cmip6s$t.mean, probs = seq(0,1,0.1)) 

t.breaks <- c(-50,-40,-30,-20,-10,0,10,20,30)
colscale <- c("#08519c", "#3182bd", "#6baed6", "#9ecae1", "#c6dbef", "#fddbc7", "#f4a582", "#d6604d", "#b2182b", "#67001f")
```

```{r map-historical-changes-quantiles-temperature, fig.dim = c(5,10)}
map_list <- list()

for(i in c("1850_1880", "1970_2000","1985_2015")){
  g <- ggplot() + geom_raster(data = subset(cmip6s, period == i & model == "historical"), aes(x=lon, y = lat,  fill = t.mean))+
    borders(colour = "grey60")+
    binned_scale(aesthetics = "fill", breaks = t.breaks, palette = function(x) colscale,
                 guide = guide_legend(label.theme = element_text(angle = 0)))+
    labs(title = i, fill = "Mean annual temperature, C")+
    theme_void()+ylim(-55,90)+
    theme(legend.position = "right")
  map_list[[i]] <- g
}

ggpubr::ggarrange(plotlist = map_list, ncol = 1, nrow = 3, common.legend = T, legend = "bottom")

```

```{r map-future-changes-quantiles-temperature, fig.dim = c(10,10)}

map_list <- list()

for(i in c("SSP245","SSP370","SSP585")){
  for (j in c("2030_2060","2070_2100")){
    index <- paste(i, j, sep = " , ")
    g <- ggplot() + geom_raster(data = subset(cmip6s, period == j & model == i), aes(x=lon, y = lat,  fill = t.mean))+
      borders(colour = "grey60")+
      binned_scale(aesthetics = "fill", breaks = t.breaks, palette = function(x) colscale,
                   guide = guide_legend(label.theme = element_text(angle = 0)))+
      labs(fill = "Mean annual temperature, C", title = index)+
      theme_void()+ylim(-55,90)+
      theme(legend.position = "bottom")
    map_list[[index]] <- g
  }}



ggpubr::ggarrange(plotlist = map_list,ncol = 2, nrow = 3, common.legend = T, legend = "bottom")

#labels = c("2030-2060", "2070-2100"
```

## Precipitation 

```{r precipitation}
quantile(cmip6s$pr.mean, probs = seq(0,1,0.1)) 

p.breaks <- c(10,50,100,200,300,400,500,600,700)
colscale <- c("#08519c", "#3182bd", "#6baed6", "#9ecae1", "#c6dbef", "#fddbc7", "#f4a582", "#d6604d", "#b2182b", "#67001f")
```

```{r map-historical-changes-quantiles-precipitation, fig.dim = c(5,10)}
map_list <- list()

for(i in c("1850_1880","1970_2000","1985_2015")){
  g <- ggplot() + geom_raster(data = subset(cmip6s, period == i & model == "historical"), aes(x=lon, y = lat,  fill = pr.mean))+
    borders(colour = "grey60")+
    binned_scale(aesthetics = "fill", breaks = p.breaks, palette = function(x) rev(colscale),
                 guide = guide_legend(label.theme = element_text(angle = 0)))+
    labs(title = i, fill = "Annual mean precipitation")+
    theme_void()+ylim(-55,90)+
    theme(legend.position = "right")
  map_list[[i]] <- g
}

ggpubr::ggarrange(plotlist = map_list, ncol = 1, nrow = 3, common.legend = T, legend = "bottom")

```

```{r map-future-changes-quantiles-precipitation, fig.dim = c(10,10)}

map_list <- list()

for(i in c("SSP245","SSP370","SSP585")){
  for (j in c("2030_2060","2070_2100")){
    index <- paste(i, j, sep = " , ")
    g <- ggplot() + geom_raster(data = subset(cmip6s, period == j & model == i), aes(x=lon, y = lat,  fill = pr.mean))+
      borders(colour = "grey60")+
      binned_scale(aesthetics = "fill", breaks = p.breaks, palette = function(x) rev(colscale),
                   guide = guide_legend(label.theme = element_text(angle = 0)))+
      labs(fill = "Annual mean\nprecipitation, mm", title = index)+
      theme_void()+ylim(-55,90)+
      theme(legend.position = "bottom")
    map_list[[index]] <- g
  }}

ggpubr::ggarrange(plotlist = map_list,ncol = 2, nrow = 3, common.legend = T, legend = "bottom")
```




