---
title: "Compare datasets"
author: "Camille"
date: "2024-02-15"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: hide
---
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = F, warning = F, error = F, fig.align = "center", collapse = T)
```

```{r libraries}
library(raster)

library(plyr)
library(dplyr)

library(ggplot2)
library(ggpmisc)

library(viridisLite)
library(colorspace)
library(RColorBrewer)

library(sf)
library(stringr)

#display.brewer.pal(n = 11, name ="RdYlBu")
colorvec <- brewer.pal(n = 11, name ="RdYlBu")

col.martonne <- c("Desert" = colorvec[2], "Arid"= colorvec[4],"Semi-arid"= colorvec[5], "Temperate" = colorvec[8], "Humid"= colorvec[10],"Forest"= colorvec[11])
col.cat <- c("Hyperarid" = colorvec[2], "Arid"= colorvec[4],"Semi-arid"= colorvec[5], "Dry subhumid" = colorvec[8], "Humid"= colorvec[10], "NA" = "#D5D8DC")
```

Liens utiles:

* https://www.fao.org/3/x0490E/x0490e0k.htm#annex%203.%20background%20on%20physical%20parameters%20used%20in%20evapotranspiration%20computatio

```{r make-data, eval = F}
# era5vA <- read.table("era5_AI.txt") %>%
#   mutate(tas = tas-273.15, pr = mpr * 60 * 60 * 24 * 365, rsds = ssrd / (60*60*24) * 1e-6 * 60*60*24) %>% # conversion from J/m2 to J/sm2 to MJ/m2/day  
#   select(c("lon","lat","Continent","Type","Name","Acronym","lm","model","source","pr","tas","sfcWind","Rn","rsds","ea","ET0","AI","cat.AI"))
# 
# 
# write.table(era5vA, "era5vA.txt")

era5vA <- read.table("ERA5/et0_30y.txt") %>% mutate(tas = t2m - 273.15, pr = mpr * 60 * 60 * 24 * 365, rsds = ssrd / 86400 * 86400 * 1e-6, sfcWind = wind, model = "historical", source = "ERA5") %>%  select(c("lon","lat","Continent","Type","Name","Acronym","lm","model","source","pr","tas","sfcWind","Rn","rsds","ea","ET0","AI","cat.AI"))

write.table(era5vA, "era5vA.txt")

wdcvA <- read.table("wdc_AI.txt") %>% mutate(rsds = srad*1e-3, ea = vapr, ET0 = ET0/365) %>% # srad in MJ/m2/day, ET0 originnally in mm/year
 select(c("lon","lat","Continent","Type","Name","Acronym","lm","model","source","pr","tas","sfcWind","Rn", "rsds","ea","ET0","AI","cat.AI"))


write.table(wdcvA, "wdcvA.txt")


cmip6 <- read.table("cmip6_AI.txt")

cmip6vA <- cmip6 %>% subset(period == "1970_2000") %>%
   group_by(lon,lat,Continent, Type, Name, Acronym, lm, model) %>% 
  dplyr::summarise(tas = mean(tas, na.rm = T) - 273.15, pr = mean(pr, na.rm = T) * 60 * 60 * 24 * 365,
                   sfcWind = mean(sfcWind, na.rm = T), Rn = mean(Rn, na.rm = T),
                   rsds = mean(rsds, na.rm = T)  * 1e-6 * 60*60*24, # from W/m2 to MJ/m2/day
                   ea = mean(ea, na.rm = T), ET0 = mean(ET0, na.rm = T), AI = mean(AI, na.rm = T))%>%
  ungroup() %>%  
  mutate(source = "CMIP6")


breaks.unesco <- c(-1,0.03,0.2,0.5,0.65,Inf)
cat.unesco <- c("(-1,0.03]" = "Hyperarid", "(0.03,0.2]" = "Arid", "(0.2,0.5]" = "Semi-arid", "(0.5,0.65]" = "Dry subhumid","(0.65,Inf]" = "Humid")

cmip6vA$cat.AI <-  cmip6vA$AI %>% cut(breaks = breaks.unesco) %>% 
  revalue(cat.unesco)

write.table(cmip6vA, "cmip6vA.txt")

```


```{r load-data}
era5vA <- read.table("era5vA.txt")
wdcvA <- read.table("wdcvA.txt")
cmip6vA <- read.table("cmip6vA.txt")

```

# Aridity Index 


```{r plot-compare, fig.dim = c(12,9)}

calib <- rbind(cmip6vA, era5vA, wdcvA)

breaks.unesco <- c(-1,0.03,0.2,0.5,0.65,Inf)
cat.unesco <- c("(-1,0.03]" = "Hyperarid", "(0.03,0.2]" = "Arid", "(0.2,0.5]" = "Semi-arid", "(0.5,0.65]" = "Dry subhumid","(0.65,Inf]" = "Humid")

calib$cat.AI <-  calib$AI %>% cut(breaks = breaks.unesco) %>% 
  revalue(cat.unesco)

cowplot::plot_grid(ggplot(data = subset(calib, source == "Worldclim")) + geom_raster(aes(x=lon, y = lat,  fill = cat.AI))+
    scale_fill_manual(values = col.cat)+
    labs(title = "Wordclim, 1970-2000", fill = "")+
    theme_void()+
    theme(legend.position = "bottom"),

ggplot(data = subset(calib, source == "ERA5")) + geom_raster(aes(x=lon, y = lat,  fill = cat.AI))+
    scale_fill_manual(values = col.cat)+
    labs(title = "ERA5, 1970-2000", fill = "")+
    theme_void()+
    theme(legend.position = "bottom"),

ggplot(data = subset(calib, source == "CMIP6")) + geom_raster(aes(x=lon, y = lat,  fill = cat.AI))+
    scale_fill_manual(values = col.cat)+
    labs(title = "CMIP6, 1970-2000", fill = "")+
    theme_void()+
    theme(legend.position = "bottom"))
```


```{r compare-by-region, eval = T}
# table_calib_c <- calib %>% select(c("Continent","source","cat.AI")) %>%
#   reshape2::melt(id.vars = c("Continent","source")) %>%
#   group_by(Continent,source,value) %>%
#   summarise(count = n()) %>% mutate(perc = round(count/sum(count)*100,1)) %>%
#   ungroup()
# 
# table_calib_c$value[is.na(table_calib_c$value) == T] <- "NA"
# 
# ggplot(table_calib_c)+
#   geom_col(aes(x = Continent, y = perc, fill = value, position = "fill"))+
#   scale_fill_manual(values = col.cat, breaks = c("Hyperarid","Arid","Semi-arid","Dry subhumid","Humid","NA"))+
#   theme_minimal()+
#   theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))+
#   facet_grid(rows = vars(source))+
#   labs(x = "Continent", y = "%", fill = "Aridity\ncategory", y = "Proportion of aridity category")
# 


table_calib <- calib %>% select(c("Continent","Name","Acronym","source","cat.AI")) %>%
  reshape2::melt(id.vars = c("Continent","Name","Acronym","source")) %>%
  group_by(Continent,Name,Acronym,source,value) %>%
  summarise(count = n()) %>% mutate(perc = round(count/sum(count)*100,1)) %>%
  ungroup() %>% filter(!Continent %in% c("SOUTHERN","PACIFIC","POLAR","ATLANTIC","INDIAN","ARCTIC"))

table_calib$value[is.na(table_calib$value) == T] <- "NA"
table_calib$Continent[which(table_calib$Continent == "EUROPE-AFRICA")] <- "EUROPE"
table_calib$Continent[which(table_calib$Continent == "CENTRAL-AMERICA")] <- "CENTRAL\nAMERICA"
table_calib$Continent[which(table_calib$Continent == "NORTH-AMERICA")] <- "NORTH\nAMERICA"
table_calib$Continent[which(table_calib$Continent == "SOUTH-AMERICA")] <- "SOUTH\nAMERICA"

table_calib$Continent <- factor(table_calib$Continent, levels = c("AFRICA","EUROPE","ASIA","OCEANIA","NORTH\nAMERICA","CENTRAL\nAMERICA","SOUTH\nAMERICA"))

ggplot(table_calib)+
  geom_col(aes(x = Acronym, y = perc, fill = value, position = "fill"))+
  scale_fill_manual(values = col.cat, breaks = c("Hyperarid","Arid","Semi-arid","Dry subhumid","Humid","NA"))+
  scale_y_continuous(position = "right")+
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5), 
        legend.position = "bottom", 
        strip.text.x = element_text(size = 6))+
  facet_grid(source ~ Continent, switch = "y", scales = "free_x", space = "free")+
  labs(x = "", y = "%", fill = "Aridity\ncategory")

```

# Comparison of datasets 

```{r calib-wide}
vars <- c("lon","lat","Continent","Name","Acronym","tas","pr","sfcWind","Rn","rsds","ea","ET0","AI","cat.AI")
merge.vars <- c("lon","lat","Continent","Name","Acronym")
calib.wide <- select(era5vA, vars) %>%
               merge(select(wdcvA, vars), 
                     by = merge.vars) %>%
          merge(select(cmip6vA, vars), 
                by = merge.vars) %>%
          setNames(c(merge.vars,
                     "tas.era5","pr.era5", "sfcWind.era5","Rn.era5","rsds.era5","ea.era5","ET0.era5","AI.era5","cat.AI.era5",
                     "tas.wdc","pr.wdc", "sfcWind.wdc","Rn.wdc","rsds.wdc","ea.wdc","ET0.wdc","AI.wdc","cat.AI.wdc",
                     "tas.cmip6","pr.cmip6", "sfcWind.cmip6","Rn.cmip6","rsds.cmip6","ea.cmip6","ET0.cmip6","AI.cmip6","cat.AI.cmip6"))
```

## CMIP6 & ERA5 compared to WDC

### Maps

In red: when CMIP6 or ERA5 are warmer than Wordlclim

```{r map-diff-tas, fig.dim = c(10,3)}
cowplot::plot_grid(
  
ggplot(calib.wide)+
  geom_raster(aes(x=lon, y = lat, fill = tas.era5-tas.wdc))+
   scale_fill_continuous_divergingx(palette = "RdYlBu", mid = 0, rev = T, name = "Temperature in °C: ERA5 - WDC")+
  theme_void()+
  theme(legend.position = "bottom"),

ggplot(calib.wide)+
  geom_raster(aes(x=lon, y = lat, fill = tas.cmip6-tas.wdc))+
   scale_fill_continuous_divergingx(palette = "RdYlBu", mid = 0, rev = T, name = "Temperature in °C: CMIP6 - WDC")+
  theme_void()+
  theme(legend.position = "bottom")
)
```

In blue: when CMIP6 or ERA5 are wetter than Worldclim

```{r map-diff-pr, fig.dim = c(10,3)}
cowplot::plot_grid(
  
  ggplot(calib.wide)+
  geom_raster(aes(x=lon, y = lat, fill = pr.era5-pr.wdc))+
   scale_fill_continuous_divergingx(palette = "RdYlBu", mid = 0, name = "Precipitation in mm/y: ERA5 - WDC", guide = "legend")+
  theme_void()+
  theme(legend.position = "bottom"),

ggplot(calib.wide)+
  geom_raster(aes(x=lon, y = lat, fill = pr.cmip6-pr.wdc))+
   scale_fill_continuous_divergingx(palette = "RdYlBu", mid = 0, name = "Precipitation in mm/y: CMIP6 - WDC", guide = "legend")+
  theme_void()+
  theme(legend.position = "bottom")
)
```

In blue: when wind speed is higher for CMIP6 or ERA5 compared to Worldclim

```{r map-diff-sfcWind, fig.dim = c(10,3)}
cowplot::plot_grid(
  
ggplot(calib.wide)+
  geom_raster(aes(x=lon, y = lat, fill = sfcWind.era5-sfcWind.wdc))+
   scale_fill_continuous_divergingx(palette = "RdYlBu", mid = 0, name = "Wind speed in m/s: ERA5 - WDC", guide = "legend")+
  theme_void()+
  theme(legend.position = "bottom"),

ggplot(calib.wide)+
  geom_raster(aes(x=lon, y = lat, fill = sfcWind.cmip6-sfcWind.wdc))+
   scale_fill_continuous_divergingx(palette = "RdYlBu", mid = 0, name = "Wind speed in m/s: CMIP6 - WDC", guide = "legend")+
  theme_void()+
  theme(legend.position = "bottom")
)
```

### Scatterplot 

Representing for scatterplot only tas, pr and sfcWind (Rn need checking in Wdc, ea needs checking in CMIP6)

```{r scatterplot-diff, fig.dim = c(10,20)}
cowplot::plot_grid(
  
ggplot(calib.wide,aes(x = tas.wdc))+geom_point(aes( y = tas.era5, col = "ERA5"), alpha =0.4, shape = 1)+
  geom_point(aes(y=tas.cmip6, col = "CMIP6"), alpha = 0.4, shape = 1)+
  stat_poly_eq(aes(y = tas.era5, col = "ERA5")) +
  stat_poly_eq(aes(y = tas.cmip6, col = "CMIP6"), label.x = "right",label.y = "bottom") +
  geom_abline(slope = 1, intercept = 0)+
  labs(x="Temperature in °C, WDC", y = "Temperature in °C", col = "Dataset")+
  scale_color_manual(values = c(colorvec[3], colorvec[9]))+
  theme_minimal()+
  theme(legend.position = "none"),

ggplot(calib.wide,aes(x = pr.wdc))+geom_point(aes( y = pr.era5, col = "ERA5"), alpha =0.4, shape = 1)+
  geom_point(aes(y=pr.cmip6, col = "CMIP6"), alpha = 0.4, shape = 1)+
  stat_poly_eq(aes(y = pr.era5, col = "ERA5")) +
  stat_poly_eq(aes(y = pr.cmip6, col = "CMIP6"), label.x = "right",label.y = "bottom") +
  geom_abline(slope = 1, intercept = 0)+
  labs(x="Annual precipitation, mm, WDC", y = "Annual precipitation, mm", col = "Dataset")+
  scale_color_manual(values = c(colorvec[3], colorvec[9]))+
  theme_minimal()+
  theme(legend.position = "none"),

ggplot(calib.wide,aes(x = sfcWind.wdc))+geom_point(aes( y = sfcWind.era5, col = "ERA5"), alpha =0.4, shape = 1)+
  geom_point(aes(y=sfcWind.cmip6, col = "CMIP6"), alpha = 0.4, shape = 1)+
  stat_poly_eq(aes(y = sfcWind.era5, col = "ERA5")) +
  stat_poly_eq(aes(y = sfcWind.cmip6, col = "CMIP6"), label.x = "right",label.y = "bottom") +
  geom_abline(slope = 1, intercept = 0)+
  labs(x="Wind speed m/s, WDC", y = "Wind speed m/s", col = "Dataset")+
  scale_color_manual(values = c(colorvec[3], colorvec[9]))+
  theme_minimal()+
  theme(legend.position = "none"),

ggplot(calib.wide,aes(x = Rn.wdc))+geom_point(aes( y = Rn.era5, col = "ERA5"), alpha =0.4, shape = 1)+
   geom_point(aes(y=Rn.cmip6, col = "CMIP6"), alpha = 0.4, shape = 1)+
  stat_poly_eq(aes(y = Rn.era5, col = "ERA5")) +
  stat_poly_eq(aes(y = Rn.cmip6, col = "CMIP6"), label.x = "right",label.y = "bottom") +
   geom_abline(slope = 1, intercept = 0)+
   labs(x="Rn, MJ/M2/m2, WDC, °C", y = "Rn, MJ/M2/m2", col = "Dataset")+
   scale_color_manual(values = c(colorvec[3], colorvec[9]))+
   theme_minimal()+
  theme(legend.position = "none"),

ggplot(calib.wide,aes(x = rsds.wdc))+geom_point(aes( y = rsds.era5, col = "ERA5"), alpha =0.4, shape = 1)+
   geom_point(aes(y=rsds.cmip6, col = "CMIP6"), alpha = 0.4, shape = 1)+
  stat_poly_eq(aes(y = rsds.era5, col = "ERA5")) +
  stat_poly_eq(aes(y = rsds.cmip6, col = "CMIP6"), label.x = "right",label.y = "bottom") +
   geom_abline(slope = 1, intercept = 0)+
   labs(x="Rsds, W/m2, WDC, °C", y = "Rsds, W/m2", col = "Dataset")+
   scale_color_manual(values = c(colorvec[3], colorvec[9]))+
   theme_minimal()+
  theme(legend.position = "none"),

ggplot(calib.wide,aes(x = ea.wdc))+geom_point(aes(y = ea.era5, col = "ERA5"), alpha =0.4, shape = 1)+
   geom_point(aes(y=ea.cmip6, col = "CMIP6"), alpha = 0.4, shape = 1)+
  stat_poly_eq(aes(y = ea.era5, col = "ERA5")) +
  stat_poly_eq(aes(y = ea.cmip6, col = "CMIP6"), label.x = "right",label.y = "bottom") +
   geom_abline(slope = 1, intercept = 0)+
   labs(x="ea, kPa, WDC, °C", y = "ea, kPa", col = "Dataset")+
   scale_color_manual(values = c(colorvec[3], colorvec[9]))+
   theme_minimal()+
  theme(legend.position = "none"),

ggplot(calib.wide,aes(x = ET0.wdc))+geom_point(aes(y = ET0.era5, col = "ERA5"), alpha =0.4, shape = 1)+
   geom_point(aes(y=ET0.cmip6, col = "CMIP6"), alpha = 0.4, shape = 1)+
  stat_poly_eq(aes(y = ET0.era5, col = "ERA5")) +
  stat_poly_eq(aes(y = ET0.cmip6, col = "CMIP6"), label.x = "right",label.y = "bottom") +
   geom_abline(slope = 1, intercept = 0)+
   labs(x="ET0, mm/day, WDC", y = "ET0, mm/day", col = "Dataset")+
   scale_color_manual(values = c(colorvec[3], colorvec[9]))+
   theme_minimal()+
  theme(legend.position = "none"),

ggplot(calib.wide,aes(x = sfcWind.wdc))+geom_point(aes( y = sfcWind.era5, col = "ERA5"), alpha =1, shape = 1)+
  geom_point(aes(y=sfcWind.cmip6, col = "CMIP6"), alpha = 1, shape = 1)+
  geom_point(aes( y = sfcWind.era5), alpha =1, shape = 1, col = "white")+
  geom_point(aes(y=sfcWind.cmip6), alpha = 1, shape = 1, col = "white")+
  labs(x="Wind speed m/s, WDC", y = "Wind speed m/s", col = "Dataset")+
  scale_color_manual(values = c(colorvec[3], colorvec[9]))+
  theme_void(base_size=15)+
  theme(legend.position = "top"),


ncol = 2
)

```

## CMIP6 & WDC compared to ERA5

### Maps

In red: when CMIP6 or WDC are warmer than ERA5

```{r map-diff-tas-2, fig.dim = c(10,3)}
cowplot::plot_grid(
  
ggplot(calib.wide)+
  geom_raster(aes(x=lon, y = lat, fill = tas.wdc-tas.era5))+
   scale_fill_continuous_divergingx(palette = "RdYlBu", mid = 0, rev = T, name = "Temperature in °C: WDC - ERA5")+
  theme_void()+
  theme(legend.position = "bottom"),

ggplot(calib.wide)+
  geom_raster(aes(x=lon, y = lat, fill = tas.cmip6-tas.era5))+
   scale_fill_continuous_divergingx(palette = "RdYlBu", mid = 0, rev = T, name = "Temperature in °C: CMIP6 - ERA5")+
  theme_void()+
  theme(legend.position = "bottom")
)
```

In blue: when CMIP6 or WDC are wetter than ERA5

```{r map-diff-pr-2, fig.dim = c(10,3)}
cowplot::plot_grid(
  
  ggplot(calib.wide)+
  geom_raster(aes(x=lon, y = lat, fill = pr.wdc-pr.era5))+
   scale_fill_continuous_divergingx(palette = "RdYlBu", mid = 0, name = "Precipitation in mm/y: WDC - ERA5", guide = "legend")+
  theme_void()+
  theme(legend.position = "bottom"),

ggplot(calib.wide)+
  geom_raster(aes(x=lon, y = lat, fill = pr.cmip6-pr.era5))+
   scale_fill_continuous_divergingx(palette = "RdYlBu", mid = 0, name = "Precipitation in mm/y: CMIP6 - ERA5", guide = "legend")+
  theme_void()+
  theme(legend.position = "bottom")
)
```

In blue: when wind speed is higher for CMIP6 or WDC compared to ERA5

```{r map-diff-sfcWind-2, fig.dim = c(10,3)}
cowplot::plot_grid(
  
ggplot(calib.wide)+
  geom_raster(aes(x=lon, y = lat, fill = sfcWind.wdc-sfcWind.era5))+
   scale_fill_continuous_divergingx(palette = "RdYlBu", mid = 0, name = "Wind speed in m/s: WDC - ERA5", guide = "legend")+
  theme_void()+
  theme(legend.position = "bottom"),

ggplot(calib.wide)+
  geom_raster(aes(x=lon, y = lat, fill = sfcWind.cmip6-sfcWind.era5))+
   scale_fill_continuous_divergingx(palette = "RdYlBu", mid = 0, name = "Wind speed in m/s: CMIP6 - ERA5", guide = "legend")+
  theme_void()+
  theme(legend.position = "bottom")
)
```

### Scatterplot 

Representing for scatterplot only tas, pr and sfcWind (Rn need checking in Wdc, ea needs checking in CMIP6)

```{r scatterplot-diff-2, fig.dim = c(10,20)}
cowplot::plot_grid(
  
ggplot(calib.wide,aes(x = tas.era5))+geom_point(aes(y = tas.wdc, col = "WDC"), alpha =0.4, shape = 1)+
  geom_point(aes(y=tas.cmip6, col = "CMIP6"), alpha = 0.4, shape = 1)+
  stat_poly_eq(aes(y = tas.wdc, col = "WDC")) +
  stat_poly_eq(aes(y = tas.cmip6, col = "CMIP6"), label.x = "right",label.y = "bottom") +
  geom_abline(slope = 1, intercept = 0)+
  labs(x="Temperature in °C, ERA5", y = "Temperature in °C", col = "Dataset")+
  scale_color_manual(values = c(colorvec[3], colorvec[9]))+
  theme_minimal()+
  theme(legend.position = "none"),

ggplot(calib.wide,aes(x = pr.era5))+geom_point(aes( y = pr.wdc, col = "WDC"), alpha =0.4, shape = 1)+
  geom_point(aes(y=pr.cmip6, col = "CMIP6"), alpha = 0.4, shape = 1)+
    stat_poly_eq(aes(y = pr.wdc, col = "WDC")) +
  stat_poly_eq(aes(y = pr.cmip6, col = "CMIP6"), label.x = "right",label.y = "bottom") +
  geom_abline(slope = 1, intercept = 0)+
  labs(x="Annual precipitation, mm, ERA5", y = "Annual precipitation, mm", col = "Dataset")+
  scale_color_manual(values = c(colorvec[3], colorvec[9]))+
  theme_minimal()+
  theme(legend.position = "none"),

ggplot(calib.wide,aes(x = sfcWind.era5))+geom_point(aes( y = sfcWind.wdc, col = "WDC"), alpha =0.4, shape = 1)+
  geom_point(aes(y=sfcWind.cmip6, col = "CMIP6"), alpha = 0.4, shape = 1)+
    stat_poly_eq(aes(y = sfcWind.wdc, col = "WDC")) +
  stat_poly_eq(aes(y = sfcWind.cmip6, col = "CMIP6"), label.x = "right",label.y = "bottom") +
  geom_abline(slope = 1, intercept = 0)+
  labs(x="Wind speed m/s, ERA5", y = "Wind speed m/s", col = "Dataset")+
  scale_color_manual(values = c(colorvec[3], colorvec[9]))+
  theme_minimal()+
  theme(legend.position = "none"),

ggplot(calib.wide,aes(x = Rn.era5))+geom_point(aes( y = Rn.wdc, col = "WDC"), alpha =0.4, shape = 1)+
   geom_point(aes(y=Rn.cmip6, col = "CMIP6"), alpha = 0.4, shape = 1)+
    stat_poly_eq(aes(y = Rn.wdc, col = "WDC")) +
  stat_poly_eq(aes(y = Rn.cmip6, col = "CMIP6"), label.x = "right",label.y = "bottom") +
   geom_abline(slope = 1, intercept = 0)+
   labs(x="Rn, MJ/m2/day, ERA5, °C", y = "Rn, W/MJ/m2/day", col = "Dataset")+
   scale_color_manual(values = c(colorvec[3], colorvec[9]))+
   theme_minimal()+
  theme(legend.position = "none"),

ggplot(calib.wide,aes(x = rsds.era5))+geom_point(aes( y = rsds.wdc, col = "WDC"), alpha =0.4, shape = 1)+
   geom_point(aes(y=rsds.cmip6, col = "CMIP6"), alpha = 0.4, shape = 1)+
    stat_poly_eq(aes(y = rsds.wdc, col = "WDC")) +
  stat_poly_eq(aes(y = rsds.cmip6, col = "CMIP6"), label.x = "right",label.y = "bottom") +
   geom_abline(slope = 1, intercept = 0)+
   labs(x="rsds, W/m2, ERA5, °C", y = "rsds, W/m2", col = "Dataset")+
   scale_color_manual(values = c(colorvec[3], colorvec[9]))+
   theme_minimal()+
  theme(legend.position = "none"),

ggplot(calib.wide,aes(x = ea.era5))+geom_point(aes( y = ea.wdc, col = "WDC"), alpha =0.4, shape = 1)+
   geom_point(aes(y=ea.cmip6, col = "CMIP6"), alpha = 0.4, shape = 1)+
    stat_poly_eq(aes(y = ea.wdc, col = "WDC")) +
  stat_poly_eq(aes(y = ea.cmip6, col = "CMIP6"), label.x = "right",label.y = "bottom") +
   geom_abline(slope = 1, intercept = 0)+
   labs(x="ea, kPa, ERA5", y = "ea, kPa", col = "Dataset")+
   scale_color_manual(values = c(colorvec[3], colorvec[9]))+
   theme_minimal()+
  theme(legend.position = "none"),

ggplot(calib.wide,aes(x = ET0.era5))+geom_point(aes(y = ET0.wdc, col = "WDC"), alpha =0.4, shape = 1)+
   geom_point(aes(y=ET0.cmip6, col = "CMIP6"), alpha = 0.4, shape = 1)+
    stat_poly_eq(aes(y = ET0.wdc, col = "WDC")) +
  stat_poly_eq(aes(y = ET0.cmip6, col = "CMIP6"), label.x = "right",label.y = "bottom") +
   geom_abline(slope = 1, intercept = 0)+
   labs(x="ET0, mm/day, ERA5", y = "ET0, mm/day", col = "Dataset")+
   scale_color_manual(values = c(colorvec[3], colorvec[9]))+
   theme_minimal()+
  theme(legend.position = "none"),

ggplot(calib.wide,aes(x = sfcWind.era5))+geom_point(aes( y = sfcWind.wdc, col = "WDC"), alpha =0.4, shape = 1)+
  geom_point(aes(y=sfcWind.cmip6, col = "CMIP6"), alpha = 0.4, shape = 1)+
  geom_point(aes( y = sfcWind.wdc), alpha =1, shape = 1, col = "white")+
  geom_point(aes(y=sfcWind.cmip6), alpha = 1, shape = 1, col = "white")+
  labs(x="Wind speed m/s, ERA5", y = "Wind speed m/s", col = "Dataset")+
  scale_color_manual(values = c(colorvec[3], colorvec[9]))+
  theme_void(base_size=15)+
  theme(legend.position = "top"),

ncol = 2
)


```



```{r exit}
knitr::knit_exit()
```

```{r compare-with NDVI}
# Download Gimms NDVI
library(gimms)

ndvi.ref <- downloadGimms(x= 2000,y=2000,dsn = "NDVI")
ndvi.max <- monthlyComposite(ndvi.ref,monthlyIndices(ndvi.ref))
saveRDS(ndvi.max,"ndvi.max.rds")

```
